### 2021-11-16 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
initial analysis

---
title: "LOINC2HPO_EPIC AND SCC"
output:
  html_document:
    df_print: paged
---

## GOAL
This is to LOINC2HPO analysis of both EPIC and SCC labs. 
```{r message=FALSE}
library(tidyverse)
library(glue)
library(readxl)
require(DBI)
require(RPostgres)
require(egg)
require(gridExtra)
source('.db_credential.R')
dsca <- DBI::dbConnect(RPostgres::Postgres(), host='localhost', user=user, dbname = dbname, port = 2345, password=password)
```

## ANALYSIS

### Prepare the lab tables

Run the following script to create the denormalized table for lab_scc and lab_epic. After this step, we should have three tables for SCC lab, EPIC lab, and combined. And each table should have LOINC and mapped HPO. 
```{r}
# sql_folder_path <- '/Users/xingminzhang/git/s4-pd-dmsdw/pd-dmsdw-loinc2hpo-app/src/main/resources/sql'
# sql_denormalized_scc_lab_2020july <- read_file(glue('{sql_folder_path}/loinc2hpo_SCC_MSDW.sql'))
# sql_denormalized_epic_lab_2020july <- read_file(glue('{sql_folder_path}/loinc2hpo_Epic_MSDW.sql'))
# sql_combine_Epic_SCC <- read_file(glue('{sql_folder_path}/loinc2hpo_combine_Epic_SCC.sql')) 
# 
# # make sure the lab table after LOINC2HPO transformation exists
# # if not, we can create it here; but the intended behavior is that we use the Java commandline app for this

```

```{sql connection=dsca, echo=FALSE}
drop table if exists denormalized_scc_lab_2020july;
```
```{sql connection=dsca}
CREATE TEMP TABLE denormalized_scc_lab_2020july as
WITH 
	lab_scc_2020q2 as (
		SELECT l.medical_record_number , l.encounter_key , l.calendar_date , l.time_of_day_key , l.clinical_result_numeric ,  trim(from l.unit_of_measure_numeric) as unit_of_measure_numeric , l.reference_range , l.procedure_key,
		case when l.abnormal_flag is null then 'N' else l.abnormal_flag end as abnormal_flag 
		FROM hai_az_prod.lab_scc_2020july l
		--where medical_record_number in (2966097, 3164683, 3869795)
	), 
	lab_scc_abnormal_flag_mapping AS (
		SELECT case when abf.abnormal_flag is NULL then 'N' else abf.abnormal_flag end as abnormal_flag, abf.mapto 
		FROM hai_az_prod.lab_scc_abnormal_flag_mapping abf
	)
SELECT lab.medical_record_number, lab.encounter_key , lab.calendar_date  as lab_date, dtod.clock_time_24_hour || ':00' as lab_time, lab.clinical_result_numeric , fdp.context_name, fdp.context_procedure_code AS local_test_code , loinc.loinc , lab.clinical_result_numeric AS lab_result_numeric_value, lab.unit_of_measure_numeric As unit_of_measure ,lab.reference_range AS normal_range , lab.abnormal_flag , abf.mapto AS interpretation, l2h.hpotermid , l2h.isnegated 
FROM lab_scc_2020q2 lab 
left JOIN hai_az_prod.fd_procedure fdp using (procedure_key)
left JOIN hai_az_prod.loinc_mapping loinc on fdp.context_procedure_code = loinc.code and (lab.unit_of_measure_numeric = loinc.unit or (lab.unit_of_measure_numeric is NULL and loinc.unit is NULL))
left JOIN lab_scc_abnormal_flag_mapping abf using (abnormal_flag)
left JOIN hai_az_prod.loinc2hpo l2h on loinc.loinc = l2h.loincid and abf.mapto = l2h.code
join prod_msdw.d_time_of_day dtod using (time_of_day_key);
```

```{sql connection=dsca}
DROP TABLE IF EXISTS denormalized_epic_lab_2020july;
```

```{sql connection=dsca}
create temp table denormalized_epic_lab_2020july as
with interpreted as (
	SELECT lab.mrn, lab.order_date , lab.lab_time , lab.test_code , lab.test_result_value , lab.unit_of_measurement , lab.reference_range ,
	case 
			when test_result_value_final :: REAL < range_low_final ::REAL then 'L' 
			when test_result_value_final :: REAL > range_hi_final :: REAL then 'H'
			else 'N' 
	end as interpretation
	FROM hai_az_test.lab_epic_2020july lab
	--where mrn in (3710567, 8014250, 3144637)
	)
SELECT lab.mrn AS medical_record_number, NULL AS encounter_key, lab.order_date as lab_date, lab.lab_time, 'EPIC' AS context_name, lab.test_code AS local_test_code , local2loinc.loinc , lab.test_result_value AS lab_result_numeric_value, lab.unit_of_measurement AS unit_of_measure , lab.reference_range AS normal_range , NULL AS abnormal_flag,  interpretation,
loinc2hpo.hpotermid , loinc2hpo.isnegated 
FROM interpreted lab
LEFT JOIN hai_az_prod.loinc_mapping local2loinc on lab.test_code :: VARCHAR = local2loinc.code and (lab.unit_of_measurement = local2loinc.unit OR (lab.unit_of_measurement is NULL and local2loinc.unit is NULL))
LEFT JOIN hai_az_prod.loinc2hpo loinc2hpo on local2loinc.loinc = loinc2hpo.loincid and interpretation = loinc2hpo.code 
where local2loinc."source" = 'EPIC'; 
```

Join the above tables together
```{sql connection=dsca}
CREATE VIEW denormalized_scc_and_epic_lab_2020july AS
WITH epic_denormalized AS (
  SELECT medical_record_number, lab_date, lab_time, context_name, local_test_code :: VARCHAR, loinc,  lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated
  FROM denormalized_epic_lab_2020july
), 
scc_denormalized AS (
  SELECT medical_record_number, lab_date, lab_time, context_name, local_test_code, loinc, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated
  FROM denormalized_scc_lab_2020july
)
SELECT * FROM scc_denormalized 
UNION all 
SELECT * FROM epic_denormalized; 
```

Right now, we have three sets of denormalized labs, SCC, EPIC and their union.
```{sql connection=dsca}
SELECT * FROM denormalized_scc_lab_2020july limit 4;
```

```{sql connection=dsca}
SELECT * FROM denormalized_epic_lab_2020july limit 4;
```

```{r}
lab_tables <- c('denormalized_scc_lab_2020july', 'denormalized_epic_lab_2020july', 'denormalized_scc_and_epic_lab_2020july')
```

### Summary statistics before running any HPO inference

```{r}
abnormal_phenotypes_before_inference <- function(lab_table_name){
  results <- list()
  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_before_infer_per_patient;')
  sql <- glue::glue('create temp table n_abnormal_before_infer_per_patient AS
        SELECT medical_record_number , count(*) as n_abnormal_before_infer, count(distinct hpotermid ) as n_uniq_abnormal_before_infer
        FROM {lab_table_name}
        where hpotermid is not null and not isnegated
        group by medical_record_number ;')
  dbExecute(dsca, sql)    
  # median of HPO-coded phenotypes per patient from abnormal lab findings
  median_abnormal_hpo <- dbGetQuery(dsca, 'SELECT median(n_abnormal_before_infer) FROM n_abnormal_before_infer_per_patient;')[1,1]
  results$n_abnormal_before_infer <- median_abnormal_hpo

  # median of unique HPO-coded phenotypes per patient from abnormal lab findings
  median_uniq_abnormal_before_infer <- dbGetQuery(dsca, 'SELECT median(n_uniq_abnormal_before_infer) FROM n_abnormal_before_infer_per_patient;')[1,1]
  results$n_uniq_abnormal_before_infer <- median_uniq_abnormal_before_infer

  n_abnormal_before_infer_per_patient <- dbGetQuery(dsca, 
        'SELECT n_abnormal_before_infer, count(*) as n_patient 
      FROM n_abnormal_before_infer_per_patient
      GROUP BY n_abnormal_before_infer
      order by n_abnormal_before_infer;')
  results$n_abnormal_before_infer_per_patient <- n_abnormal_before_infer_per_patient

  n_uniq_abnormal_before_infer_per_patient <- dbGetQuery(dsca, 
          'SELECT n_uniq_abnormal_before_infer, count(*) as n_patient 
            FROM n_abnormal_before_infer_per_patient
            GROUP BY n_uniq_abnormal_before_infer
            order by n_uniq_abnormal_before_infer;')
  results$n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_before_infer_per_patient

  return (results)
}

# the input is the third item in the above function
plot_patient_distribution_by_n_abnormal_phenotypes_before_infer <- function(df){
  n_abnormal_before_infer_per_patient <- df
  n_abnormal_before_infer_per_patient <- n_abnormal_before_infer_per_patient %>% mutate(n_abnormal_before_infer = as.integer(n_abnormal_before_infer), n_patient = as.integer(n_patient))

## roll up the ones > 100
  df_temp <- n_abnormal_before_infer_per_patient %>% mutate(n_abnormal_before_infer = ifelse(n_abnormal_before_infer > 100, 100, n_abnormal_before_infer)) %>% group_by(n_abnormal_before_infer) %>% summarise(n_patient = sum(n_patient))

  p <- ggplot(df_temp) + 
    geom_bar(aes(x=n_abnormal_before_infer, y = n_patient), stat='identity') +
    scale_x_continuous(limits = c(0, 105)) + 
    xlab("# HPO-coded phenotypes") + ylab("patient count") + 
    ggtitle("before HPO inference") +
    theme(axis.text = element_text(size = 8), title = element_text(size = 10))

  rm(df_temp)
  return (p)
}

plot_patient_distribution_by_n_uniq_abnormal_phenotypes_before_infer <- function(df){
  n_uniq_abnormal_before_infer_per_patient <- df
  n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_before_infer_per_patient %>% 
  mutate(n_uniq_abnormal_before_infer = as.integer(n_uniq_abnormal_before_infer), n_patient = as.integer(n_patient))

  p <- ggplot(n_uniq_abnormal_before_infer_per_patient) + 
    geom_bar(aes(x=n_uniq_abnormal_before_infer, y = n_patient), stat='identity') +
    scale_x_continuous(limits = c(0, 105)) +
    xlab("# HPO-coded phenotypes (unique)") + ylab("patient count") + 
    ggtitle("before HPO inference") +
    theme(axis.text = element_text(size = 8), title = element_text(size = 10))

  return (p)
}
```

```{r}
results_before_infer <- lapply(lab_tables, function(x) abnormal_phenotypes_before_inference(x))
```

Median of # of abnormal phenotypes before running HPO inference
```{r}
as.data.frame(list(lab_source = lab_tables, median_abnormal = sapply(results_before_infer, function(x) x$n_abnormal_before_infer), median_uniq_abnormal = sapply(results_before_infer, function(x) x$n_uniq_abnormal_before_infer)) )
```

What are most prevalent abnormal phenotypes? "Prevalent" is defined as unique number of patients having an abnormal phenotype. 
```{sql connection=dsca, max.output=-1}
SELECT l.hpotermid, hpo.label, l.n :: INTEGER
FROM 
      (SELECT hpotermid , count(distinct medical_record_number) as n
      FROM denormalized_scc_and_epic_lab_2020july
      WHERE hpotermid is not null and not isnegated
      group by hpotermid 
      order by n desc
      LIMIT 50) as l
LEFT JOIN hai_az_prod.hpo hpo on l.hpotermid = hpo.termid ;
```

### Summary statistics after running HPO inference

Count the number of HPO phenotypes per patient after inference. Note inference was only done for abnormal labs. It is tricky to infer based on normal findings.

```{r}
abnormal_phenotypes_after_inference <- function(lab_table_name){
  results <- list()
  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_with_inferred_per_patient;')
  sql <- glue::glue('create temp table n_abnormal_with_inferred_per_patient AS
with abnormal_lab as 
	(SELECT * FROM {lab_table_name} where hpotermid is not null and not isnegated),
abnormal_lab_after_infer as 
	(SELECT l.medical_record_number, l.lab_date, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance 
	FROM abnormal_lab l 
	left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair."current")
SELECT medical_record_number, count(*) as n_abnormal_with_inferred, count(DISTINCT ancestor) as n_uniq_abnormal_with_inferred
FROM abnormal_lab_after_infer 
group by medical_record_number;')
  dbExecute(dsca, sql)    
  # median of HPO-coded phenotypes per patient from abnormal lab findings
  median_abnormal_hpo <- dbGetQuery(dsca, 'SELECT median(n_abnormal_with_inferred) FROM n_abnormal_with_inferred_per_patient;')[1,1]
  results$n_abnormal_after_infer <- median_abnormal_hpo

  # median of unique HPO-coded phenotypes per patient from abnormal lab findings
  median_uniq_abnormal_after_infer <- dbGetQuery(dsca, 'SELECT median(n_uniq_abnormal_with_inferred) FROM n_abnormal_with_inferred_per_patient;')[1,1]
  results$n_uniq_abnormal_after_infer <- median_uniq_abnormal_after_infer

  n_abnormal_after_infer_per_patient <- dbGetQuery(dsca, 
        'SELECT n_abnormal_with_inferred, count(*) as n_patient 
        FROM n_abnormal_with_inferred_per_patient 
        group by n_abnormal_with_inferred 
        order by n_abnormal_with_inferred;')
  results$n_abnormal_before_infer_per_patient <- n_abnormal_after_infer_per_patient

  n_uniq_abnormal_after_infer_per_patient <- dbGetQuery(dsca, 
          'SELECT n_uniq_abnormal_with_inferred, count(*) as n_patient 
          FROM n_abnormal_with_inferred_per_patient 
          group by n_uniq_abnormal_with_inferred 
          order by n_uniq_abnormal_with_inferred;')
  results$n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_after_infer_per_patient

  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_with_inferred_per_patient;')

  return (results)
}

# the input is the third item in the above function
plot_patient_distribution_by_n_abnormal_phenotypes_after_infer <- function(df){
  n_abnormal_with_inferred_per_patient <- df
  n_abnormal_with_inferred_per_patient <- n_abnormal_with_inferred_per_patient %>% mutate(n_abnormal_with_inferred = as.integer(n_abnormal_with_inferred), n_patient = as.integer(n_patient))
n_abnormal_with_inferred_per_patient %>% head()

  MAX_N_ABNORMAL_WITH_INFERRED <- max(n_abnormal_with_inferred_per_patient$n_abnormal_with_inferred)
  df_temp <- n_abnormal_with_inferred_per_patient %>% mutate(strata = cut(n_abnormal_with_inferred_per_patient$n_abnormal_with_inferred, breaks = c(seq(0,100, 10), seq(101, 1001, 100), seq(1002, MAX_N_ABNORMAL_WITH_INFERRED + 30000, 30000)))) %>% group_by(strata) %>% summarise(n_patient = sum(n_patient))

  p <- ggplot(df_temp) +
    geom_bar(aes(x = strata, y = n_patient), stat = 'identity') + theme(axis.text.x = element_text(angle = 30)) +
    xlab("# HPO-coded phenotypes") + ylab("patient count") + 
    ggtitle("after HPO inference") +
    theme(axis.text = element_text(size = 4), title = element_text(size = 10))

  rm(df_temp)
  return (p)
}

plot_patient_distribution_by_n_uniq_abnormal_phenotypes_after_infer <- function(df){
  n_uniq_abnormal_with_inferred_per_patient <- df
  n_uniq_abnormal_with_inferred_per_patient <- n_uniq_abnormal_with_inferred_per_patient %>% mutate(n_uniq_abnormal_with_inferred = as.integer(n_uniq_abnormal_with_inferred), n_patient = as.integer(n_patient))

  MAX_N_ABNORMAL_WITH_INFERRED <- max(n_uniq_abnormal_with_inferred_per_patient$n_uniq_abnormal_with_inferred)
  df_temp <- n_uniq_abnormal_with_inferred_per_patient %>% mutate(strata = cut(n_uniq_abnormal_with_inferred, breaks = c(seq(0,100, 10), seq(101, 1001, 100), seq(1002, MAX_N_ABNORMAL_WITH_INFERRED + 30000, 30000)))) %>% group_by(strata) %>% summarise(n_patient = sum(n_patient))

  p <- ggplot(df_temp) +
    geom_bar(aes(x = strata, y = n_patient), stat = 'identity') + theme(axis.text.x = element_text(angle = 45)) +
    xlab("# HPO-coded phenotypes (unique)") + ylab("patient count") + 
    ggtitle("after HPO inference") +
    theme(axis.text = element_text(size = 4), title = element_text(size = 10))
  rm(df_temp)

  return (p)
}
```

```{r}
results_after_infer <- lapply(lab_tables, abnormal_phenotypes_after_inference)
```

```{r}
as.data.frame(list(lab_source = lab_tables, median_abnormal = sapply(results_after_infer, function(x) x$n_abnormal_after_infer), median_uniq_abnormal = sapply(results_after_infer, function(x) x$n_uniq_abnormal_after_infer)) )
```

### Show patient distribution plots before and after inference on one plot

```{r}
# input is the list item in results_before_infer, and results_after_infer
plot_grid <- function(result_before_infer, result_after_infer){
  # patient distribution ~ abnormal phenotypes before inference
  df_abnormal_before <- result_before_infer[[3]]
  p1 <- plot_patient_distribution_by_n_abnormal_phenotypes_before_infer(df_abnormal_before)
  # patient distribution ~ uniq abnormal phenotypes before inference
  df_uniq_abnormal_before <- result_before_infer[[4]]
  p2 <- plot_patient_distribution_by_n_uniq_abnormal_phenotypes_before_infer(df_uniq_abnormal_before)
  # patient distribution ~ abnormal phenotypes after inference
  df_abnormal_after <- result_after_infer[[3]]
  p3 <- plot_patient_distribution_by_n_abnormal_phenotypes_after_infer(df_abnormal_after)
  # patient distribution ~ uniq abnormal phenotypes after inference
  df_uniq_abnormal_after <- result_after_infer[[4]]
  p4 <- plot_patient_distribution_by_n_uniq_abnormal_phenotypes_after_infer(df_uniq_abnormal_after)

  width = unit(8, "cm")
  height = unit(3, "cm")
  p <- gridExtra::grid.arrange(
    set_panel_size(p1, width = width, height = height),
    set_panel_size(p2, width = width, height = height),
    set_panel_size(p3, width = width, height = height),
    set_panel_size(p4, width = width, height = height),
    ncol=2)

  return (p)
}
```

Patient distribution by looking at SCC labs alone
```{r}
# Lab_SCC
plot_grid(results_before_infer[[1]], results_after_infer[[1]])
```

Patient distribution by looking at EPIC labs alone
```{r}
# Lab_EPIC
plot_grid(results_before_infer[[2]], results_after_infer[[2]])
```

Patient distribution by looking at all labs combined
```{r}
# Lab_SCC_and_EPIC
plot_grid(results_before_infer[[3]], results_after_infer[[3]])
```

We can also look at the most prevalent phenotypes after inference. It is not fair to compare all labs because parent terms are always going to be more prevalent. So we just look at level 1 terms (terms right below "Phenotypic Abnormality").
Look at patients at level = 1 (right below phenotypic abnormality)
```{sql connection=dsca, max.output=-1}
with abnormal_lab as 
	(SELECT * FROM denormalized_scc_and_epic_lab_2020july where hpotermid is not null and not isnegated),
abnormal_lab_after_infer as 
	(SELECT l.medical_record_number, l.lab_date, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance , hpo.distancetoroot, hpo.label
	FROM abnormal_lab l 
	left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair."current"
	left join hai_az_prod.hpo hpo on pair.ancestor = hpo.termid)
SELECT ancestor, label, count(distinct medical_record_number) :: INTEGER as n_patient FROM abnormal_lab_after_infer where distancetoroot = 1 group by ancestor, label
ORDER BY n_patient DESC;
```

```{sql connection=dsca}
DROP VIEW IF EXISTS denormalized_scc_and_epic_lab_2020july;
```

### Analyze the tested phenotypes

Whether a lab test is ordered or not can sometimes be very informative. In this section, we will analyze the most tested phenotypes.

Some phenotypes can be directly tested, such as "Increased LDL cholesterol concentration". Other phenotypes can be inferred from directly testable phenotypes, such as "Abnormal Leukocyte count", which are inferred from specific tests on different blood cell counts. We collect the count of phenotypes directly tested, or including indirectly tested ones as well.

```{r}
CACHE_DIR = 'cache'
if (!dir.exists(CACHE_DIR)){
  dir.create(CACHE_DIR)
}

get_tested_phenotype_rank_direct <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_phenotype_rank_direct.csv')

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
      with testable_phenotypes_direct as (
    	select lab.medical_record_number , lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly 
    	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
    	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
    	where testable.is_tested_directly 
    ),
    testable_phenotype_direct_freq as (
    	select termid, count(*) as n_record, count(distinct medical_record_number) as n_patient
    	from testable_phenotypes_direct
    	group by termid 
    )
    select testable.*, hpo.label
    from testable_phenotype_direct_freq testable
    join hai_az_prod.hpo hpo on testable.termid = hpo.termid
    order by n_record desc;")

    write.csv(df, file = df_path, row.names=FALSE)
  }

  df <- read.csv(df_path)

  return (df)
}

get_tested_phenotype_rank_all <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_phenotype_rank_all.csv')

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
               with testable_phenotypes_direct as (
              	select lab.medical_record_number , lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly 
              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
              	--where testable.is_tested_directly 
              ),
              testable_phenotype_direct_freq as (
              	select termid, count(*) as n_record, count(distinct medical_record_number) as n_patient
              	from testable_phenotypes_direct
              	group by termid 
              )
              select testable.*, hpo.label, hpo.distancetoroot 
              from testable_phenotype_direct_freq testable
              join hai_az_prod.hpo hpo on testable.termid = hpo.termid
              order by n_record desc;
              ")

    write.csv(df, file = df_path, row.names=FALSE)
  }

  df <- read.csv(df_path)

  return (df)
}

tested_phenotype_rank_direct <- get_tested_phenotype_rank_direct()
tested_phenotype_rank_all <- get_tested_phenotype_rank_all()
```

For the directly tested phenotypes, we can observe what are the most tested abnormalities in one healthcare system. 
```{r}
tested_phenotype_rank_direct %>% head(10) %>% formattable::formattable()
```

For the directly/indirectly tested phenotypes, we can determine at the higher level, what are the most tested abnormalities. 
```{r max.output=-1}
tested_phenotype_rank_all %>% filter(distancetoroot == 1) %>% arrange(-n_record) %>% select(-distancetoroot)
```

For both cases, it would be interesting to learn about the changes of the lab phenotypes over time, and among different races.

```{r}
get_tested_phenotype_rank_direct_by_race <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_phenotype_rank_direct_by_race.csv')

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
              with testable_phenotypes_direct as (
            	select lab.medical_record_number , person.race, lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly 
            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
            	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
            	join hai_az_prod.person_msdw_2020july person using (medical_record_number)
            	where testable.is_tested_directly 
            ),
            testable_phenotype_direct_freq as (
            	select termid, race, count(*) as n_record, count(distinct medical_record_number) as n_patient
            	from testable_phenotypes_direct
            	group by termid, race
            )
            select testable.*, hpo.label
            from testable_phenotype_direct_freq testable
            join hai_az_prod.hpo hpo on testable.termid = hpo.termid
            order by termid, race;")

    write.csv(df, file = df_path, row.names=FALSE)
  }

  df <- read.csv(df_path)

  return (df)
}

get_tested_phenotype_rank_all_by_race <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_phenotype_rank_all_by_race.csv')

  if (force_rerun | !file.exists(df_path)){
    df <-  dbGetQuery(dsca, "
              with testable_phenotypes_direct as (
              	select lab.medical_record_number , person.race, lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly 
              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
              	join hai_az_prod.person_msdw_2020july person using (medical_record_number)
              	--where testable.is_tested_directly 
              ),
              testable_phenotype_direct_freq as (
              	select termid, race, count(*) as n_record, count(distinct medical_record_number) as n_patient
              	from testable_phenotypes_direct
              	group by termid, race
              )
              select testable.*, hpo.label, hpo.distancetoroot
              from testable_phenotype_direct_freq testable
              join hai_az_prod.hpo hpo on testable.termid = hpo.termid
              order by termid, race;")

    write.csv(df, file = df_path, row.names=FALSE)
  }

  df <- read.csv(df_path)

  return (df)
}

tested_phenotype_rank_direct_by_race <- get_tested_phenotype_rank_direct_by_race()
tested_phenotype_rank_all_by_race <- get_tested_phenotype_rank_all_by_race()
```

For each directly measured phenotype, we can determine whether race and the tests are correlated by some form of Chi-squared teste.
```{r}
tested_phenotype_rank_direct_by_race %>% filter(race == 'African American') %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
tested_phenotype_rank_direct_by_race %>% filter(race == 'Asian') %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
tested_phenotype_rank_direct_by_race %>% filter(race == 'White') %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
```

```{r}
tested_phenotype_rank_all_by_race %>% filter(race == 'African American' & distancetoroot == 1) %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
tested_phenotype_rank_all_by_race %>% filter(race == 'Asian' & distancetoroot == 1) %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
tested_phenotype_rank_all_by_race %>% filter(race == 'White' & distancetoroot == 1) %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
```

Plot the anual change of tested phenotypes.

```{r}
get_tested_phenotype_rank_all_by_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_phenotype_rank_all_by_year.csv')

  if (force_rerun | !file.exists(df_path)){
    df <-  dbGetQuery(dsca, "
              with testable_phenotypes_direct as (
              	select lab.medical_record_number , person.race, lab.context_name , lab.lab_date , lab.lab_time , date_part(year, lab_date) as lab_year, lab.loinc , testable.termid , testable.is_tested_directly 
              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
              	join hai_az_prod.person_msdw_2020july person using (medical_record_number)
              	--where testable.is_tested_directly 
              ),
              testable_phenotype_direct_freq as (
              	select termid, lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
              	from testable_phenotypes_direct
              	group by termid, lab_year
              )
              select testable.*, hpo.label, hpo.distancetoroot 
              from testable_phenotype_direct_freq testable
              join hai_az_prod.hpo hpo on testable.termid = hpo.termid
              order by termid, lab_year;
              ")

    write.csv(df, file = df_path, row.names=FALSE)
  }

  df <- read.csv(df_path)

  return (df)
}

tested_phenotype_rank_all_by_year <- get_tested_phenotype_rank_all_by_year()

plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df = tested_phenotype_rank_all_by_year){
  hpo_terms_selected <- df %>% filter(termid %in% terms_of_interest) %>% select(termid, label) %>% distinct()

  p <- df %>% filter(termid %in% terms_of_interest) %>% 
    filter(lab_year <= 2019 & lab_year > 2000) %>%
    ggplot(aes(x = lab_year, y = n_patient, group = termid, color = termid)) + 
    geom_point() +
    geom_line() +
    scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " "))

  return (p)
}
```

```{r}
hpo_level_1_terms <- tested_phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid)
```

Some trends are pretty interesting, such as the spike of Abnormal cellular phenotype. The term has three children, but only one is tested: HP:0011017 Abnormal cellular physiology

```{r}
children_abnormal_cellular_physiology <- c('HP:0003220', 'HP:0011018', 'HP:0025463', 'HP:0011019', 'HP:0003254', 'HP:4000056', 'HP:0011133', 'HP:0004371', 'HP:0031377', 'HP:0031409', 'HP:0012103', 'HP:0032576', 'HP:0003204', 'HP:0004356', 'HP:0032571', 'HP:0002916', 'HP:0003224', 'HP:0003575', 'HP:0003358')
plot_phenotypes_of_interest_by_year(terms_of_interest = children_abnormal_cellular_physiology)
```

Again, the above plot shows that it is driven by one HP:0031409 Abnormal lymphocyte physiology. 
```{r}
children_of_interest <- c('HP:0011840', 'HP:0030886', 'HP:0012177', 'HP:0005372', 'HP:0031378')
plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest)
```
```{r}
children_of_interest <- c('HP:0002847', 'HP:0002959', 'HP:0005384', 'HP:0041076', 'HP:0010701')
plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest)
```

```{r}
children_of_interest <- c('HP:0010702', 'HP:0032336', 'HP:0410241', 'HP:0410240', 'HP:0410243', 'HP:0410242', 'HP:0410244', 'HP:0004313')
plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest)
```

```{r}
children_of_interest <- c('HP:0410367', 'HP:0031047', 'HP:0003237', 'HP:0003212', 'HP:0003261', 'HP:0410369', 'HP:0410371', 'HP:0410246', 'HP:0003496')
plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest)
```

### 2021-11-16 build.py

@@ -1,0 +1,16 @@
+#   -*- coding: utf-8 -*-
+from pybuilder.core import use_plugin, init
+
+use_plugin("python.core")
+use_plugin("python.unittest")
+use_plugin("python.flake8")
+use_plugin("python.distutils")
+
+
+name = "s4-loinc2hpo-msdw"
+default_task = "publish"
+
+
+@init
+def set_properties(project):
+    pass

### 2021-11-23 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
phenotypic diversity analysis

@@
 
-### Prepare the lab tables
+```{r}
+CACHE_DIR = 'cache'
+if (!dir.exists(CACHE_DIR)){
+  dir.create(CACHE_DIR)
+}
 
-Run the following script to create the denormalized table for lab_scc and lab_epic.
-After this step, we should have three tables for SCC lab, EPIC lab, and combined.
-And each table should have LOINC and mapped HPO.
-```{r}
-# sql_folder_path <- '/Users/xingminzhang/git/s4-pd-dmsdw/pd-dmsdw-loinc2hpo-app/src/main/resources/sql'
-# sql_denormalized_scc_lab_2020july <- read_file(glue('{sql_folder_path}/loinc2hpo_SCC_MSDW.sql'))
-# sql_denormalized_epic_lab_2020july <- read_file(glue('{sql_folder_path}/loinc2hpo_Epic_MSDW.sql'))
-# sql_combine_Epic_SCC <- read_file(glue('{sql_folder_path}/loinc2hpo_combine_Epic_SCC.sql')) 
-# 
-# # make sure the lab table after LOINC2HPO transformation exists
-# # if not, we can create it here; but the intended behavior is that we use the Java commandline app for this
+get_db_data <- function(sql, df_path, force_rerun=FALSE, cache_dir=CACHE_DIR){
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, sql)
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
 
+  df <- read.csv(df_path)
+
+  return (df)
+}
 ```
 
-```{sql connection=dsca, echo=FALSE}
-drop table if exists denormalized_scc_lab_2020july;
-```
-```{sql connection=dsca}
-CREATE TEMP TABLE denormalized_scc_lab_2020july as
-WITH 
-	lab_scc_2020q2 as (
-		SELECT l.medical_record_number , l.encounter_key , l.calendar_date , l.time_of_day_key , l.clinical_result_numeric ,  trim(from l.unit_of_measure_numeric) as unit_of_measure_numeric , l.reference_range , l.procedure_key,
-		case when l.abnormal_flag is null then 'N' else l.abnormal_flag end as abnormal_flag 
-		FROM hai_az_prod.lab_scc_2020july l
-		--where medical_record_number in (2966097, 3164683, 3869795)
-	), 
-	lab_scc_abnormal_flag_mapping AS (
-		SELECT case when abf.abnormal_flag is NULL then 'N' else abf.abnormal_flag end as abnormal_flag, abf.mapto 
-		FROM hai_az_prod.lab_scc_abnormal_flag_mapping abf
+### Analyze changes of observed phenotypes
+We will used the persisted lab test table for this analysis. 
+```{r}
+get_phenotype_rank_all_by_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/observed_phenotype_rank_all_by_year.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, "
+      with abnormal_lab as 
+      	(select * from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab where hpotermid is not null and not isnegated),
+      abnormal_lab_after_infer as 
+      	(SELECT l.medical_record_number, l.lab_date, date_part(year, lab_date) as lab_year, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance
+      	FROM abnormal_lab l 
+      	left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair.current),
+      lab_freq_by_year as (
+      	select ancestor as termid, lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient from abnormal_lab_after_infer group by ancestor, lab_year
       )
-SELECT lab.medical_record_number, lab.encounter_key , lab.calendar_date  as lab_date, dtod.clock_time_24_hour || ':00' as lab_time, lab.clinical_result_numeric , fdp.context_name, fdp.context_procedure_code AS local_test_code , loinc.loinc , lab.clinical_result_numeric AS lab_result_numeric_value, lab.unit_of_measure_numeric As unit_of_measure ,lab.reference_range AS normal_range , lab.abnormal_flag , abf.mapto AS interpretation, l2h.hpotermid , l2h.isnegated
-FROM lab_scc_2020q2 lab 
-left JOIN hai_az_prod.fd_procedure fdp using (procedure_key)
-left JOIN hai_az_prod.loinc_mapping loinc on fdp.context_procedure_code = loinc.code and (lab.unit_of_measure_numeric = loinc.unit or (lab.unit_of_measure_numeric is NULL and loinc.unit is NULL))
-left JOIN lab_scc_abnormal_flag_mapping abf using (abnormal_flag)
-left JOIN hai_az_prod.loinc2hpo l2h on loinc.loinc = l2h.loincid and abf.mapto = l2h.code
-join prod_msdw.d_time_of_day dtod using (time_of_day_key);
-```
+      select *
+      from lab_freq_by_year
+      join hai_az_prod.hpo h using (termid)
+      order by lab_year, distancetoroot;")
 
-```{sql connection=dsca}
-DROP TABLE IF EXISTS denormalized_epic_lab_2020july;
-```
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
 
-```{sql connection=dsca}
-create temp table denormalized_epic_lab_2020july as
-with interpreted as (
-	SELECT lab.mrn, lab.order_date , lab.lab_time , lab.test_code , lab.test_result_value , lab.unit_of_measurement , lab.reference_range ,
-	case 
-			when test_result_value_final :: REAL < range_low_final ::REAL then 'L' 
-			when test_result_value_final :: REAL > range_hi_final :: REAL then 'H'
-			else 'N' 
-	end as interpretation
-	FROM hai_az_test.lab_epic_2020july lab
-	--where mrn in (3710567, 8014250, 3144637)
-	)
-SELECT lab.mrn AS medical_record_number, NULL AS encounter_key, lab.order_date as lab_date, lab.lab_time, 'EPIC' AS context_name, lab.test_code AS local_test_code , local2loinc.loinc , lab.test_result_value AS lab_result_numeric_value, lab.unit_of_measurement AS unit_of_measure , lab.reference_range AS normal_range , NULL AS abnormal_flag,  interpretation,
-loinc2hpo.hpotermid , loinc2hpo.isnegated 
-FROM interpreted lab
-LEFT JOIN hai_az_prod.loinc_mapping local2loinc on lab.test_code :: VARCHAR = local2loinc.code and (lab.unit_of_measurement = local2loinc.unit OR (lab.unit_of_measurement is NULL and local2loinc.unit is NULL))
-LEFT JOIN hai_az_prod.loinc2hpo loinc2hpo on local2loinc.loinc = loinc2hpo.loincid and interpretation = loinc2hpo.code 
-where local2loinc."source" = 'EPIC'; 
-```
+  df <- read.csv(df_path)
 
-Join the above tables together
-```{sql connection=dsca}
-CREATE VIEW denormalized_scc_and_epic_lab_2020july AS
-WITH epic_denormalized AS (
-  SELECT medical_record_number, lab_date, lab_time, context_name, local_test_code :: VARCHAR, loinc,  lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated
-  FROM denormalized_epic_lab_2020july
-), 
-scc_denormalized AS (
-  SELECT medical_record_number, lab_date, lab_time, context_name, local_test_code, loinc, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated
-  FROM denormalized_scc_lab_2020july
-)
-SELECT * FROM scc_denormalized 
-UNION all 
-SELECT * FROM epic_denormalized; 
-```
+  return (df)
+}
 
-Right now, we have three sets of denormalized labs, SCC, EPIC and their union.
-```{sql connection=dsca}
-SELECT * FROM denormalized_scc_lab_2020july limit 4;
-```
-
-```{sql connection=dsca}
-SELECT * FROM denormalized_epic_lab_2020july limit 4;
+phenotype_rank_all_by_year <- get_phenotype_rank_all_by_year()
 ```
@@
 ```{r}
-lab_tables <- c('denormalized_scc_lab_2020july', 'denormalized_epic_lab_2020july', 'denormalized_scc_and_epic_lab_2020july')
-```
+plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df){
+  hpo_terms_selected <- df %>% filter(termid %in% terms_of_interest) %>% select(termid, label) %>% distinct()
 
-### Summary statistics before running any HPO inference
+  p <- df %>% filter(termid %in% terms_of_interest) %>% 
+    filter(lab_year <= 2019 & lab_year > 2000) %>%
+    ggplot(aes(x = lab_year, y = n_patient, group = termid, color = termid)) + 
+    geom_point() +
+    geom_line() +
+    scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " "))
 
-```{r}
-abnormal_phenotypes_before_inference <- function(lab_table_name){
-  results <- list()
-  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_before_infer_per_patient;')
-  sql <- glue::glue('create temp table n_abnormal_before_infer_per_patient AS
-        SELECT medical_record_number , count(*) as n_abnormal_before_infer, count(distinct hpotermid ) as n_uniq_abnormal_before_infer
-        FROM {lab_table_name}
-        where hpotermid is not null and not isnegated
-        group by medical_record_number ;')
-  dbExecute(dsca, sql)    
-  # median of HPO-coded phenotypes per patient from abnormal lab findings
-  median_abnormal_hpo <- dbGetQuery(dsca, 'SELECT median(n_abnormal_before_infer) FROM n_abnormal_before_infer_per_patient;')[1,1]
-  results$n_abnormal_before_infer <- median_abnormal_hpo
-
-  # median of unique HPO-coded phenotypes per patient from abnormal lab findings
-  median_uniq_abnormal_before_infer <- dbGetQuery(dsca, 'SELECT median(n_uniq_abnormal_before_infer) FROM n_abnormal_before_infer_per_patient;')[1,1]
-  results$n_uniq_abnormal_before_infer <- median_uniq_abnormal_before_infer
-
-  n_abnormal_before_infer_per_patient <- dbGetQuery(dsca, 
-        'SELECT n_abnormal_before_infer, count(*) as n_patient 
-      FROM n_abnormal_before_infer_per_patient
-      GROUP BY n_abnormal_before_infer
-      order by n_abnormal_before_infer;')
-  results$n_abnormal_before_infer_per_patient <- n_abnormal_before_infer_per_patient
-
-  n_uniq_abnormal_before_infer_per_patient <- dbGetQuery(dsca, 
-          'SELECT n_uniq_abnormal_before_infer, count(*) as n_patient 
-            FROM n_abnormal_before_infer_per_patient
-            GROUP BY n_uniq_abnormal_before_infer
-            order by n_uniq_abnormal_before_infer;')
-  results$n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_before_infer_per_patient
-
-  return (results)
-}
-
-# the input is the third item in the above function
-plot_patient_distribution_by_n_abnormal_phenotypes_before_infer <- function(df){
-  n_abnormal_before_infer_per_patient <- df
-  n_abnormal_before_infer_per_patient <- n_abnormal_before_infer_per_patient %>% mutate(n_abnormal_before_infer = as.integer(n_abnormal_before_infer), n_patient = as.integer(n_patient))
-
-## roll up the ones > 100
-  df_temp <- n_abnormal_before_infer_per_patient %>% mutate(n_abnormal_before_infer = ifelse(n_abnormal_before_infer > 100, 100, n_abnormal_before_infer)) %>% group_by(n_abnormal_before_infer) %>% summarise(n_patient = sum(n_patient))
-
-  p <- ggplot(df_temp) + 
-    geom_bar(aes(x=n_abnormal_before_infer, y = n_patient), stat='identity') +
-    scale_x_continuous(limits = c(0, 105)) + 
-    xlab("# HPO-coded phenotypes") + ylab("patient count") + 
-    ggtitle("before HPO inference") +
-    theme(axis.text = element_text(size = 8), title = element_text(size = 10))
-
-  rm(df_temp)
   return (p)
@@
 
-plot_patient_distribution_by_n_uniq_abnormal_phenotypes_before_infer <- function(df){
-  n_uniq_abnormal_before_infer_per_patient <- df
-  n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_before_infer_per_patient %>% 
-  mutate(n_uniq_abnormal_before_infer = as.integer(n_uniq_abnormal_before_infer), n_patient = as.integer(n_patient))
-
-  p <- ggplot(n_uniq_abnormal_before_infer_per_patient) + 
-    geom_bar(aes(x=n_uniq_abnormal_before_infer, y = n_patient), stat='identity') +
-    scale_x_continuous(limits = c(0, 105)) +
-    xlab("# HPO-coded phenotypes (unique)") + ylab("patient count") + 
-    ggtitle("before HPO inference") +
-    theme(axis.text = element_text(size = 8), title = element_text(size = 10))
-
-  return (p)
-}
-```
-
-```{r}
-results_before_infer <- lapply(lab_tables, function(x) abnormal_phenotypes_before_inference(x))
-```
-
-Median of # of abnormal phenotypes before running HPO inference
-```{r}
-as.data.frame(list(lab_source = lab_tables, median_abnormal = sapply(results_before_infer, function(x) x$n_abnormal_before_infer), median_uniq_abnormal = sapply(results_before_infer, function(x) x$n_uniq_abnormal_before_infer)) )
-```
-
-What are most prevalent abnormal phenotypes? "Prevalent" is defined as unique number of patients having an abnormal phenotype. 
-```{sql connection=dsca, max.output=-1}
-SELECT l.hpotermid, hpo.label, l.n :: INTEGER
-FROM 
-      (SELECT hpotermid , count(distinct medical_record_number) as n
-      FROM denormalized_scc_and_epic_lab_2020july
-      WHERE hpotermid is not null and not isnegated
-      group by hpotermid 
-      order by n desc
-      LIMIT 50) as l
-LEFT JOIN hai_az_prod.hpo hpo on l.hpotermid = hpo.termid ;
-```
-
-### Summary statistics after running HPO inference
-
-Count the number of HPO phenotypes per patient after inference. Note inference was only done for abnormal labs. It is tricky to infer based on normal findings.
-
-```{r}
-abnormal_phenotypes_after_inference <- function(lab_table_name){
-  results <- list()
-  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_with_inferred_per_patient;')
-  sql <- glue::glue('create temp table n_abnormal_with_inferred_per_patient AS
-with abnormal_lab as 
-	(SELECT * FROM {lab_table_name} where hpotermid is not null and not isnegated),
-abnormal_lab_after_infer as 
-	(SELECT l.medical_record_number, l.lab_date, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance 
-	FROM abnormal_lab l 
-	left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair."current")
-SELECT medical_record_number, count(*) as n_abnormal_with_inferred, count(DISTINCT ancestor) as n_uniq_abnormal_with_inferred
-FROM abnormal_lab_after_infer 
-group by medical_record_number;')
-  dbExecute(dsca, sql)    
-  # median of HPO-coded phenotypes per patient from abnormal lab findings
-  median_abnormal_hpo <- dbGetQuery(dsca, 'SELECT median(n_abnormal_with_inferred) FROM n_abnormal_with_inferred_per_patient;')[1,1]
-  results$n_abnormal_after_infer <- median_abnormal_hpo
-
-  # median of unique HPO-coded phenotypes per patient from abnormal lab findings
-  median_uniq_abnormal_after_infer <- dbGetQuery(dsca, 'SELECT median(n_uniq_abnormal_with_inferred) FROM n_abnormal_with_inferred_per_patient;')[1,1]
-  results$n_uniq_abnormal_after_infer <- median_uniq_abnormal_after_infer
-
-  n_abnormal_after_infer_per_patient <- dbGetQuery(dsca, 
-        'SELECT n_abnormal_with_inferred, count(*) as n_patient 
-        FROM n_abnormal_with_inferred_per_patient 
-        group by n_abnormal_with_inferred 
-        order by n_abnormal_with_inferred;')
-  results$n_abnormal_before_infer_per_patient <- n_abnormal_after_infer_per_patient
-
-  n_uniq_abnormal_after_infer_per_patient <- dbGetQuery(dsca, 
-          'SELECT n_uniq_abnormal_with_inferred, count(*) as n_patient 
-          FROM n_abnormal_with_inferred_per_patient 
-          group by n_uniq_abnormal_with_inferred 
-          order by n_uniq_abnormal_with_inferred;')
-  results$n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_after_infer_per_patient
-
-  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_with_inferred_per_patient;')
-
-  return (results)
-}
-
-# the input is the third item in the above function
-plot_patient_distribution_by_n_abnormal_phenotypes_after_infer <- function(df){
-  n_abnormal_with_inferred_per_patient <- df
-  n_abnormal_with_inferred_per_patient <- n_abnormal_with_inferred_per_patient %>% mutate(n_abnormal_with_inferred = as.integer(n_abnormal_with_inferred), n_patient = as.integer(n_patient))
-n_abnormal_with_inferred_per_patient %>% head()
-
-  MAX_N_ABNORMAL_WITH_INFERRED <- max(n_abnormal_with_inferred_per_patient$n_abnormal_with_inferred)
-  df_temp <- n_abnormal_with_inferred_per_patient %>% mutate(strata = cut(n_abnormal_with_inferred_per_patient$n_abnormal_with_inferred, breaks = c(seq(0,100, 10), seq(101, 1001, 100), seq(1002, MAX_N_ABNORMAL_WITH_INFERRED + 30000, 30000)))) %>% group_by(strata) %>% summarise(n_patient = sum(n_patient))
-
-  p <- ggplot(df_temp) +
-    geom_bar(aes(x = strata, y = n_patient), stat = 'identity') + theme(axis.text.x = element_text(angle = 30)) +
-    xlab("# HPO-coded phenotypes") + ylab("patient count") + 
-    ggtitle("after HPO inference") +
-    theme(axis.text = element_text(size = 4), title = element_text(size = 10))
-
-  rm(df_temp)
-  return (p)
-}
-
-plot_patient_distribution_by_n_uniq_abnormal_phenotypes_after_infer <- function(df){
-  n_uniq_abnormal_with_inferred_per_patient <- df
-  n_uniq_abnormal_with_inferred_per_patient <- n_uniq_abnormal_with_inferred_per_patient %>% mutate(n_uniq_abnormal_with_inferred = as.integer(n_uniq_abnormal_with_inferred), n_patient = as.integer(n_patient))
-
-  MAX_N_ABNORMAL_WITH_INFERRED <- max(n_uniq_abnormal_with_inferred_per_patient$n_uniq_abnormal_with_inferred)
-  df_temp <- n_uniq_abnormal_with_inferred_per_patient %>% mutate(strata = cut(n_uniq_abnormal_with_inferred, breaks = c(seq(0,100, 10), seq(101, 1001, 100), seq(1002, MAX_N_ABNORMAL_WITH_INFERRED + 30000, 30000)))) %>% group_by(strata) %>% summarise(n_patient = sum(n_patient))
-
-  p <- ggplot(df_temp) +
-    geom_bar(aes(x = strata, y = n_patient), stat = 'identity') + theme(axis.text.x = element_text(angle = 45)) +
-    xlab("# HPO-coded phenotypes (unique)") + ylab("patient count") + 
-    ggtitle("after HPO inference") +
-    theme(axis.text = element_text(size = 4), title = element_text(size = 10))
-  rm(df_temp)
-
-  return (p)
-}
-```
-
-```{r}
-results_after_infer <- lapply(lab_tables, abnormal_phenotypes_after_inference)
-```
-
-```{r}
-as.data.frame(list(lab_source = lab_tables, median_abnormal = sapply(results_after_infer, function(x) x$n_abnormal_after_infer), median_uniq_abnormal = sapply(results_after_infer, function(x) x$n_uniq_abnormal_after_infer)) )
-```
-
-### Show patient distribution plots before and after inference on one plot
-
-```{r}
-# input is the list item in results_before_infer, and results_after_infer
-plot_grid <- function(result_before_infer, result_after_infer){
-  # patient distribution ~ abnormal phenotypes before inference
-  df_abnormal_before <- result_before_infer[[3]]
-  p1 <- plot_patient_distribution_by_n_abnormal_phenotypes_before_infer(df_abnormal_before)
-  # patient distribution ~ uniq abnormal phenotypes before inference
-  df_uniq_abnormal_before <- result_before_infer[[4]]
-  p2 <- plot_patient_distribution_by_n_uniq_abnormal_phenotypes_before_infer(df_uniq_abnormal_before)
-  # patient distribution ~ abnormal phenotypes after inference
-  df_abnormal_after <- result_after_infer[[3]]
-  p3 <- plot_patient_distribution_by_n_abnormal_phenotypes_after_infer(df_abnormal_after)
-  # patient distribution ~ uniq abnormal phenotypes after inference
-  df_uniq_abnormal_after <- result_after_infer[[4]]
-  p4 <- plot_patient_distribution_by_n_uniq_abnormal_phenotypes_after_infer(df_uniq_abnormal_after)
-
-  width = unit(8, "cm")
-  height = unit(3, "cm")
-  p <- gridExtra::grid.arrange(
-    set_panel_size(p1, width = width, height = height),
-    set_panel_size(p2, width = width, height = height),
-    set_panel_size(p3, width = width, height = height),
-    set_panel_size(p4, width = width, height = height),
-    ncol=2)
-
-  return (p)
-}
-```
-
-Patient distribution by looking at SCC labs alone
-```{r}
-# Lab_SCC
-plot_grid(results_before_infer[[1]], results_after_infer[[1]])
-```
-
-Patient distribution by looking at EPIC labs alone
-```{r}
-# Lab_EPIC
-plot_grid(results_before_infer[[2]], results_after_infer[[2]])
-```
-
-Patient distribution by looking at all labs combined
-```{r}
-# Lab_SCC_and_EPIC
-plot_grid(results_before_infer[[3]], results_after_infer[[3]])
-```
-
-We can also look at the most prevalent phenotypes after inference.
-It is not fair to compare all labs because parent terms are always going to be more prevalent.
-So we just look at level 1 terms (terms right below "Phenotypic Abnormality").
-Look at patients at level = 1 (right below phenotypic abnormality)
-```{sql connection=dsca, max.output=-1}
-with abnormal_lab as 
-	(SELECT * FROM denormalized_scc_and_epic_lab_2020july where hpotermid is not null and not isnegated),
-abnormal_lab_after_infer as 
-	(SELECT l.medical_record_number, l.lab_date, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance , hpo.distancetoroot, hpo.label
-	FROM abnormal_lab l 
-	left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair."current"
-	left join hai_az_prod.hpo hpo on pair.ancestor = hpo.termid)
-SELECT ancestor, label, count(distinct medical_record_number) :: INTEGER as n_patient FROM abnormal_lab_after_infer where distancetoroot = 1 group by ancestor, label
-ORDER BY n_patient DESC;
-```
-
-```{sql connection=dsca}
-DROP VIEW IF EXISTS denormalized_scc_and_epic_lab_2020july;
+hpo_level_1_terms <- phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
+plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=phenotype_rank_all_by_year)
 ```
@@
 ```{r}
-CACHE_DIR = 'cache'
-if (!dir.exists(CACHE_DIR)){
-  dir.create(CACHE_DIR)
-}
-
 get_tested_phenotype_rank_direct <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
@@
 tested_phenotype_rank_all_by_year <- get_tested_phenotype_rank_all_by_year()
-
-plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df = tested_phenotype_rank_all_by_year){
-  hpo_terms_selected <- df %>% filter(termid %in% terms_of_interest) %>% select(termid, label) %>% distinct()
-
-  p <- df %>% filter(termid %in% terms_of_interest) %>% 
-    filter(lab_year <= 2019 & lab_year > 2000) %>%
-    ggplot(aes(x = lab_year, y = n_patient, group = termid, color = termid)) + 
-    geom_point() +
-    geom_line() +
-    scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " "))
-
-  return (p)
-}
 ```
@@
 hpo_level_1_terms <- tested_phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
-plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid)
+plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year)
 ```
@@
 children_abnormal_cellular_physiology <- c('HP:0003220', 'HP:0011018', 'HP:0025463', 'HP:0011019', 'HP:0003254', 'HP:4000056', 'HP:0011133', 'HP:0004371', 'HP:0031377', 'HP:0031409', 'HP:0012103', 'HP:0032576', 'HP:0003204', 'HP:0004356', 'HP:0032571', 'HP:0002916', 'HP:0003224', 'HP:0003575', 'HP:0003358')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_abnormal_cellular_physiology)
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_abnormal_cellular_physiology, df=tested_phenotype_rank_all_by_year)
 ```
@@
 children_of_interest <- c('HP:0011840', 'HP:0030886', 'HP:0012177', 'HP:0005372', 'HP:0031378')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest)
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
 ```
@@
 children_of_interest <- c('HP:0002847', 'HP:0002959', 'HP:0005384', 'HP:0041076', 'HP:0010701')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest)
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
 ```
@@
 children_of_interest <- c('HP:0010702', 'HP:0032336', 'HP:0410241', 'HP:0410240', 'HP:0410243', 'HP:0410242', 'HP:0410244', 'HP:0004313')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest)
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
 ```
@@
 children_of_interest <- c('HP:0410367', 'HP:0031047', 'HP:0003237', 'HP:0003212', 'HP:0003261', 'HP:0410369', 'HP:0410371', 'HP:0410246', 'HP:0003496')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest)
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
 ```

### 2021-12-01 src/main/r/numeric_vs_hpo_for_acute_kidney_failure.Rmd
compare raw and hpo of kidney labs for acute kidney failure

---
title: "R Notebook"
output: html_notebook
---

## Goal
Compare the performance of raw lab values vs HPO interpretations for acute kidney failure.

The analysis process is very similar to the liver function analysis.

Algorithm:
- Define a cohort of patients coded with acute kidney failure as the primay diagnosis of a visit;
- Get all the lab tests with selected LOINC code; 
- For each patient, the labs within X days (e.g. 7 days) before the diagnosis date is considered to be associated with + outcome; and the labs further outside of the window is considered to be associated with - outcome; 
- For each LOINC code, we will prepare a dataframe, separate into training and test set, train a logistic regression classifier using either raw numerical value or HPO value (observed or not observed), build the ROC curve with the test set, and finally compare the ROC curve to determine whether they are statistically different.

## Analysis

### retrieve data
```{r message=FALSE}
library(tidyverse)
library(glue)
library(lubridate)
library(readxl)
require(DBI)
require(RPostgres)
require(egg)
require(gridExtra)
source('.db_credential.R')
source('functions.R')
dsca <- DBI::dbConnect(RPostgres::Postgres(), host='localhost', user=user, dbname = dbname, port = 2345, password=password)

# cache directory to read from or write to
CACHE_DIR <- './cache'
if (!dir.exists(CACHE_DIR)){
  dir.create(CACHE_DIR)
}
```

Create a cohort for patients with abnormal liver function diagnoses.
```{r}
sql <- read_file('../sql/acute_kidney_failure_cohort.sql')
print(sql)
n<- dbExecute(dsca, sql)
```

Retrieve liver function diagnoses information for the cohort.
```{r}
sql <- "SELECT * FROM cohort"
df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'acute_kidney_failure_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)

df_cohort %>% nrow()
```

Retrieve liver function lab tests for the cohort. 
```{r}
sql <- read_file('../sql/get_labs_for_acute_kidney_failure.sql')
df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_acute_kidney_failure.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
df_labs %>% nrow()
```

The first thing is to check how do the raw values of lab test look like before diagnoses of abnormal liver function. We want to choose a time window during which the liver function labs are highly correlated with the diagnosis, and another time window well, which is before the first window and corresponds to a *normal* liver function labs. The plot below shows that roughly we can choose -180 as the time cutoff: within this time cutoff, i.e. (-180, 0) days, the latest lab test tends to be abnormal and correspond to an abnormal liver function diagnosis; outside of this time cutoff, i.e. 180 days before a known liver abnormal function diagnosis, the lab test tends to be normal.  
```{r}
plot_lab_raw_values_vs_rel_date <- function(lab_loinc, ylim=NULL){
  df <- df_labs %>% 
    filter(loinc == lab_loinc) %>% 
    filter(rel_date <= 0 & rel_date > -730) %>% 
    mutate(lab_result_numeric_value = as.numeric(lab_result_numeric_value))

  p <- ggplot(df, aes(x = rel_date, y = lab_result_numeric_value)) +
    geom_point(size = 0.1, alpha=0.1) +
    xlab('lab date (relative to diagnosis)')

  if (!is.null(ylim)){
    p <- p + coord_cartesian(ylim = ylim)
  }

  p <- p + theme_bw() + 
    theme(panel.grid = element_blank(),
          axis.text = element_text(size = 5),
          axis.title = element_text(size = 8), 
          strip.text.x = element_text(margin = margin(0.3, 0, 0.3, 0), size = 5),
          plot.title = element_text(size = 8))

  return (p)
}

# setClass('LiverLab', representation(lab_loinc = 'character', lab_name = 'character', ylim_min = "numeric", ylim_max = "numeric"))
# liver_lab_ALT <- new('LiverLab', lab_loinc = '1742-6', lab_name = 'Alanine aminotransferase', ylim_min = 0, ylim_max = 500)
# liver_lab_GGT <- new('LiverLab', lab_loinc = '2324-2', lab_name = 'Gamma glutamyl transferase', ylim_min = 0, ylim_max = 1000)
# liver_lab_AST <- new('LiverLab', lab_loinc = '1920-8', lab_name = 'Aspartate aminotransferase', ylim_min = 0, ylim_max = 500)
# liver_lab_ALP <- new('LiverLab', lab_loinc = '6768-6', lab_name = 'Alkaline phosphatase', ylim_min = 0, ylim_max = 1000)
# liver_lab_total_protein <- new('LiverLab', lab_loinc = '2885-2', lab_name = 'Total protein', ylim_min = 0, ylim_max = 15)
# liver_lab_serum_albumin <- new('LiverLab', lab_loinc = '1751-7', lab_name = 'Serum albumin', ylim_min = 0, ylim_max = 20)
# liver_lab_PT <- new('LiverLab', lab_loinc = '5902-2', lab_name = 'Prothrombin time', ylim_min = 0, ylim_max = 50)
# liver_lab_total_bulirubin <- new('LiverLab', lab_loinc = '1975-2', lab_name = 'Total bilirubin', ylim_min = 0, ylim_max = 15)
# liver_lab_direct_bulirubin <- new('LiverLab', lab_loinc = '1968-7', lab_name = 'Direct bilirubin', ylim_min = 0, ylim_max = 15)
# 
# liver_labs <- c(liver_lab_ALT, 
#                 liver_lab_GGT, 
#                 liver_lab_AST, 
#                 liver_lab_ALP, 
#                 liver_lab_total_protein, 
#                 liver_lab_serum_albumin, 
#                 liver_lab_total_bulirubin, 
#                 liver_lab_direct_bulirubin)
# 
# plots <- map(liver_labs, function(x) plot_lab_raw_values_vs_rel_date(lab_loinc = x@lab_loinc, ylim = c(x@ylim_min, x@ylim_max)))
# p <- arrangeGrob(sapply(plots, function(x) x[[1]]), ncol = 2)

p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2160-0', ylim = c(0, 20)) + ggtitle('Serum creatinine')

p2 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '3094-0', ylim = c(0, 150)) + ggtitle('Urea nitrogen in serum')

p3 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '6299-2', ylim = c(0, 200)) + ggtitle('Urea nitrogen in blood')

p4 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2823-3', ylim = c(0, 10)) + ggtitle('Potassium in serum')

# seems the test strip only reports positive or negative outcomes
p5 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '5804-0', ylim = c(0, 500)) + ggtitle('protein in urine')

# very sparse data
p6 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2889-4', ylim = c(0, 200)) + ggtitle('Protein in 24h urine')

# very very sparse data
p7 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2888-6', ylim = c(0, 500)) + ggtitle('Protein in urine')

p <- gridExtra::arrangeGrob(p1, p2, p3, p4, p5, p6, p7, ncol = 3)

ggsave(filename = 'cache/kidney_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
```

![liver function labs](./cache/kidney_labs_raw_data_distribution_by_rel_date.png)

Use -180 as the time cutoff and separate the patient journey into two sections: 0 - 180 days before the first diagnosis of abnormal liver function, and 180 days before the first diagnosis of liver abnormal functions. Within each section, we will pick the latest lab measurement for each LOINC lab test (features) and assign either + (within 180 days) or - outcome (outside of 180 days). 
```{r}
# prepare data entries for + and - outcomes
# each lab result is associated with an outcome
# briefly, the following snippet uses the following algorithm:
# calculate the relative date to the first abnormal liver function diagnosis: if before diagnosis, - days otherwise + days
# then cut the relative days into a few windows (-Inf, -365), (-365, -30), (-30, -7), (-7, 0) and (0, Inf)
# for each window, we only use the closest one to diagnosis date as a representative, although we can use other methods such as the mean
# at last, only the (-7, 0) window has + outcome, meaning having abnormal liver function; all other windows have normal liver function

df_labs_and_outcomes <- df_labs %>% filter(loinc %in% c('2160-0', '3094-0', '6299-2', '2823-3')) %>%
  as_tibble() %>%
  select(-min_diag_date, -rel_date) %>% ## these two columns can be removed from the sql query
  ##filter(medical_record_number %in% c('2715515', '3087585')) %>% ## test with a few patient
  left_join(df_cohort, by = 'medical_record_number') %>%
  mutate(lab_date = ymd(lab_date), min_diag_date = ymd(min_diag_date)) %>%
  mutate(lab_date_rel_to_diag = as.duration(interval(min_diag_date, lab_date)) / ddays()) %>%
  mutate(window_num = cut(lab_date_rel_to_diag, c(-Inf, -180, 0, Inf))) %>%
##select(medical_record_number, loinc, lab_date, min_diag_date, lab_date_rel_to_diag, window_num)
  filter(window_num %in% c('(-Inf,-180]', '(-180,0]')) %>%
  arrange(medical_record_number, loinc, -lab_date_rel_to_diag) %>%
  group_by(medical_record_number, loinc, window_num) %>%
  mutate(sequence = row_number()) %>%
  #select(medical_record_number, loinc, lab_date, min_diag_date, lab_date_rel_to_diag, window_num, sequence) %>%
  ungroup() %>%
  filter(sequence == 1) %>%
  mutate(outcome = if_else(window_num == '(-180,0]', 'abnormal', 'normal'))

# the following shows that we have pretty balanced classes for each lab test
df_labs_and_outcomes %>% group_by(loinc, outcome) %>% summarise(n = n())
```

Define a list of functions to run statistical comparisons.

```{r}
require(caret)
require(broom)
require(pROC)

# preprocess: convert the HPO column (and the isnegated column) into three separate features:
# is_hpoterm_observed_for_H, is_hpoterm_observed_for_L and is_hpoterm_observed_for_N
# pass in the dataframe of lab tests of one LOINC, return a new dataframe with the three new features
create_hpo_features <- function(df_one_loinc){
  df_hpo_features <- df_one_loinc %>%
    mutate(dummy_value = 1) %>%
    pivot_wider(names_from = interpretation, names_prefix = 'is_hpoterm_observed_for_', values_from = dummy_value, values_fill = 0) %>%
    select(lab_result_numeric_value, starts_with('is_hpoterm_observed_for_'), outcome)

  df_hpo_features <- df_hpo_features %>%
    mutate(outcome = as.factor(outcome)) %>%
    mutate(outcome = relevel(outcome, 'normal')) %>%
    mutate(lab_result_numeric_value = as.numeric(lab_result_numeric_value))

  return (df_hpo_features)
}

train_test_split <- function(df, p, ...){
  trainIndex = createDataPartition(df$outcome, p = p, ...)
  df_train <- df[trainIndex, ]
  df_test <- df[-trainIndex, ]
  return (list(df_train=df_train, df_test=df_test))
}

lr_for_hpo_H <- function(df_train, df_test){
  model_continuous <- glm(formula = 'outcome ~ lab_result_numeric_value', data = df_train, family = binomial(link = 'logit'))
  model_hpo <- glm(formula = 'outcome ~ is_hpoterm_observed', data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_H), family = binomial(link = 'logit'))
  prediction_continuous <- predict(model_continuous, df_test, type = 'response')
  prediction_hpo <- predict(model_hpo, df_test %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_H), type = 'response')
  roc_continuous <- roc(df_test$outcome, prediction_continuous)
  roc_hpo <- roc(df_test$outcome, prediction_hpo)
  return (list(model_continuous, model_hpo, prediction_continuous, prediction_hpo, roc_continuous, roc_hpo))
}

lr_for_hpo_L <- function(df_train, df_test){
  model_continuous <- glm(formula = 'outcome ~ lab_result_numeric_value', data = df_train, family = binomial(link = 'logit'))
  model_hpo <- glm(formula = 'outcome ~ is_hpoterm_observed', data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_L), family = binomial(link = 'logit'))
  prediction_continuous <- predict(model_continuous, df_test, type = 'response')
  prediction_hpo <- predict(model_hpo, df_test %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_L), type = 'response')
  roc_continuous <- roc(df_test$outcome, prediction_continuous)
  roc_hpo <- roc(df_test$outcome, prediction_hpo)
  return (list(model_continuous, model_hpo, prediction_continuous, prediction_hpo, roc_continuous, roc_hpo))
}

compare_rocs <- function(roc1, roc2){
  results <- broom::tidy(roc.test(roc1, roc2, method = "bootstrap"))
  return (results)
}

pipeline_compare_for_outcome_prediction <- function(df_one_loinc) {

  df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc)
  df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE)

  df_train <- df_train_test$df_train
  df_test <- df_train_test$df_test

  # check if we have H for analysis, and/or L
  valid_lab_outcomes <- df_one_loinc %>% filter(!is.na(hpotermid) & !isnegated) %>% select(interpretation, hpotermid) %>% distinct()

  roc_comparisons <- list()
  if ('H' %in% valid_lab_outcomes$interpretation){
    lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test)
    roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
    roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'H'))
    roc_comparisons$H <- roc_comparison
  }

  if ('L' %in% valid_lab_outcomes$interpretation){
    lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
    roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
    roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'L'))
    roc_comparisons$L <- roc_comparison
  }

  roc_comparisons_combined <- bind_rows(roc_comparisons)

  roc_comparisons_combined
}

example_loinc = '2160-0'
df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == example_loinc)
pipeline_compare_for_outcome_prediction(df_one_loinc = df_one_loinc)
```

```{r}
require(furrr)
plan(multisession, workers = 4)

kidney_analysis <- df_labs_and_outcomes %>%
  group_by(loinc) %>%
  nest() %>%
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
  select(-data) %>%
  unnest(result)

kidney_analysis
```

```{r}
df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '1742-6')

df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc)

df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE)

df_train <- df_train_test$df_train
df_test <- df_train_test$df_test

lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test)
roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'H'))
roc_comparisons$H <- roc_comparison

plot(lr_H[[5]])
plot(lr_H[[6]])

roc.test(lr_H[[5]], lr_H[[6]], method = "bootstrap")
```

### 2021-12-01 src/main/r/numeric_vs_hpo_for_liver_function.Rmd
compare raw and hpo of kidney labs for acute kidney failure

--- 2021-12-01 src/main/r/numeric_vs_hpo_for_acute_kidney_failure.Rmd
@@
 ## Goal
-Compare the performance of raw lab values vs HPO interpretations for acute kidney failure.
-
-The analysis process is very similar to the liver function analysis.
+Compare the performance of raw lab values vs HPO interpretations for liver function prediction.
 
 Algorithm:
-- Define a cohort of patients coded with acute kidney failure as the primay diagnosis of a visit;
+- Define a cohort of patients coded with abnormal liver function, and the earliest date of diagnosis;
 - Get all the lab tests with selected LOINC code; 
@@
 ```{r}
-sql <- read_file('../sql/acute_kidney_failure_cohort.sql')
+sql <- read_file('../sql/liver_abnormal_function_cohort.sql')
 print(sql)
@@
 sql <- "SELECT * FROM cohort"
-df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'acute_kidney_failure_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'liver_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
 
@@
 ```{r}
-sql <- read_file('../sql/get_labs_for_acute_kidney_failure.sql')
-df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_acute_kidney_failure.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
-df_labs %>% nrow()
+sql <- read_file('../sql/get_labs_for_liver_function.sql')
+df_liver_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_liver_function.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
+df_liver_labs %>% nrow()
 ```
@@
 ```{r}
-plot_lab_raw_values_vs_rel_date <- function(lab_loinc, ylim=NULL){
-  df <- df_labs %>% 
+plot_liver_lab_raw_values_vs_rel_date <- function(lab_loinc, ylim=NULL){
+  df <- df_liver_labs %>% 
     filter(loinc == lab_loinc) %>% 
@@
 # 
-# plots <- map(liver_labs, function(x) plot_lab_raw_values_vs_rel_date(lab_loinc = x@lab_loinc, ylim = c(x@ylim_min, x@ylim_max)))
+# plots <- map(liver_labs, function(x) plot_liver_lab_raw_values_vs_rel_date(lab_loinc = x@lab_loinc, ylim = c(x@ylim_min, x@ylim_max)))
 # p <- arrangeGrob(sapply(plots, function(x) x[[1]]), ncol = 2)
 
-p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2160-0', ylim = c(0, 20)) + ggtitle('Serum creatinine')
+## LOINC 1742-6 Alanine aminotransferase
+p1 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1742-6', ylim = c(0, 500)) + ggtitle('Alanine aminotransferase')
 
-p2 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '3094-0', ylim = c(0, 150)) + ggtitle('Urea nitrogen in serum')
+## LOINC 2324-2 Gamma glutamyl transferase
+p2 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '2324-2', ylim = c(0, 1000)) + ggtitle('Gamma glutamyl transferase')
 
-p3 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '6299-2', ylim = c(0, 200)) + ggtitle('Urea nitrogen in blood')
+## LOINC 1920-8 Aspartate aminotransferase
+p3 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1920-8', ylim = c(0, 500)) + ggtitle('Aspartate aminotransferase')
 
-p4 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2823-3', ylim = c(0, 10)) + ggtitle('Potassium in serum')
+## LOINC 6768-6 Alkaline phosphatase
+p4 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '6768-6', ylim = c(0, 1000)) + ggtitle('Alkaline phosphatase')
 
-# seems the test strip only reports positive or negative outcomes
-p5 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '5804-0', ylim = c(0, 500)) + ggtitle('protein in urine')
+## LOINC 2885-2 total protein: if low, maybe reflecting liver or kidney problem
+p5 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '2885-2', ylim = c(0, 15)) + ggtitle('Total protein')
 
-# very sparse data
-p6 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2889-4', ylim = c(0, 200)) + ggtitle('Protein in 24h urine')
+## LOINC 1751-7 serum albumin
+p6 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1751-7', ylim = c(0, 7.5)) + ggtitle('Serum albumin')
 
-# very very sparse data
-p7 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2888-6', ylim = c(0, 500)) + ggtitle('Protein in urine')
+## LOINC 5902-2 Prothrombin time
+p7 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '5902-2', ylim = c(0, 50)) + ggtitle('PT')
 
-p <- gridExtra::arrangeGrob(p1, p2, p3, p4, p5, p6, p7, ncol = 3)
+## LOINC 1975-2 Bilirubin total
+p8 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1975-2', ylim = c(0, 15)) + ggtitle('Bilirubin total')
 
-ggsave(filename = 'cache/kidney_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
+## LOINC 1968-7 Bilirubin direct
+p9 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1968-7', ylim = c(0, 15)) + ggtitle('Bilirubin direct')
+
+p <- gridExtra::arrangeGrob(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3)
+
+ggsave(filename = 'cache/liver_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
 ```
 
-![liver function labs](./cache/kidney_labs_raw_data_distribution_by_rel_date.png)
+![liver function labs](./cache/liver_labs_raw_data_distribution_by_rel_date.png)
+
+Machine learning part:
+  For each LOINC lab test
+    prepare dataframe for each 
+    data split
+    train logistic regression with either raw numeric value or HPO terms
+    build ROC with test set
+    run statistical comparison to compare the two models
 
@@
 
-df_labs_and_outcomes <- df_labs %>% filter(loinc %in% c('2160-0', '3094-0', '6299-2', '2823-3')) %>%
+df_liver_labs_and_outcomes <- df_liver_labs %>% 
   as_tibble() %>%
@@
 # the following shows that we have pretty balanced classes for each lab test
-df_labs_and_outcomes %>% group_by(loinc, outcome) %>% summarise(n = n())
+df_liver_labs_and_outcomes %>% group_by(loinc, outcome) %>% summarise(n = n())
 ```
@@
 
-pipeline_compare_for_outcome_prediction <- function(df_one_loinc) {
+pipeline_compare_for_liver_function <- function(df_one_loinc) {
 
@@
 
-example_loinc = '2160-0'
-df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == example_loinc)
-pipeline_compare_for_outcome_prediction(df_one_loinc = df_one_loinc)
+example_loinc = '1742-6'
+df_one_loinc <- df_liver_labs_and_outcomes %>% filter(loinc == example_loinc)
+pipeline_compare_for_liver_function(df_one_loinc = df_one_loinc)
 ```
@@
 
-kidney_analysis <- df_labs_and_outcomes %>%
+liver_analysis <- df_liver_labs_and_outcomes %>%
   group_by(loinc) %>%
   nest() %>%
-  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
+  mutate(result = future_map(data, pipeline_compare_for_liver_function)) %>%
   select(-data) %>%
@@
 
-kidney_analysis
+liver_analysis
 ```
@@
 ```{r}
-df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '1742-6')
+df_one_loinc <- df_liver_labs_and_outcomes %>% filter(loinc == '1742-6')
 

### 2021-12-01 src/main/r/functions.R

@@ -1,0 +1,24 @@
+require(data.table)
+
+#' Get dataset from database; if data is already cached, then directly retrieve from cache
+#' @param cache_dir, cache directory
+#' @param file_name, filename used for reading from or writing to cache directory
+#' @param db_access_callback, a callback to retrieve data from database
+#' @param conn, necessary for db_access_call_back
+#' @param statement, sql query, necessary for db_access_call_back
+#' @param force_rerun, if true, get data from database regardless of whether file exists in cache or not, and then write to cache directory  
+get_df_from_db_or_cache <- function(cache_dir, file_name, 
+                                    db_access_callback = NULL,
+                                    conn = NULL, 
+                                    statement = NULL,
+                                    force_rerun = FALSE, 
+                                    ...){
+  df_path <- file.path(cache_dir, file_name)
+  if (force_rerun | !file.exists(df_path)){
+    df <- db_access_callback(conn, statement)
+    write.csv(df, file = df_path, row.names = FALSE)
+  }
+  
+  # data.table::fread is typically preferred
+  df <- data.table::fread(df_path, ...)
+}

### 2021-12-01 src/main/r/numeric_vs_hpo_for_acute_kidney_failure.Rmd
minor tweak

@@
 title: "R Notebook"
-output: html_notebook
+output:
+  html_document:
+    df_print: paged
 ---
@@
   mutate(lab_date_rel_to_diag = as.duration(interval(min_diag_date, lab_date)) / ddays()) %>%
-  mutate(window_num = cut(lab_date_rel_to_diag, c(-Inf, -180, 0, Inf))) %>%
+  mutate(window_num = cut(lab_date_rel_to_diag, c(-Inf, -100, 0, Inf))) %>%
 ##select(medical_record_number, loinc, lab_date, min_diag_date, lab_date_rel_to_diag, window_num)
-  filter(window_num %in% c('(-Inf,-180]', '(-180,0]')) %>%
+  filter(window_num %in% c('(-Inf,-100]', '(-100,0]')) %>%
   arrange(medical_record_number, loinc, -lab_date_rel_to_diag) %>%
@@
   filter(sequence == 1) %>%
-  mutate(outcome = if_else(window_num == '(-180,0]', 'abnormal', 'normal'))
+  mutate(outcome = if_else(window_num == '(-100,0]', 'abnormal', 'normal'))
 
@@
 
-```{r}
+```{r max.output=-1}
 require(furrr)
@@
 ```{r}
-df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '1742-6')
+df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '2160-0')
 
@@
 roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
-roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'H'))
-roc_comparisons$H <- roc_comparison
+roc_comparison <- bind_cols(roc_comparison)
+roc_comparison$H <- roc_comparison
 

### 2021-12-10 src/main/r/numeric_vs_hpo_for_CRC.Rmd
add code for aplastic anemia and CRC

--- 2021-12-01 src/main/r/numeric_vs_hpo_for_acute_kidney_failure.Rmd
@@
 ## Goal
-Compare the performance of raw lab values vs HPO interpretations for acute kidney failure.
-
-The analysis process is very similar to the liver function analysis.
-
-Algorithm:
-- Define a cohort of patients coded with acute kidney failure as the primay diagnosis of a visit;
-- Get all the lab tests with selected LOINC code; 
-- For each patient, the labs within X days (e.g.
-7 days) before the diagnosis date is considered to be associated with + outcome; and the labs further outside of the window is considered to be associated with - outcome;
-- For each LOINC code, we will prepare a dataframe, separate into training and test set, train a logistic regression classifier using either raw numerical value or HPO value (observed or not observed), build the ROC curve with the test set, and finally compare the ROC curve to determine whether they are statistically different.
+Compare the performance of raw lab values vs HPO interpretations for colorectal cancer.
 
@@
 
-Create a cohort for patients with abnormal liver function diagnoses.
+Create a cohort for patients with colorectal cancer.
 ```{r}
-sql <- read_file('../sql/acute_kidney_failure_cohort.sql')
+sql <- read_file('../sql/colorectal_cancer_cohort.sql')
 print(sql)
@@
 
-Retrieve liver function diagnoses information for the cohort.
+Retrieve diagnoses information for the cohort.
 ```{r}
 sql <- "SELECT * FROM cohort"
-df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'acute_kidney_failure_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'colorectal_cancer_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
 
@@
 
-Retrieve liver function lab tests for the cohort. 
+Retrieve related lab tests for the cohort. 
 ```{r}
-sql <- read_file('../sql/get_labs_for_acute_kidney_failure.sql')
-df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_acute_kidney_failure.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
+sql <- read_file('../sql/get_labs_for_colorectal_cancer.sql')
+df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_colorectal_cancer.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
 df_labs %>% nrow()
@@
 
-p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2160-0', ylim = c(0, 20)) + ggtitle('Serum creatinine')
+p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '6690-2', ylim = c(0, 50)) + ggtitle('Leukocytes')
 
-p2 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '3094-0', ylim = c(0, 150)) + ggtitle('Urea nitrogen in serum')
+p2 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '789-8', ylim = c(0, 10)) + ggtitle('RBC')
 
-p3 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '6299-2', ylim = c(0, 200)) + ggtitle('Urea nitrogen in blood')
+p3 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '718-7', ylim = c(0, NA)) + ggtitle('Hemoglobin')
 
-p4 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2823-3', ylim = c(0, 10)) + ggtitle('Potassium in serum')
+p4 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '4544-3', ylim = c(0, NA)) + ggtitle('Hemotocrit')
 
-# seems the test strip only reports positive or negative outcomes
-p5 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '5804-0', ylim = c(0, 500)) + ggtitle('protein in urine')
+p5 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '787-2', ylim = c(40, 140)) + ggtitle('MCV')
 
-# very sparse data
-p6 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2889-4', ylim = c(0, 200)) + ggtitle('Protein in 24h urine')
+p6 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '785-6', ylim = c(0, NA)) + ggtitle('MCH')
 
-# very very sparse data
-p7 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2888-6', ylim = c(0, 500)) + ggtitle('Protein in urine')
+p7 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '786-4', ylim = c(20, 40)) + ggtitle('MCHC')
 
-p <- gridExtra::arrangeGrob(p1, p2, p3, p4, p5, p6, p7, ncol = 3)
+p8 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '21000-5', ylim = c(0, NA)) + ggtitle('EDW')
 
-ggsave(filename = 'cache/kidney_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
+p9 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '788-0', ylim = c(0, NA)) + ggtitle('EDW')
+
+p10 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '777-3', ylim = c(0, 1000)) + ggtitle('Platelets')
+
+p11 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '32207-3', ylim = c(0, NA)) + ggtitle('Platelet dist width')
+
+p12 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '32623-1', ylim = c(0, 20)) + ggtitle('Platelet mean volume')
+
+p <- gridExtra::arrangeGrob(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol = 4)
+
+ggsave(filename = 'cache/CRC_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 8, height = 6)
 ```
 
-![liver function labs](./cache/kidney_labs_raw_data_distribution_by_rel_date.png)
+![liver function labs](./cache/CRC_labs_raw_data_distribution_by_rel_date.png)
 
@@
 
-df_labs_and_outcomes <- df_labs %>% filter(loinc %in% c('2160-0', '3094-0', '6299-2', '2823-3')) %>%
+df_labs_and_outcomes <- df_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
   as_tibble() %>%
@@
   mutate(lab_date_rel_to_diag = as.duration(interval(min_diag_date, lab_date)) / ddays()) %>%
-  mutate(window_num = cut(lab_date_rel_to_diag, c(-Inf, -100, 0, Inf))) %>%
+  mutate(window_num = cut(lab_date_rel_to_diag, c(-Inf, -180, 0, Inf))) %>%
 ##select(medical_record_number, loinc, lab_date, min_diag_date, lab_date_rel_to_diag, window_num)
-  filter(window_num %in% c('(-Inf,-100]', '(-100,0]')) %>%
+  filter(window_num %in% c('(-Inf,-180]', '(-180,0]')) %>%
   arrange(medical_record_number, loinc, -lab_date_rel_to_diag) %>%
@@
   filter(sequence == 1) %>%
-  mutate(outcome = if_else(window_num == '(-100,0]', 'abnormal', 'normal'))
+  mutate(outcome = if_else(window_num == '(-180,0]', 'abnormal', 'normal'))
 
@@
 
-example_loinc = '2160-0'
+example_loinc = '787-2'
 df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == example_loinc)
@@
 
-```{r max.output=-1}
+```{r max.output=-1, message=FALSE}
 require(furrr)
@@
 
-kidney_analysis <- df_labs_and_outcomes %>%
+crc_analysis <- df_labs_and_outcomes %>%
   group_by(loinc) %>%
@@
 
-kidney_analysis
+crc_analysis
 ```
 
-```{r}
-df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '2160-0')
+```{r message=FALSE}
+df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '4544-3')
 
@@
 
-lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test)
-roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
+lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
+roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
 roc_comparison <- bind_cols(roc_comparison)
-roc_comparison$H <- roc_comparison
 
-plot(lr_H[[5]])
-plot(lr_H[[6]])
+plot(lr_L[[5]])
+plot(lr_L[[6]])
 
-roc.test(lr_H[[5]], lr_H[[6]], method = "bootstrap")
+roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap")
 ```
+
+## Use separate cohort for control
+
+The above section used the same CRC cohort as their controls (patients were assumed CRC free one year before their diagnoses).
+In this section, we will define a separate cohort as controls.
+The control cohort are entirely free from any cancer diagnosis.
+We will start from a random sample of cancer-free patients.
+
+Retrieve the demographic table for cases, and then for controls. 
+Retrieve the lab table for cases, and then for controls. 
+Select the labs within 180 days of diagnosis for analysis;
+Select the latest labs for the control patients for analysis.
+
+```{r}
+sql <- read_file('../sql/colorectal_cancer_cohort_demographics.sql')
+df_case_demographics <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'demographics_colorectal_cancer_cases.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+
+sql <- read_file('../sql/colorectal_cancer_control_cohort.sql')
+n<- dbExecute(dsca, sql)
+
+sql <- 'SELECT * FROM control_cohort'
+df_control_demographics <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'demographics_colorectal_cancer_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+
+sql <- read_file('../sql/get_labs_for_colorectal_cancer_control.sql')
+df_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_sampled_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = TRUE)
+```
+
+```{r}
+# labs for cases
+df_labs_and_outcomes_cases <- df_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
+  as_tibble() %>%
+  select(-min_diag_date, -rel_date) %>% ## these two columns can be removed from the sql query
+  #filter(medical_record_number %in% c('1006680', '3087585')) %>% ## test with a few patient
+  left_join(df_cohort, by = 'medical_record_number') %>%
+  mutate(lab_date = ymd(lab_date), min_diag_date = ymd(min_diag_date)) %>%
+  mutate(lab_date_rel_to_diag = as.duration(interval(min_diag_date, lab_date)) / ddays()) %>%
+  filter(lab_date_rel_to_diag >= -180 & lab_date_rel_to_diag <= 0) %>%
+  arrange(medical_record_number, loinc, -lab_date_rel_to_diag) %>%
+  group_by(medical_record_number, loinc) %>%
+  mutate(sequence = row_number()) %>%
+  ungroup() %>%
+  filter(sequence == 1) %>%
+  mutate(outcome = 'abnormal')
+
+# labs for controls
+df_labs_and_outcomes_controls <- df_control_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
+  as_tibble() %>%
+  #filter(medical_record_number == '7007952')  %>%
+  #mutate(lab_date = as.Date(lab_date)) %>%
+  arrange(medical_record_number, loinc, desc(lab_date)) %>%
+  group_by(medical_record_number, loinc) %>%
+  mutate(sequence = row_number()) %>%
+  ungroup() %>%
+  filter(sequence == 1) %>%
+  mutate(outcome = 'normal')
+
+nrow(df_labs_and_outcomes_cases)
+nrow(df_labs_and_outcomes_controls)
+
+df_labs_and_outcomes_case_control <- bind_rows(
+  df_labs_and_outcomes_cases %>% select(medical_record_number, lab_date, loinc, local_test_code, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated, outcome),
+  df_labs_and_outcomes_controls %>% select(medical_record_number, lab_date, loinc, local_test_code, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated, outcome)
+)
+
+rm(df_labs_and_outcomes_cases, df_labs_and_outcomes_controls)
+
+df_labs_and_outcomes_case_control %>% sample_n(10)
+```
+```{r max.output=-1, message=FALSE}
+require(furrr)
+plan(multisession, workers = 4)
+
+crc_separate_controls_analysis <- df_labs_and_outcomes_case_control %>%
+  group_by(loinc) %>%
+  nest() %>%
+  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
+  select(-data) %>%
+  unnest(result)
+
+crc_separate_controls_analysis
+```
+
+```{r message=FALSE}
+df_one_loinc <- df_labs_and_outcomes_case_control %>% filter(loinc == '4544-3')
+
+df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc)
+
+df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE)
+
+df_train <- df_train_test$df_train
+df_test <- df_train_test$df_test
+
+lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
+roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
+roc_comparison <- bind_cols(roc_comparison)
+
+plot(lr_L[[5]])
+plot(lr_L[[6]])
+
+roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap")
+```

### 2021-12-10 src/main/r/numeric_vs_hpo_for_blood_disorder.Rmd
add code for aplastic anemia and CRC

--- 2021-12-01 src/main/r/numeric_vs_hpo_for_acute_kidney_failure.Rmd
@@
 ## Goal
-Compare the performance of raw lab values vs HPO interpretations for acute kidney failure.
-
-The analysis process is very similar to the liver function analysis.
-
-Algorithm:
-- Define a cohort of patients coded with acute kidney failure as the primay diagnosis of a visit;
-- Get all the lab tests with selected LOINC code; 
-- For each patient, the labs within X days (e.g.
-7 days) before the diagnosis date is considered to be associated with + outcome; and the labs further outside of the window is considered to be associated with - outcome;
-- For each LOINC code, we will prepare a dataframe, separate into training and test set, train a logistic regression classifier using either raw numerical value or HPO value (observed or not observed), build the ROC curve with the test set, and finally compare the ROC curve to determine whether they are statistically different.
+Compare the performance of raw lab values vs HPO interpretations for seme blood disorders
 
@@
 
-Create a cohort for patients with abnormal liver function diagnoses.
+Create a cohort for patients with aplastic anemia.
 ```{r}
-sql <- read_file('../sql/acute_kidney_failure_cohort.sql')
+sql <- read_file('../sql/aplastic_anemia.sql')
 print(sql)
@@
 
-Retrieve liver function diagnoses information for the cohort.
+Retrieve diagnoses information for the cohort.
 ```{r}
 sql <- "SELECT * FROM cohort"
-df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'acute_kidney_failure_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'aplastic_anemia_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
 
@@
 
-Retrieve liver function lab tests for the cohort. 
+Retrieve related lab tests for the cohort. 
 ```{r}
-sql <- read_file('../sql/get_labs_for_acute_kidney_failure.sql')
-df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_acute_kidney_failure.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
+sql <- read_file('../sql/get_labs_for_aplastic_anemia.sql')
+df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_aplastic_anemia.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
 df_labs %>% nrow()
@@
 
-p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2160-0', ylim = c(0, 20)) + ggtitle('Serum creatinine')
+p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '789-8', ylim = c(0, 10)) + ggtitle('RBC')
 
-p2 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '3094-0', ylim = c(0, 150)) + ggtitle('Urea nitrogen in serum')
+p2 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '6690-2', ylim = c(0, 25)) + ggtitle('CBC')
 
-p3 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '6299-2', ylim = c(0, 200)) + ggtitle('Urea nitrogen in blood')
+p3 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '777-3', ylim = c(0, 1000)) + ggtitle('Platelet')
 
-p4 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2823-3', ylim = c(0, 10)) + ggtitle('Potassium in serum')
+p4 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '751-8', ylim = c(0, 25)) + ggtitle('neutrophil')
 
-# seems the test strip only reports positive or negative outcomes
-p5 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '5804-0', ylim = c(0, 500)) + ggtitle('protein in urine')
+p <- gridExtra::arrangeGrob(p1, p2, p3, p4, ncol = 2)
 
-# very sparse data
-p6 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2889-4', ylim = c(0, 200)) + ggtitle('Protein in 24h urine')
-
-# very very sparse data
-p7 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2888-6', ylim = c(0, 500)) + ggtitle('Protein in urine')
-
-p <- gridExtra::arrangeGrob(p1, p2, p3, p4, p5, p6, p7, ncol = 3)
-
-ggsave(filename = 'cache/kidney_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
+ggsave(filename = 'cache/anemia_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 5, height = 5)
 ```
 
-![liver function labs](./cache/kidney_labs_raw_data_distribution_by_rel_date.png)
+![liver function labs](./cache/anemia_labs_raw_data_distribution_by_rel_date.png)
 
@@
 
-df_labs_and_outcomes <- df_labs %>% filter(loinc %in% c('2160-0', '3094-0', '6299-2', '2823-3')) %>%
+df_labs_and_outcomes <- df_labs %>%
   as_tibble() %>%
@@
 
-example_loinc = '2160-0'
+example_loinc = '789-8'
 df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == example_loinc)
@@
 
-kidney_analysis <- df_labs_and_outcomes %>%
+anemia_analysis <- df_labs_and_outcomes %>%
   group_by(loinc) %>%
@@
 
-kidney_analysis
+anemia_analysis
 ```
 
-```{r}
-df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '2160-0')
+```{r message=FALSE}
+df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '789-8')
 
@@
 
-lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test)
-roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
+lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
+roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
 roc_comparison <- bind_cols(roc_comparison)
-roc_comparison$H <- roc_comparison
 
-plot(lr_H[[5]])
-plot(lr_H[[6]])
+plot(lr_L[[5]])
+plot(lr_L[[6]])
 
-roc.test(lr_H[[5]], lr_H[[6]], method = "bootstrap")
+roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap")
 ```

### 2021-12-12 src/main/r/numeric_vs_hpo_for_CRC.Rmd
rerun after propensity score matching

@@
 sql <- read_file('../sql/get_labs_for_colorectal_cancer_control.sql')
-df_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_sampled_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = TRUE)
+df_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_sampled_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
 ```
@@
 ```
+
 ```{r max.output=-1, message=FALSE}
 require(furrr)
-plan(multisession, workers = 4)
+plan(multisession, workers=8)
+start_time <- Sys.time()
 
@@
   nest() %>%
-  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
+  ungroup() %>%
+  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, .options = furrr_options(seed = T))) %>%
   select(-data) %>%
   unnest(result)
+
+end_time <- Sys.time()
+
+print(end_time - start_time)
 
@@
 ```
+
+# Age and race matching
+The above analysis just sampled the control cohort randomly. In this section, we will determine whether we need to match sex, age and race.
+
+```{r}
+## because of our processing, cases without desired labs are not counted here
+df_labs_and_outcomes_case_control %>% select(medical_record_number, outcome) %>%
+  distinct() %>%
+  left_join(bind_rows(df_case_demographics, df_control_demographics), by = 'medical_record_number') %>%
+  group_by(outcome, race) %>%
+  summarise(n = n())
+
+df_demographics_with_outcomes <- bind_rows(df_case_demographics %>% select(-min_diag_date) %>% mutate(outcome = 'abnormal'),
+          df_control_demographics %>% mutate(outcome = 'normal')) %>% mutate(age = as.duration(interval(date_of_birth, today())) / dyears())
+
+df_demographics_with_outcomes <- df_demographics_with_outcomes %>%
+  filter(gender %in% c('Male', 'Female')) %>%
+  filter_all(all_vars(!is.na(.))) %>% 
+  mutate(outcome = as.factor(outcome), 
+         gender = as.factor(gender),
+         race = as.factor(race)) %>% 
+  mutate(outcome = relevel(outcome, 'normal'),
+         gender = relevel(gender, 'Male'),
+         race = relevel(race, 'White'))
+
+## determine if gender and race are associated with outcome: CRC or no CRC
+ncase <- nrow(df_demographics_with_outcomes[df_demographics_with_outcomes$outcome == 'abnormal',])
+model <- glm(formula = outcome ~ gender + race + age, data = df_demographics_with_outcomes %>% filter(outcome == 'abnormal') %>% union_all(df_demographics_with_outcomes %>% filter(outcome == 'normal') %>% sample_n(ncase)),
+                    family = binomial)
+
+anova(model, test = 'Chisq')
+```
+```{r}
+prediction <- predict(model, newdata = df_demographics_with_outcomes , type = 'response', na.action = na.pass)
+
+df_demographics_with_outcomes$propensity_score <- prediction
+
+ggplot(df_demographics_with_outcomes, aes(x = outcome, y = propensity_score)) +
+  geom_violin()
+```
+```{r}
+require(MatchIt)
+
+# it is really slow. have to kill it before finishing
+# match_outcome <- matchit(outcome ~ gender + race + age, method = "nearest", data = df_demographics_with_outcomes %>% select(-propensity_score))
+
+# try this stackoverflow method
+# https://stackoverflow.com/questions/29672088/speedup-matchit
+# still slow, but works
+df_demographics_with_outcomes %>% colnames()
+m.out <- matchit(outcome ~ medical_record_number, data = df_demographics_with_outcomes, method = "nearest", distance = df_demographics_with_outcomes$propensity_score)
+save(m.out, file='m.out.obj')
+matched_unique_ids_etc <- match.data(m.out, data = df_demographics_with_outcomes)
+
+matched_unique_ids_etc %>% group_by(outcome) %>% summarise(n =n())
+ggplot(matched_unique_ids_etc, aes(x = outcome, y = propensity_score)) +
+  geom_violin()
+
+ggplot(matched_unique_ids_etc, aes(x = outcome, y = age)) +
+  geom_violin()
+```
+We need to put back the identifiers of controls to the database in order to selectively retrieve their labs. 
+```{r}
+dbExecute(dsca, "create table if not exists matched_control_cohort (medical_record_number VARCHAR)")
+dbExecute(dsca, "truncate table matched_control_cohort")
+
+insert_mrns <- function(mrns_str, dbconn = dsca){
+  sql <- glue('insert into matched_control_cohort values {mrns_str};')
+  message(sql)
+  dbExecute(dbconn, sql)
+}
+
+matched_unique_ids_etc %>% filter(outcome == 'normal') %>% 
+  rowwise() %>%
+  mutate(batch_number = sample(seq(10), size = 1, replace = TRUE)) %>%
+  ungroup() %>%
+  group_by(batch_number) %>%
+  summarise(sql_query = str_c(str_c("('", medical_record_number, "')"), collapse = ",")) %>% 
+  ungroup() %>%
+  mutate(sql_query = walk(sql_query, insert_mrns, dbconn = dsca))
+
+# retrieve labs for the matched control cohort
+sql <- read_file('../sql/get_labs_for_colorectal_cancer_matched_control.sql')
+df_matched_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_matched_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = TRUE)
+
+# drop the matched control cohort
+# persist it for a while for now
+# n <- dbExecute(dsca, "DROP TABLE IF EXISTS matched_control_cohort")
+# assertthat::are_equal(n, 0)
+```
+
+Rerun the analysis with the matched controls.
+
+```{r}
+# labs for cases
+df_labs_and_outcomes_cases <- df_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
+  as_tibble() %>%
+  select(-min_diag_date, -rel_date) %>% ## these two columns can be removed from the sql query
+  #filter(medical_record_number %in% c('1006680', '3087585')) %>% ## test with a few patient
+  left_join(df_cohort, by = 'medical_record_number') %>%
+  mutate(lab_date = ymd(lab_date), min_diag_date = ymd(min_diag_date)) %>%
+  mutate(lab_date_rel_to_diag = as.duration(interval(min_diag_date, lab_date)) / ddays()) %>%
+  filter(lab_date_rel_to_diag >= -180 & lab_date_rel_to_diag <= 0) %>%
+  arrange(medical_record_number, loinc, -lab_date_rel_to_diag) %>%
+  group_by(medical_record_number, loinc) %>%
+  mutate(sequence = row_number()) %>%
+  ungroup() %>%
+  filter(sequence == 1) %>%
+  mutate(outcome = 'abnormal')
+
+# labs for controls
+df_labs_and_outcomes_matched_controls <- df_matched_control_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
+  as_tibble() %>%
+  #filter(medical_record_number == '7007952')  %>%
+  #mutate(lab_date = as.Date(lab_date)) %>%
+  arrange(medical_record_number, loinc, desc(lab_date)) %>%
+  group_by(medical_record_number, loinc) %>%
+  mutate(sequence = row_number()) %>%
+  ungroup() %>%
+  filter(sequence == 1) %>%
+  mutate(outcome = 'normal')
+
+nrow(df_labs_and_outcomes_cases)
+nrow(df_labs_and_outcomes_matched_controls)
+
+df_labs_and_outcomes_case_matched_control <- bind_rows(
+  df_labs_and_outcomes_cases %>% select(medical_record_number, lab_date, loinc, local_test_code, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated, outcome),
+  df_labs_and_outcomes_matched_controls %>% select(medical_record_number, lab_date, loinc, local_test_code, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated, outcome)
+)
+
+rm(df_labs_and_outcomes_cases, df_labs_and_outcomes_matched_controls)
+
+df_labs_and_outcomes_case_matched_control %>% sample_n(10)
+```
+
+```{r max.output=-1, message=FALSE}
+require(furrr)
+plan(multisession, workers=4)
+start_time <- Sys.time()
+
+crc_case_matched_controls_analysis <- df_labs_and_outcomes_case_matched_control %>%
+  group_by(loinc) %>%
+  nest() %>%
+  ungroup() %>%
+  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, .options = furrr_options(seed = T))) %>%
+  select(-data) %>%
+  unnest(result)
+
+end_time <- Sys.time()
+
+print(end_time - start_time)
+
+crc_case_matched_controls_analysis
+```

### 2021-12-14 src/main/r/numeric_vs_hpo_for_CRC.Rmd
include demographics in model

@@
     pivot_wider(names_from = interpretation, names_prefix = 'is_hpoterm_observed_for_', values_from = dummy_value, values_fill = 0) %>%
-    select(lab_result_numeric_value, starts_with('is_hpoterm_observed_for_'), outcome)
+    select(lab_result_numeric_value, starts_with('is_hpoterm_observed_for_'), any_of(c('age', 'gender', 'race')), outcome)
 
@@
 
-lr_for_hpo_H <- function(df_train, df_test){
-  model_continuous <- glm(formula = 'outcome ~ lab_result_numeric_value', data = df_train, family = binomial(link = 'logit'))
-  model_hpo <- glm(formula = 'outcome ~ is_hpoterm_observed', data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_H), family = binomial(link = 'logit'))
+lr_for_hpo_H <- function(df_train, df_test, include_demographics=FALSE){
+  if (include_demographics){
+    formula_continuous = 'outcome ~ lab_result_numeric_value + age + gender + race'
+    formula_hpo = 'outcome ~ is_hpoterm_observed + age + gender + race'
+  } else {
+    formula_continuous = 'outcome ~ lab_result_numeric_value'
+    formula_hpo = 'outcome ~ is_hpoterm_observed'
+  }
+  model_continuous <- glm(formula = formula_continuous, data = df_train, family = binomial(link = 'logit'))
+  model_hpo <- glm(formula = formula_hpo, data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_H), family = binomial(link = 'logit'))
   prediction_continuous <- predict(model_continuous, df_test, type = 'response')
@@
 
-lr_for_hpo_L <- function(df_train, df_test){
-  model_continuous <- glm(formula = 'outcome ~ lab_result_numeric_value', data = df_train, family = binomial(link = 'logit'))
-  model_hpo <- glm(formula = 'outcome ~ is_hpoterm_observed', data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_L), family = binomial(link = 'logit'))
+lr_for_hpo_L <- function(df_train, df_test, include_demographics=FALSE){
+  if (include_demographics){
+    formula_continuous = 'outcome ~ lab_result_numeric_value + age + gender + race'
+    formula_hpo = 'outcome ~ is_hpoterm_observed + age + gender + race'
+  } else {
+    formula_continuous = 'outcome ~ lab_result_numeric_value'
+    formula_hpo = 'outcome ~ is_hpoterm_observed'
+  }
+
+  model_continuous <- glm(formula = formula_continuous, data = df_train, family = binomial(link = 'logit'))
+  model_hpo <- glm(formula = formula_hpo, data = df_train %>% rename(is_hpoterm_observed = is_hpoterm_observed_for_L), family = binomial(link = 'logit'))
   prediction_continuous <- predict(model_continuous, df_test, type = 'response')
@@
 
-pipeline_compare_for_outcome_prediction <- function(df_one_loinc) {
+pipeline_compare_for_outcome_prediction <- function(df_one_loinc, include_demographics=FALSE) {
 
@@
   if ('H' %in% valid_lab_outcomes$interpretation){
-    lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test)
+    lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test, include_demographics=include_demographics)
     roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
@@
   if ('L' %in% valid_lab_outcomes$interpretation){
-    lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
+    lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test, include_demographics=include_demographics)
     roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
@@
 
-ggplot(matched_unique_ids_etc, aes(x = outcome, y = age)) +
-  geom_violin()
+plot_match_age <- ggplot(matched_unique_ids_etc, aes(x = outcome, y = age, fill = outcome)) +
+  geom_violin(draw_quantiles = c(0.5)) +
+  theme(legend.position = 'none')
+
+plot_match_gender <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = gender)) +
+  geom_bar(stat = 'count')
+
+plot_match_race <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = race)) +
+  geom_bar(stat = 'count')
+
+plot_match <- gridExtra::arrangeGrob(plot_match_age, plot_match_gender, plot_match_race, layout_matrix = rbind(c(1,2), c(3,3)), ncol = 1)
+plot(plot_match)
 ```
@@
 df_labs_and_outcomes_matched_controls <- df_matched_control_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
+  # my current control still contains CRC patients
+  # TODO: further purify controls
+  filter(!(medical_record_number %in% df_labs_and_outcomes_cases$medical_record_number)) %>%
   as_tibble() %>%
@@
 ```
+
+Combine demographics to see whether the AUC can be improved. 
+```{r}
+df_demographics_case_matched_control <- bind_rows(df_case_demographics %>% select(-min_diag_date), 
+                                                  df_control_demographics %>% filter(medical_record_number %in% 
+                                                                                       unique(df_matched_control_labs$medical_record_number)) %>%
+                                                    filter(!(medical_record_number %in% df_case_demographics$medical_record_number)))
+
+plan(multisession, workers=4)
+start_time <- Sys.time()
+
+crc_case_matched_controls_analysis_with_demographics <- df_labs_and_outcomes_case_matched_control %>%
+  left_join(df_demographics_case_matched_control, by = 'medical_record_number') %>%
+  mutate(age = as.duration(interval(date_of_birth, lab_date)) / dyears()) %>%
+  filter(!(race %in% c('African American', 'American Indian or Alaska Native', 'Native Hawaiian or Other Pacific Islander'))) %>%
+  # mutate(gender = as.factor(gender),
+  #        race = as.factor(race)) %>% 
+  # mutate(gender = relevel(gender, 'Male'),
+  #        race = relevel(race, 'White')) %>%
+  group_by(loinc) %>%
+  nest() %>%
+  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, include_demographics=TRUE, .options = furrr_options(seed = T))) %>%
+  select(-data) %>%
+  unnest(result)
+
+end_time <- Sys.time()
+
+print(end_time - start_time)
+
+crc_case_matched_controls_analysis_with_demographics
+```

### 2022-03-08 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
update

@@
 
+### Collect patient basic demographics
+```{r}
+get_patient_race_sex_distribution <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/patient_count_by_race_sex.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, "
+            select race, gender, count(*) as n
+            from hai_az_prod.person_msdw_2020july pmj 
+            group by race, gender 
+            order by race, gender;
+            ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+
+}
+
+patient_race_sex_distribution <- get_patient_race_sex_distribution()
+
+# recode all unknown gender as one category
+patient_race_sex_distribution <- patient_race_sex_distribution %>% 
+  mutate(gender = recode(gender, 'MSDW_NOT APPL' = 'Unknown')) %>%
+  mutate(gender = recode(gender, 'MSDW_OTHER' = 'Unknown')) %>%
+  mutate(gender = recode(gender, 'MSDW_UNKNOWN' = 'Unknown')) %>%
+  mutate(gender = recode(gender, 'NOT AVAILABLE'= 'Unknown')) %>%
+  mutate(gender = recode(gender, .missing = 'Unknown'))
+
+# recode "African American" as "Black or African American"
+# TODO: update the race map
+patient_race_sex_distribution <- patient_race_sex_distribution %>%
+  mutate(race = recode(race, 'African American' = 'Black or African American'))
+```
+
+```{r}
+countByGender <- patient_race_sex_distribution %>% 
+  group_by(gender) %>% 
+  summarise(n = sum(n))
+
+plot_by_sex <- ggplot(countByGender) + 
+  geom_bar(aes(x = 1, y = n / sum(n), fill = gender), stat = "identity") + 
+  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral", Indeterminant = 'Cyan', Unknown = 'Orange')) + 
+  xlab("") + ylab("patient count") +
+  coord_polar(theta = "y") + 
+  theme_bw() + 
+  theme(panel.grid = element_blank(), legend.position = c(0.25, 0.7), 
+        axis.text = element_blank(), axis.ticks = element_blank(), 
+        axis.title = element_text(size = 8), legend.text = element_text(size = 8), 
+        legend.title = element_text(size = 10), 
+        legend.background = element_rect(fill=alpha('white', 0.1)))
+
+ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.png'), plot = plot_by_sex, width = 2.5, height = 2.5)
+plot_by_sex
+
+```
+
+```{r}
+countByRace_Gender <- patient_race_sex_distribution
+
+countByRace_Gender$race <- if_else(countByRace_Gender$race == 'Unknown', 'Unreported', countByRace_Gender$race)
+
+plot_by_race  <- ggplot(countByRace_Gender %>% 
+         filter(gender %in% c('Male', 'Female')) %>% 
+         mutate(race = as.factor(race), n = if_else(gender == "Male", n, -n))) + 
+  geom_bar(aes(x = reorder(race, -n, FUN=sum), y = n, fill = gender, group=gender), stat="identity") + 
+##  scale_y_continuous(breaks = seq(-40,40, 10), labels = abs(seq(-40,40, 10))) +
+  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral")) + 
+  scale_y_continuous(labels = abs) +
+  xlab("") + ylab("patient count") +
+  coord_flip() + 
+  theme_bw() + 
+  theme(panel.grid = element_blank(), legend.position = "right", 
+        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
+        legend.text = element_text(size = 8), legend.title = element_text(size = 10), 
+        legend.background = element_rect(fill=alpha('white', 0.1)))
+
+ggsave(filename = glue("{CACHE_DIR}/plotByRace.png"), plot = plot_by_race, width = 5, height = 2)
+plot_by_race
+```
+
+Patient age analysis: find the birth years for all the patients and find their current age
+```{r}
+get_patient_birth_year_distribution <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/patient_count_by_birth_year_sex.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, "
+            with person_birth_year as (
+            	select medical_record_number , date_of_birth , gender , date_part('Year', date_of_birth) as birth_year 
+            	from hai_az_prod.person_msdw_2020july pmj
+            )
+            select birth_year, gender, count(*) as n 
+            from person_birth_year 
+            group by birth_year , gender 
+            order by birth_year, gender;
+            ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+
+}
+
+patient_birth_year_sex_distribution <- get_patient_birth_year_distribution()
+
+# lump together sex that can be treated as Unknown
+patient_birth_year_sex_distribution <- patient_birth_year_sex_distribution %>% 
+  mutate(gender = recode(gender, 'MSDW_NOT APPL' = 'Unknown')) %>%
+  mutate(gender = recode(gender, 'MSDW_OTHER' = 'Unknown')) %>%
+  mutate(gender = recode(gender, 'MSDW_UNKNOWN' = 'Unknown')) %>%
+  mutate(gender = recode(gender, 'NOT AVAILABLE'= 'Unknown')) %>%
+  mutate(gender = recode(gender, .missing = 'Unknown')) %>%
+  # sample_n(50) %>%
+  # ignore ~ 1 million patients whoes birth_year is unknown
+  # filter(!is.na(birth_year)) %>%
+  # 2 patients have birth_year after 2020, which is clearly error as our database cut off is 2020
+  filter(birth_year <= 2020) %>%
+  mutate(birth_year_bucket = cut(birth_year, breaks = c(1800, seq(1931, 2021, by = 10)), dig.lab = 4)) %>%
+  group_by(birth_year_bucket, gender) %>% 
+  summarise(n = sum(n)) %>% 
+  arrange(birth_year_bucket)
+
+plot_by_birth_year_sex <- patient_birth_year_sex_distribution %>% ungroup() %>% 
+  filter(gender %in% c('Male', 'Female')) %>% 
+  mutate(n = if_else(gender == 'Female', -n, n)) %>%
+  mutate(birth_year_bucket = factor(birth_year_bucket, levels = rev(levels(birth_year_bucket)))) %>% 
+  # change bucket name (1800,1931] to '<= 1931'
+  # change bucket name (2011, 2021] to '(2011, 2020]' -- our database cutoff is 2020
+  mutate(birth_year_bucket = recode(birth_year_bucket, '(1800,1931]' = '<=1931')) %>% 
+  mutate(birth_year_bucket = recode(birth_year_bucket, '(2011,2021]' = '(2011,2020]')) %>%{
+    ggplot(.) +
+  geom_bar(aes(x = birth_year_bucket, y = n, fill = gender), stat = 'identity') + 
+  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral")) + 
+  scale_y_continuous(labels = abs) +
+  xlab('') + ylab('patient count') +
+  coord_flip() + 
+      theme_bw() + 
+  theme(panel.grid = element_blank(), legend.position = "right", 
+        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
+        legend.text = element_text(size = 8), legend.title = element_text(size = 10), 
+        legend.background = element_rect(fill=alpha('white', 0.1)))
+  }
+
+ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.png"), plot = plot_by_birth_year_sex, width = 4, height = 3)
+plot_by_birth_year_sex
+
+```
+
+### Count patient distribution by the number of HPO terms per patient 
+```{r}
+
+```
+
+### Collect patient counts for those with lab data at each year
+
+```{r}
+get_tested_patient_count_by_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/tested_patient_count_by_year.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, "
+            with scc_lab as (
+            	select medical_record_number , calendar_date as lab_date, date_part('Year', lab_date) as lab_year
+            	from hai_az_prod.lab_scc_2020july lsj
+            ), 
+            epic_lab as (
+            	select mrn as medical_record_number, order_date as lab_date, date_part('Year', lab_date) as lab_year
+            	from hai_az_test.lab_epic_2020july lej
+            ),
+            scc_and_epic as (
+            	select * 
+            	from scc_lab 
+            	union all
+            	select * 
+            	from epic_lab
+            )
+            select lab_year, count(distinct medical_record_number) as n_patient
+            from scc_and_epic 
+            group by lab_year
+            order by lab_year;
+            ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+tested_patient_count_by_year <- get_tested_patient_count_by_year()
+```
+
 ### Analyze changes of observed phenotypes
@@
 ```{r}
-plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df){
+plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df, normalize=TRUE){
   hpo_terms_selected <- df %>% filter(termid %in% terms_of_interest) %>% select(termid, label) %>% distinct()
+
+  y_str = ifelse(normalize, 'n_patient/n_tested', 'n_patient')
+  y_label = ifelse(normalize, 'patient fraction among tested', 'patient count')
 
@@
     filter(lab_year <= 2019 & lab_year > 2000) %>%
-    ggplot(aes(x = lab_year, y = n_patient, group = termid, color = termid)) + 
+    ggplot(aes(x = lab_year, y = eval(parse(text=y_str)), group = termid, color = termid)) + 
     geom_point() +
     geom_line() +
-    scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " "))
+    scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " ")) +
+    ylab(y_label)
 
@@
 hpo_level_1_terms <- phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
-plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=phenotype_rank_all_by_year)
+plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=phenotype_rank_all_by_year, normalize = FALSE)
+```
+
+```{r}
+hpo_level_1_terms <- phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
+plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df= phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
 ```
@@
     df <- dbGetQuery(dsca, "
-               with testable_phenotypes_direct as (
+               with testable_phenotypes_all as (
               	select lab.medical_record_number , lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly 
@@
               ),
-              testable_phenotype_direct_freq as (
+              testable_phenotype_all_freq as (
               	select termid, count(*) as n_record, count(distinct medical_record_number) as n_patient
-              	from testable_phenotypes_direct
+              	from testable_phenotypes_all
               	group by termid 
@@
               select testable.*, hpo.label, hpo.distancetoroot 
-              from testable_phenotype_direct_freq testable
+              from testable_phenotype_all_freq testable
               join hai_az_prod.hpo hpo on testable.termid = hpo.termid
@@
 
-Plot the anual change of tested phenotypes.
+Plot the annual change of tested phenotypes.
 
@@
 hpo_level_1_terms <- tested_phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
-plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year)
+plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year, normalize = FALSE)
+```
+```{r}
+plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
 ```
@@
 
+Find what local lab codes are mapped to HP:0010702 Increased circulating antibody level and find which are tested more often after 2017.
 ```{r}
-children_of_interest <- c('HP:0410367', 'HP:0031047', 'HP:0003237', 'HP:0003212', 'HP:0003261', 'HP:0410369', 'HP:0410371', 'HP:0410246', 'HP:0003496')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
+get_tested_phenotype_spike_by_loinc_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/tested_phenotype_spike_by_loinc_year.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <-  dbGetQuery(dsca, "
+              with lab_spike as (
+            	select context_name , lab.loinc , ltp.termid , date_part('Year', lab_date) as lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
+            	join hai_az_prod.loinc_testable_phenotypes ltp on lab.loinc = ltp.loincid 
+            	where ltp.termid = 'HP:0010702'
+            	group by context_name , lab.loinc , ltp.termid, date_part('Year', lab_date)
+            )
+            select l.*
+            from lab_spike l
+            order by context_name , loinc , lab_year;
+              ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+get_tested_phenotype_spike_by_localcode_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/tested_phenotype_spike_by_localcode_year.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <-  dbGetQuery(dsca, "
+              with lab_spike as (
+            	select context_name , local_test_code , ltp.termid, date_part('Year', lab_date) as lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
+            	join hai_az_prod.loinc_testable_phenotypes ltp on lab.loinc = ltp.loincid 
+            	where ltp.termid = 'HP:0010702'
+            	group by context_name , local_test_code ,ltp.termid,  date_part('Year', lab_date)
+            ),
+            local_lab_names as (
+            	select distinct source, code, test_name
+            	from hai_az_prod.loinc_mapping lm 
+            )
+            select l.*, r.test_name
+            from lab_spike l
+            join local_lab_names r on l.local_test_code = r.code
+            order by context_name , local_test_code , lab_year;
+              ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+tested_phenotype_spike_by_loinc_year <- get_tested_phenotype_spike_by_loinc_year()
+tested_phenotype_spike_by_localcode_year <- get_tested_phenotype_spike_by_localcode_year()
 ```
+
+We can see the spike is caused by LOINC:10834-0 Globulin [Mass/volume] in Serum by calculation.
+
+The SCC spike is for LOINC:2465-3 IgG [Mass/volume] in Serum or Plasma. The increase is significant, but the raw count is still negligible. 
+```{r}
+tested_phenotype_spike_by_loinc_year %>% 
+  left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year') %>%
+    filter(lab_year <= 2019 & lab_year > 2000) %>%
+  #filter(loinc == '10834-0') %>%
+    ggplot(aes(x = lab_year, y = n_patient/n_tested, group = loinc, color = loinc)) + 
+    geom_point() +
+    geom_line() +
+    #scale_color_discrete(breaks = loinc, labels = str_c('LOINC', loinc, sep = ":")) +
+    ylab('patient fraction among tested') +
+  facet_wrap(~context_name, nrow = 2, scales = 'free_y')
+```
+
+Similarly, if we use local codes for the plot, we can also see that EPIC lab test 3277 caused the spike, which has a name of "GLOBULIN". 
+```{r}
+tested_phenotype_spike_by_localcode_year %>% 
+  left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year') %>%
+    filter(lab_year <= 2019 & lab_year > 2000) %>%
+  group_by(context_name, local_test_code) %>%
+  mutate(min_year = min(lab_year), max_year = max(lab_year), max_min_year = max_year - min_year) %>%
+  filter(max_min_year > 2) %>%
+    ggplot(aes(x = lab_year, y = n_patient/n_tested, group = local_test_code, color = local_test_code)) + 
+    geom_point() +
+    geom_line() +
+    #scale_color_discrete(breaks = loinc, labels = str_c('LOINC', loinc, sep = ":")) +
+    ylab('patient fraction among tested') +
+  theme(legend.position = 'bottom') +
+  facet_wrap(~context_name, nrow = 2, scales = 'free_y')
+```

### 2022-03-08 src/main/r/biomarker.Rmd
update

---
title: "Biomarker screening"
output: html_notebook
---

In this section, we try to screen the phenotypes for biomarkers associated with pregression to severe asthma.

```{r message=FALSE}
library(tidyverse)
library(glue)
library(readxl)
require(DBI)
require(RPostgres)
require(egg)
require(gridExtra)
require(lubridate)
require(furrr)
source('.db_credential.R')
dsca <- DBI::dbConnect(RPostgres::Postgres(), host='localhost', user=user, dbname = dbname, port = 2345, password=password)
```

## Cohort 
The starting cohort are patients with any asthma (J45) and asthma-related diagnosis.

Within this starting cohort, we find one subset with severe asthma diagnosis. We further find a subset that has non-severe asthma diagnosis before they progressed to the severe form. These patients are the cases.

Another subcohort are the ones that have never been associated with severe asthma. These patients are the controls.

Final table has three columns:

- medical_record_number
- cohort: case or control
- t0: earliest severe asthma dx date for cases and last asthma diagnosis for controls

```{sql connection=dsca}
-- dx of asthma and asthma-related symptoms
create temp table asthma_related_dx as 
select medical_record_number , calendar_date , context_name , context_diagnosis_code 
from hai_az_prod.diagnosis_2020july dj 
where context_diagnosis_code ~* '^J45' or context_diagnosis_code ~* '^R06';
```

```{sql connection=dsca}
-- severe asthma
create temp table severe_asthma_patients as
select medical_record_number , min(calendar_date) as min_calendar_date
from hai_az_prod.diagnosis_2020july dj 
where context_diagnosis_code ~* '^J45\\.5'
group by medical_record_number;  
```

A subset of severe asthma patients who were progressed from non-severe to severe. These are cases. 
```{sql connection=dsca}
-- severe asthma progressed from non-severe: 2593
create temp table progressed_to_severe_asthma_patients as
select distinct medical_record_number 
from hai_az_prod.diagnosis_2020july dj 
join severe_asthma_patients using (medical_record_number)
where (context_diagnosis_code ~* '^J45(\\.)?[2349]' or context_diagnosis_code ~* '^R06') and calendar_date < min_calendar_date;
```

Patients who had asthma but not severe. These are controls. 
```{sql connection=dsca}
-- control : 323420
create temp table never_severe_asthma_patients as
select medical_record_number, max(calendar_date) as max_calendar_date, random() as random 
from asthma_related_dx
where medical_record_number not in (select medical_record_number from severe_asthma_patients)
group by medical_record_number ;
```

Combine cases and controls in one table. For now, we will use the 
```{sql connection=dsca}
-- combined cohort 
create temp table cohort as 
select medical_record_number , min_calendar_date as t0, 'case' as case_or_control
from severe_asthma_patients 
join progressed_to_severe_asthma_patients using (medical_record_number)
union all
select medical_record_number , max_calendar_date as t0, 'control' as case_or_control
from never_severe_asthma_patients
where random < 2593.0 * 2 / 323420;
```

## Extraction of data

We need to extract the following information:

- the cohort table: medical_record_number, cohort, t0
- tested phenotypess: medical_record_number, termid, testdate
- observed phenotypes: medical_record_number, termid, testdate
- demographics: medical_record_number, sex, race, birthdate

```{sql connection=dsca, output.var=cohort}
select * from cohort;
```

```{sql connection=dsca, output.var=tested_phenotypes}
with testable_phenotypes_all as (
	select lab.medical_record_number , lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly 
	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
	join cohort using (medical_record_number)
  )
 select medical_record_number, lab_date, termid
 from testable_phenotypes_all;
```

```{sql connection=dsca, output.var=observed_phenotypes}
with abnormal_lab as (
  select lab.*
  from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
  join cohort using (medical_record_number)
  where hpotermid is not null and not isnegated),
abnormal_lab_after_infer as (
  SELECT l.medical_record_number, l.lab_date, date_part(year, lab_date) as lab_year, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance
  FROM abnormal_lab l 
  left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair.current)
select medical_record_number , lab_date, ancestor as termid
from abnormal_lab_after_infer;
```

```{sql connection=dsca, output.var='cohort_demographics'}
select person.*
from hai_az_prod.person_msdw_2020july person
join cohort using (medical_record_number);
```

## Data transformation

- tested phenotypes: calculate testdate relative to t0; cut the relative date into yearly brackets and aggregated on yearly basis
- observed phenotypes: similar to the above
- demographics: calculate age at the time of t0

```{r}
cohort_tested_phenotypes_yearly_count <- tested_phenotypes %>%
  left_join(cohort, by = 'medical_record_number') %>%
  mutate(lab_date_rel = interval(t0, lab_date) / ddays()) %>% 
  filter(lab_date_rel < 0) %>%
  filter(lab_date_rel > -365 * 3) %>%
  mutate(lab_date_rel_bracket = cut(lab_date_rel / 365, breaks=seq(-3,0))) %>%
  group_by(medical_record_number, lab_date_rel_bracket, termid) %>% 
  summarise(termid_count = n(), .groups = 'drop')
```
```{r}
cohort_observed_phenotypes_yearly_count <- observed_phenotypes %>% 
  left_join(cohort, by = 'medical_record_number') %>%
  mutate(lab_date_rel = interval(t0, lab_date) / ddays()) %>%
  filter(lab_date_rel < 0) %>%
  filter(lab_date_rel > -365 * 3) %>%
  mutate(lab_date_rel_bracket = cut(lab_date_rel / 365, breaks=seq(-3,0))) %>%
  group_by(medical_record_number, lab_date_rel_bracket, termid) %>% 
  summarise(termid_count = n(), .groups = 'drop')
```

```{r}
## collect all the phenotypes in one dataframe
phenotype_features <- cohort_tested_phenotypes_yearly_count %>% 
  rename(tested_count = termid_count) %>%
  left_join(cohort_observed_phenotypes_yearly_count %>% rename(observed_count = termid_count), 
            by = c('medical_record_number', 'lab_date_rel_bracket', 'termid')) %>%
  mutate(observed_count = ifelse(is.na(observed_count), 0, observed_count)) %>% 
  mutate(is_tested = tested_count > 0) %>%
  mutate(is_observed = observed_count > 0)
```

The final data frame should have the following columns:
medical_record_number: patient id 
termid: phenotype
is_tested: TRUE or FALSE
is_observed: TRUE or FALSE
time_window: [-3, -2), [-2, -1), [-1, 0)
sex
race
age: age at first severe asthma diagnosis (cases) or last asthma diagnosis (control)
outcome: progressed to severe asthma or not

## Statistical analysis
For the initial analysis, the model is as below:

outcome ~ is_tested + is_tested:is_observed

```{r}
confounders_outcome <- cohort %>% 
  inner_join(cohort_demographics, by = 'medical_record_number') %>%
  mutate(age = interval(date_of_birth, t0) / dyears()) %>%
  select(medical_record_number, gender, race, age, case_or_control) %>% 
  filter(!is.na(age))

d <- phenotype_features %>% 
  filter(termid == 'HP:0001880') %>%
  filter(lab_date_rel_bracket == '(-1,0]')

d <- phenotype_features %>% 
  filter(termid == 'HP:0001873') %>%
  filter(lab_date_rel_bracket == '(-1,0]')

prepare_Xy <- function(df_phenotype_feature, df_confounders_outcome){
  df_model <- df_confounders_outcome %>% 
    left_join(df_phenotype_feature, by = 'medical_record_number') %>%
    select(medical_record_number, gender, race, age, tested_count, observed_count, is_tested, is_observed, case_or_control) %>%
    ## fill in missing values
    replace_na(list(tested_count = 0, observed_count = 0, is_tested = FALSE, is_observed = FALSE)) %>%
    ## factorize categorical variables
    mutate(gender = factor(gender)) %>%
    mutate(race = factor(race)) %>%
    mutate(is_tested = factor(is_tested)) %>%
    mutate(is_observed = factor(is_observed)) %>%
    mutate(case_or_control = factor(case_or_control)) %>%
    ## set the reference levels for categorical variables
    mutate(gender = relevel(gender, ref='Male')) %>%
    mutate(race = relevel(race, ref='White')) %>%
    mutate(is_tested = relevel(is_tested, ref = 'FALSE')) %>%
    mutate(is_observed = relevel(is_observed, ref = 'FALSE')) %>%
    mutate(case_or_control = relevel(case_or_control, ref = 'control'))
  return (df_model)
}

glm_Xy <- function(Xy){
  model <- glm(formula = 'case_or_control ~ gender + race + age + is_tested / is_observed', data = Xy, family = binomial(link = 'logit'))
  model_coeff <- broom::tidy(model)
  return (model_coeff)
}

df_model <- prepare_Xy(df_phenotype_feature = d %>% select(-lab_date_rel_bracket, -termid), 
                       df_confounders_outcome = confounders_outcome)

# df_model %>% sample_n(10)

glm_Xy(df_model) %>% filter(str_starts(term, 'is_testedTRUE'))

fit_model <- function(df_phenotype_feature, df_confounders_outcome) {
  df_model <- prepare_Xy(df_phenotype_feature, df_confounders_outcome)
  model_coeff <- data.frame(term = character(), estimate = double(), std.error = double(), statistic = double(), p.value = double())
  tryCatch({model_coeff = glm_Xy(df_model)}, error = function(e) print('cannot fit model'))
  #error = function(e) print('model did not run'))
  return (model_coeff)
}

fit_model(df_phenotype_feature = d %>% select(-lab_date_rel_bracket, -termid), 
                       df_confounders_outcome = confounders_outcome)

## explore other statistical models
model2 <- glmer(formula = 'case_or_control ~ gender + race + age + is_observed + (1 + is_observed | is_tested)', data = df_model, family = binomial(link = 'logit') )
summary(model2)
```

Screen all the phenotypes

```{r}
tic <- Sys.time()

plan(multisession, workers = 6)

screen_results2 <- phenotype_features %>% 
  #filter(termid == 'HP:0001880') %>%
  #filter(lab_date_rel_bracket == '(-1,0]') %>%
  group_by(lab_date_rel_bracket, termid) %>%
  nest() %>%
  #calling ungroup() is critical for furrr to work in parallel
  ungroup() %>%
  mutate(coefficient = future_map(data, fit_model, df_confounders_outcome = confounders_outcome)) %>%
  select(-data) %>%
  unnest(coefficient)

toc <- Sys.time()

toc - tic
```
```{r}
## how many biomarkers we have screened
## at each time window
screen_results %>% group_by(lab_date_rel_bracket) %>% summarise(n = n_distinct(termid))
```

Annotate the screen results by adding the HPO term label, and also create a few different views that rank the results differently. 
```{r}
hpo_term_list <- read.csv('/Users/xingminzhang/git/s4-pd-dmsdw/pd-dmsdw-loinc2hpo-app/src/main/resources/hp_term_list.csv', header = FALSE, col.names = c('termid', 'distance_to_root', 'label'), stringsAsFactors = FALSE)

screen_results_annotated <- screen_results %>%  filter(str_starts(term, 'is_testedTRUE')) %>% arrange(lab_date_rel_bracket) %>%
  left_join(hpo_term_list, by = 'termid') %>%
  select(lab_date_rel_bracket, termid, label, term, estimate, std.error, statistic, p.value)

write.csv(screen_results_annotated, file = '../cache/screen_results_annotated.csv', row.names = FALSE)

screen_results_annotated %>% head()
```

```{r}
# rank results by weight of is_tested
screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
  arrange(term) %>%
  mutate(term_rank = row_number()) %>%
  ungroup() %>% 
  arrange(lab_date_rel_bracket, termid) %>% 
  group_by(lab_date_rel_bracket, termid, label) %>%
  nest() %>% 
  mutate(ranking_metric_value = map_dbl(data, function(df) df$estimate[df$term == 'is_testedTRUE'])) %>% 
  arrange(-ranking_metric_value) %>% 
  mutate(overall_rank = row_number()) %>%
  unnest(cols = c(data)) %>%
  write.csv(file = '../cache/screen_results_annotated_rank_by_weight_of_is_tested.csv', row.names = FALSE)

# rank results by weight of is_testedTRUE:is_observedTRUE
screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
  arrange(term) %>%
  mutate(term_rank = row_number()) %>%
  ungroup() %>% 
  arrange(lab_date_rel_bracket, termid) %>% 
  group_by(lab_date_rel_bracket, termid, label) %>%
  nest() %>% 
  mutate(ranking_metric_value = map_dbl(data, function(df) df$estimate[df$term == 'is_testedTRUE:is_observedTRUE'])) %>% 
  arrange(-ranking_metric_value) %>% 
  mutate(overall_rank = row_number()) %>%
  unnest(cols = c(data)) %>%
  write.csv(file = '../cache/screen_results_annotated_rank_by_weight_of_is_tested_is_observed.csv', row.names = FALSE)

# rank results by p value of is_tested
screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
  arrange(term) %>%
  mutate(term_rank = row_number()) %>%
  ungroup() %>% 
  arrange(lab_date_rel_bracket, termid) %>% 
  group_by(lab_date_rel_bracket, termid, label) %>%
  nest() %>% 
  mutate(ranking_metric_value = map_dbl(data, function(df) df$p.value[df$term == 'is_testedTRUE'])) %>% 
  arrange(ranking_metric_value) %>% 
  mutate(overall_rank = row_number()) %>%
  unnest(cols = c(data)) %>%
  write.csv(file = '../cache/screen_results_annotated_rank_by_p_of_is_tested.csv', row.names = FALSE)

# rank results by p value of is_testedTRUE:is_observedTRUE
screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
  arrange(term) %>%
  mutate(term_rank = row_number()) %>%
  ungroup() %>% 
  arrange(lab_date_rel_bracket, termid) %>% 
  group_by(lab_date_rel_bracket, termid, label) %>%
  nest() %>% 
  mutate(ranking_metric_value = map_dbl(data, function(df) df$p.value[df$term == 'is_testedTRUE:is_observedTRUE'])) %>% 
  arrange(ranking_metric_value) %>% 
  mutate(overall_rank = row_number()) %>%
  unnest(cols = c(data)) %>%
  write.csv(file = '../cache/screen_results_annotated_rank_by_p_of_is_tested_is_observed.csv', row.names = FALSE)
```

## Cache data for future use
```{r}
cache_data <- TRUE

if (cache_data){
  cache_list <- list(confounders_outcome = confounders_outcome,
                     phenotype_features = phenotype_features)

  for (i in 1:length(cache_list)){
    filename <- names(cache_list)[[i]]
    write.table(cache_list[[i]], file = glue('../cache/{filename}.csv'), sep = ',', row.names = FALSE, col.names = TRUE)
  }
}
```

### 2022-03-11 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
update analysis

@@
 
+### Identify starting cohort
+The starting cohort could be the entirety of the MSDW set, but here we focus on the subcohort where we have lab data for. 
+```{sql connection=dsca}
+create temp table cohort as 
+select distinct medical_record_number 
+from hai_az_prod.lab_scc_2020july lsj 
+union 
+select distinct mrn as medical_record_number 
+from hai_az_test.lab_epic_2020july lej ;
+```
+
 ### Collect patient basic demographics
@@
             from hai_az_prod.person_msdw_2020july pmj 
+            join cohort using (medical_record_number)
             group by race, gender 
@@
 
-patient_race_sex_distribution <- get_patient_race_sex_distribution()
+patient_race_sex_distribution <- get_patient_race_sex_distribution(force_rerun = FALSE)
 
@@
 
+# the following cleaning are no longer needed after updating the race mapping file
 # recode "African American" as "Black or African American"
-# TODO: update the race map
-patient_race_sex_distribution <- patient_race_sex_distribution %>%
-  mutate(race = recode(race, 'African American' = 'Black or African American'))
+# patient_race_sex_distribution <- patient_race_sex_distribution %>%
+#   mutate(race = recode(race, 'African American' = 'Black or African American'))
 ```
@@
 plot_by_sex
-
 ```
@@
             from person_birth_year 
+            join cohort using (medical_record_number)
             group by birth_year , gender 
@@
 
-patient_birth_year_sex_distribution <- get_patient_birth_year_distribution()
+patient_birth_year_sex_distribution <- get_patient_birth_year_distribution(FALSE)
 
@@
 
+### Success rate of LOINC2HPO
+```{r}
+# among all the labs, what is the percentage that can be mapped to HPO
+get_loinc2hpo_success_metric <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/loinc2hpo_success_metric.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, "
+            with scc_lab_count as (
+            	select context_name , count(*) as n_test
+            	from hai_az_prod.lab_scc_2020july lsj 
+            	join hai_az_prod.fd_procedure fp using (procedure_key)
+            	group by context_name
+            ),
+            epic_lab as (
+            	select distinct mrn, order_date, lab_time , test_code, test_result_value  from prod_msdw.epic_lab el
+            ),
+            epic_lab_count as (
+            	select 'EPIC' as context_name, count(*) as n_test from epic_lab
+            ),
+            lab_count as (
+            	select * from scc_lab_count 
+            	union all 
+            	select * from epic_lab_count 
+            ),
+            loinc_success_count as (
+            	select context_name , count(*) as n_loinc_success
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july saelalhj 
+            	where loinc is not null
+            	group by context_name
+            ),
+            loinc2hpo_success_count as (
+            	select context_name , count(*) as n_loinc2hpo_success
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july saelalhj 
+            	where hpotermid is not null
+            	group by context_name 
+            )
+            select *, (n_loinc_success + 0.1)/n_test as loinc_success_rate, (n_loinc2hpo_success + 0.1)/n_test as loinc2hpo_success_rate
+            from lab_count
+            left join loinc_success_count using (context_name)
+            left join loinc2hpo_success_count using (context_name);
+            ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+
+}
+
+loinc2hpo_success_metric <- get_loinc2hpo_success_metric(TRUE)
+
+loinc2hpo_success_metric <- loinc2hpo_success_metric %>% mutate(across(function(x) is.numeric(x), function(x) replace_na(x, 0)))
+
+loinc2hpo_success_metric %>% mutate(across(function(x) is.double(x), function(x) round(x, digits = 3))) %>%
+  arrange(-n_test)
+```
+The overall success rate : `sum(loinc2hpo_success_metric$n_success) / sum(loinc2hpo_success_metric$n_test)`
+```{r}
+overall_success_rate <- sum(loinc2hpo_success_metric$n_success) / sum(loinc2hpo_success_metric$n_test)
+print(glue('Total Lab Results: {sum(loinc2hpo_success_metric$n_test)}, 
+           Total Lab Mapped to LOINC: {sum(loinc2hpo_success_metric$n_loinc_success)}
+           Total Lab Transformed to HPO: {sum(loinc2hpo_success_metric$n_loinc2hpo_success)}'))
+
+print(glue('Total LOINC Mapping Rate: {sum(loinc2hpo_success_metric$n_loinc_success)/sum(loinc2hpo_success_metric$n_test)},
+           Total HPO transformation Rate: {sum(loinc2hpo_success_metric$n_loinc2hpo_success)/sum(loinc2hpo_success_metric$n_test)}'))
+```
+
 ### Count patient distribution by the number of HPO terms per patient 
 ```{r}
+# prepare a summary table: summary_hpo_counts_per_patient 
+# rows: per patient
+# columns: medical_record_number, hpo, unique hpo, abnormal_hpo, unique_abnormal_hpo, unique_abnormal_hpo_inference
 
+# collect statistics for the starting cohort
+# The mapping procedure assigned an average of 633 laboratory test-derived HPO terms per individual patient, many of which were from the same laboratory tests performed at different visits.
+The tests corresponded to a mean of 57.7 unique HPO terms, of which 20.8 were abnormalities and the remainder were normal phenotypes (Fig.
+4e).
+The hierarchical structure of the HPO allows inferences to be propagated up to parent terms and their ancestors;14 using this method, we inferred an additional 51.2 HPO terms (total 73.5) based on 22.2 abnormalities for each patient
+# how many phenotypes are mapped from lab tests? # unique phenotypes? then break down to normal and abnormal
+# how many unique abnormal phenotypes for each patient? then after HPO inference
+
+get_hpo_counts_per_patient <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/hpo_counts_per_patient.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, "
+            with count_no_inference as (
+            	select medical_record_number , count(*) as n_hpo, count(distinct case when isnegated then 'NOT' else '' end || hpotermid) as n_distinct_hpo
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
+            	where hpotermid is not null
+            	group by medical_record_number  
+            ),
+            count_abnormal_no_inference as (
+            	select medical_record_number , count(distinct hpotermid) as n_distinct_abnormal_hpo
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
+            	where hpotermid is not null and not isnegated 
+            	group by medical_record_number
+            ),
+            count_abnormal_with_inference as (
+            	select medical_record_number , count(distinct ancestor) as n_distinct_abnormal_hpo_infer
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
+            	join hai_az_prod.hpo_is_a_pairs hiap on lab.hpotermid = hiap.current 
+            	where hpotermid is not null and not isnegated 
+            	group by medical_record_number
+            )
+            select c1.n_hpo, c1.n_distinct_hpo, c2.n_distinct_abnormal_hpo, c3.n_distinct_abnormal_hpo_infer 
+            from cohort p 
+            left join count_no_inference c1 using (medical_record_number) 
+            left join count_abnormal_no_inference c2 using (medical_record_number)
+            left join count_abnormal_with_inference c3 using (medical_record_number);
+            ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+hpo_counts_per_patient <- get_hpo_counts_per_patient(TRUE)
+hpo_counts_per_patient <- hpo_counts_per_patient %>% 
+  replace_na(list(n_hpo = 0, n_distinct_hpo = 0, n_distinct_abnormal_hpo = 0, n_distinct_abnormal_hpo_infer = 0))
+```
+
+Report out the patient level phenotype counts
+```{r}
+# mean
+sapply(hpo_counts_per_patient, mean)
+
+# median 
+sapply(hpo_counts_per_patient, median)
+```
+
+Plot the distribution of distinct phenotypes for each patient (including abnormal and abnormal)
+```{r}
+plot_by_phenotypes_per_patient <- ggplot(hpo_counts_per_patient) +
+  geom_histogram(aes(x = n_distinct_hpo, y = ..count../1000), binwidth = 10, fill = 'orange', color = 'black', size = 0.25) +
+  xlab('number of unique phenotypes for each patient') +
+  ylab('patient counts (in thousands)') + 
+  theme_bw() +
+  theme(panel.grid = element_blank(),
+        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
+        legend.text = element_text(size = 8), legend.title = element_text(size = 10))
+ggsave(filename = glue("{CACHE_DIR}/plotByPhenotypesPerPatient.png"), plot = plot_by_phenotypes_per_patient, width = 3, height = 2)
+plot_by_phenotypes_per_patient
+```
+Plot distribution of abnormal phenotypes (distinct) after inference
+```{r}
+plot_by_distinct_abnormal_hpo_infer_per_patient <- ggplot(hpo_counts_per_patient) +
+  geom_histogram(aes(x = n_distinct_abnormal_hpo_infer, y = ..count../1000), binwidth = 10, fill = 'orange', color = 'black', size = 0.25) +
+  xlab('number of unique phenotypes for each patient') +
+  ylab('patient counts (in thousands)') + 
+  theme_bw() +
+  theme(panel.grid = element_blank(),
+        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
+        legend.text = element_text(size = 8), legend.title = element_text(size = 10))
+ggsave(filename = glue("{CACHE_DIR}/plotByDistinctAbnormalPhenotypeInferPerPatient.png"), plot = plot_by_distinct_abnormal_hpo_infer_per_patient, width = 3, height = 2)
+plot_by_distinct_abnormal_hpo_infer_per_patient
 ```

### 2022-03-17 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
update analysis

@@
 }
+
+FORCE_RERUN <- FALSE
 ```
@@
 
-tested_patient_count_by_year <- get_tested_patient_count_by_year()
+tested_patient_count_by_year <- get_tested_patient_count_by_year(force_rerun=FORCE_RERUN)
+```
+
+Collect tested patients by year, race and sex.
+
+```{r}
+get_tested_patient_count_by_year_race_sex <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/tested_patient_count_by_year_race_sex.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, "
+            with scc_lab as (
+            	select medical_record_number , calendar_date as lab_date, date_part('Year', lab_date) as lab_year
+            	from hai_az_prod.lab_scc_2020july lsj
+            ), 
+            epic_lab as (
+            	select mrn as medical_record_number, order_date as lab_date, date_part('Year', lab_date) as lab_year
+            	from hai_az_test.lab_epic_2020july lej
+            ),
+            scc_and_epic as (
+            	select * 
+            	from scc_lab 
+            	union all
+            	select * 
+            	from epic_lab
+            ),
+            scc_and_epic_with_race as (
+            	select lab.*, p.race, p.gender
+            	from scc_and_epic lab
+            	join hai_az_prod.person_msdw_2020july p using (medical_record_number)
+            )
+            select lab_year, race, gender, count(distinct medical_record_number) as n_patient
+            from scc_and_epic_with_race
+            group by lab_year, race, gender
+            order by lab_year, race, gender;
+            ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+tested_patient_count_by_year_race_sex <- get_tested_patient_count_by_year_race_sex(force_rerun=FORCE_RERUN)
 ```
@@
 
-phenotype_rank_all_by_year <- get_phenotype_rank_all_by_year()
+phenotype_rank_all_by_year <- get_phenotype_rank_all_by_year(force_rerun = FORCE_RERUN)
+
+get_phenotype_rank_all_by_year_race_sex <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/observed_phenotype_rank_all_by_year_race_sex.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, "
+      with abnormal_lab as 
+      	(select * 
+      	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
+      	where hpotermid is not null and not isnegated
+      	),
+      abnormal_lab_after_infer as 
+      	(SELECT l.medical_record_number, l.lab_date, date_part(year, lab_date) as lab_year, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance
+      	FROM abnormal_lab l 
+      	left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair.current),
+      lab_freq_by_year_race_sex as (
+      	select ancestor as termid, lab_year, p.race, p.gender, count(*) as n_record, count(distinct medical_record_number) as n_patient 
+      	from abnormal_lab_after_infer 
+      	join hai_az_prod.person_msdw_2020july p using (medical_record_number)
+      	group by ancestor, lab_year, p.race, p.gender
+      )
+      select *
+      from lab_freq_by_year_race_sex
+      join hai_az_prod.hpo h using (termid)
+      order by lab_year, distancetoroot;")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+phenotype_rank_all_by_year_race_sex <- get_phenotype_rank_all_by_year_race_sex(force_rerun = FORCE_RERUN)
 ```
@@
 ```{r}
+fancy_scientific <- function(l) {
+     # turn in to character string in scientific notation
+     l <- format(l, scientific = TRUE)
+     # quote the part before the exponent to keep all the digits
+     l <- gsub("^(.*)e", "'\\1'e", l)
+     # turn the 'e+' into plotmath format
+     l <- gsub("e", "%*%10^", l)
+     # return this as an expression
+     parse(text=l)
+}
+
 plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df, normalize=TRUE){
@@
   y_str = ifelse(normalize, 'n_patient/n_tested', 'n_patient')
-  y_label = ifelse(normalize, 'patient fraction among tested', 'patient count')
+  y_label = ifelse(normalize, 'patient fraction normalized to database size', 'patient count')
 
@@
     scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " ")) +
-    ylab(y_label)
+    xlab('Year') +
+    ylab(y_label) +
+    theme_bw() +
+    theme(panel.grid = element_blank(),
+        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
+        legend.text = element_text(size = 8), legend.title = element_text(size = 10))
 
@@
 hpo_level_1_terms <- phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
-plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=phenotype_rank_all_by_year, normalize = FALSE)
+p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=phenotype_rank_all_by_year, normalize = FALSE)
+ggsave(filename = glue('{CACHE_DIR}/plot_abnormal_phenotypes_by_year_level_1.pdf'), plot = p, width = 8, height = 4)
+p
 ```
@@
 hpo_level_1_terms <- phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
-plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df= phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
+p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df= phenotype_rank_all_by_year %>%
+                                           left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
+ggsave(filename = glue('{CACHE_DIR}/plot_abnormal_phenotypes_by_year_level_1_normalized.pdf'), plot = p, width = 8, height = 4)
+p
+```
+
+Which trend line in the above plot are statistically significant?
+```{r}
+df= phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year')
+trendline_lm_test <- df %>% 
+  filter(distancetoroot == 1) %>% 
+  filter(lab_year > 2000) %>% 
+  mutate(fraction = n_patient / n_tested) %>%
+  select(termid, lab_year, label, fraction) %>% 
+  group_by(termid, label) %>% 
+  nest() %>% 
+  mutate(lm_model = map(data, function(d) lm(fraction ~ lab_year, data = d))) %>%
+  mutate(lm_coeff = map(lm_model, broom::tidy)) %>%
+  select(-data, -lm_model) %>% 
+  unnest(lm_coeff) %>% 
+  filter(term == 'lab_year') %>% 
+  mutate(p.value.bonferroni = p.value * 10)
+
+symnum(trendline_lm_test$p.value.bonferroni, cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, Inf), symbols = c("***", "**", "*", ".", " "))
+
+trendline_lm_test_formatted <- trendline_lm_test %>% select(termid, label, estimate, p.value, p.value.bonferroni) %>% 
+  arrange(p.value.bonferroni) %>%
+  rename(beta = estimate) %>% 
+  mutate(significance = symnum(p.value.bonferroni, cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, Inf), symbols = c("***", "**", "*", ".", " ")) ) %>% 
+  mutate(across(.cols = function(x) is.double(x), function(x) scales::scientific(x, digits = 3)))
+
+write.csv(trendline_lm_test_formatted, file = glue('{CACHE_DIR}/table_abnormal_phenotypes_by_year_level_1_normalized_trendline_lm_model.csv'), row.names=FALSE)
+
+trendline_lm_test_formatted
 ```
@@
 
-tested_phenotype_rank_direct <- get_tested_phenotype_rank_direct()
-tested_phenotype_rank_all <- get_tested_phenotype_rank_all()
+tested_phenotype_rank_direct <- get_tested_phenotype_rank_direct(force_rerun = FORCE_RERUN)
+tested_phenotype_rank_all <- get_tested_phenotype_rank_all(force_rerun = FORCE_RERUN)
 ```
@@
 
-For both cases, it would be interesting to learn about the changes of the lab phenotypes over time, and among different races.
+Plot the annual change of tested phenotypes.
+
+```{r}
+get_tested_phenotype_rank_all_by_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/tested_phenotype_rank_all_by_year.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <-  dbGetQuery(dsca, "
+              with testable_phenotypes_direct as (
+              	select lab.medical_record_number , person.race, lab.context_name , lab.lab_date , lab.lab_time , date_part(year, lab_date) as lab_year, lab.loinc , testable.termid , testable.is_tested_directly
+              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
+              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
+              	join hai_az_prod.person_msdw_2020july person using (medical_record_number)
+              	--where testable.is_tested_directly 
+              ),
+              testable_phenotype_direct_freq as (
+              	select termid, lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
+              	from testable_phenotypes_direct
+              	group by termid, lab_year
+              )
+              select testable.*, hpo.label, hpo.distancetoroot 
+              from testable_phenotype_direct_freq testable
+              join hai_az_prod.hpo hpo on testable.termid = hpo.termid
+              order by termid, lab_year;
+              ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+tested_phenotype_rank_all_by_year <- get_tested_phenotype_rank_all_by_year(force_rerun = FORCE_RERUN)
+```
+
+```{r}
+hpo_level_1_terms <- tested_phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
+p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year, normalize = FALSE)
+ggsave(filename = glue('{CACHE_DIR}/plot_tested_phenotypes_by_year_level_1.pdf'), plot = p, width = 8, height = 4)
+p
+```
+```{r}
+p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
+ggsave(filename = glue('{CACHE_DIR}/plot_tested_phenotypes_by_year_level_1_normalized.pdf'), plot = p, width = 8, height = 4)
+p
+```
+
+```{r}
+df= tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year')
+trendline_lm_test <- df %>% 
+  filter(distancetoroot == 1) %>% 
+  filter(lab_year > 2000) %>% 
+  mutate(fraction = n_patient / n_tested) %>%
+  select(termid, lab_year, label, fraction) %>% 
+  group_by(termid, label) %>% 
+  nest() %>% 
+  mutate(lm_model = map(data, function(d) lm(fraction ~ lab_year, data = d))) %>%
+  mutate(lm_coeff = map(lm_model, broom::tidy)) %>%
+  select(-data, -lm_model) %>% 
+  unnest(lm_coeff) %>% 
+  filter(term == 'lab_year') %>% 
+  mutate(p.value.bonferroni = p.value * 11)
+
+trendline_lm_test_formatted <- trendline_lm_test %>% select(termid, label, estimate, p.value, p.value.bonferroni) %>% 
+  arrange(p.value.bonferroni) %>%
+  rename(beta = estimate) %>% 
+  mutate(significance = symnum(p.value.bonferroni, cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, Inf), symbols = c("***", "**", "*", ".", " ")) ) %>% 
+  mutate(across(.cols = function(x) is.double(x), function(x) scales::scientific(x, digits = 3)))
+
+write.csv(trendline_lm_test_formatted, file = glue('{CACHE_DIR}/table_tested_phenotypes_by_year_level_1_normalized_trendline_lm_model.csv'), row.names=FALSE)
+
+trendline_lm_test_formatted
+```
+
+Some trends are pretty interesting, such as the spike of Abnormal cellular phenotype. The term has three children, but only one is tested: HP:0011017 Abnormal cellular physiology
+
+```{r}
+children_abnormal_cellular_physiology <- c('HP:0003220', 'HP:0011018', 'HP:0025463', 'HP:0011019', 'HP:0003254', 'HP:4000056', 'HP:0011133', 'HP:0004371', 'HP:0031377', 'HP:0031409', 'HP:0012103', 'HP:0032576', 'HP:0003204', 'HP:0004356', 'HP:0032571', 'HP:0002916', 'HP:0003224', 'HP:0003575', 'HP:0003358')
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_abnormal_cellular_physiology, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'))
+```
+
+Again, the above plot shows that it is driven by one HP:0031409 Abnormal lymphocyte physiology. 
+```{r}
+children_of_interest <- c('HP:0011840', 'HP:0030886', 'HP:0012177', 'HP:0005372', 'HP:0031378')
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'))
+```
+```{r}
+children_of_interest <- c('HP:0002847', 'HP:0002959', 'HP:0005384', 'HP:0041076', 'HP:0010701')
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'))
+```
+
+```{r}
+children_of_interest <- c('HP:0010702', 'HP:0032336', 'HP:0410241', 'HP:0410240', 'HP:0410243', 'HP:0410242', 'HP:0410244', 'HP:0004313')
+plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'))
+```
+
+Find what local lab codes are mapped to HP:0010702 Increased circulating antibody level and find which are tested more often after 2017.
+```{r}
+get_tested_phenotype_spike_by_loinc_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/tested_phenotype_spike_by_loinc_year.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <-  dbGetQuery(dsca, "
+              with lab_spike as (
+            	select context_name , lab.loinc , ltp.termid , date_part('Year', lab_date) as lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
+            	join hai_az_prod.loinc_testable_phenotypes ltp on lab.loinc = ltp.loincid 
+            	where ltp.termid = 'HP:0010702'
+            	group by context_name , lab.loinc , ltp.termid, date_part('Year', lab_date)
+            )
+            select l.*
+            from lab_spike l
+            order by context_name , loinc , lab_year;
+              ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+get_tested_phenotype_spike_by_localcode_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/tested_phenotype_spike_by_localcode_year.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <-  dbGetQuery(dsca, "
+              with lab_spike as (
+            	select context_name , local_test_code , ltp.termid, date_part('Year', lab_date) as lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
+            	join hai_az_prod.loinc_testable_phenotypes ltp on lab.loinc = ltp.loincid 
+            	where ltp.termid = 'HP:0010702'
+            	group by context_name , local_test_code ,ltp.termid,  date_part('Year', lab_date)
+            ),
+            local_lab_names as (
+            	select distinct source, code, test_name
+            	from hai_az_prod.loinc_mapping lm 
+            )
+            select l.*, r.test_name
+            from lab_spike l
+            join local_lab_names r on l.local_test_code = r.code
+            order by context_name , local_test_code , lab_year;
+              ")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+tested_phenotype_spike_by_loinc_year <- get_tested_phenotype_spike_by_loinc_year(force_rerun = FORCE_RERUN)
+tested_phenotype_spike_by_localcode_year <- get_tested_phenotype_spike_by_localcode_year(force_rerun = FORCE_RERUN)
+```
+
+We can see the spike is caused by LOINC:10834-0 Globulin [Mass/volume] in Serum by calculation.
+
+The SCC spike is for LOINC:2465-3 IgG [Mass/volume] in Serum or Plasma. The increase is significant, but the raw count is still negligible. 
+```{r}
+tested_phenotype_spike_by_loinc_year %>% 
+  left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year') %>%
+    filter(lab_year <= 2019 & lab_year > 2000) %>%
+  #filter(loinc == '10834-0') %>%
+    ggplot(aes(x = lab_year, y = n_patient/n_tested, group = loinc, color = loinc)) + 
+    geom_point() +
+    geom_line() +
+    #scale_color_discrete(breaks = loinc, labels = str_c('LOINC', loinc, sep = ":")) +
+    ylab('patient fraction among tested') +
+  facet_wrap(~context_name, nrow = 2, scales = 'free_y')
+```
+
+Similarly, if we use local codes for the plot, we can also see that EPIC lab test 3277 caused the spike, which has a name of "GLOBULIN". 
+```{r}
+tested_phenotype_spike_by_localcode_year %>% 
+  left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year') %>%
+    filter(lab_year <= 2019 & lab_year > 2000) %>%
+  group_by(context_name, local_test_code) %>%
+  mutate(min_year = min(lab_year), max_year = max(lab_year), max_min_year = max_year - min_year) %>%
+  filter(max_min_year > 2) %>%
+    ggplot(aes(x = lab_year, y = n_patient/n_tested, group = local_test_code, color = local_test_code)) + 
+    geom_point() +
+    geom_line() +
+    #scale_color_discrete(breaks = loinc, labels = str_c('LOINC', loinc, sep = ":")) +
+    ylab('patient fraction among tested') +
+  theme(legend.position = 'bottom') +
+  facet_wrap(~context_name, nrow = 2, scales = 'free_y')
+```
+
+### Analyze racial differences
+
+#### racial differences of tested phenotypes
+
+The section analyzed whether there are racial differences in tested phenotypes. The problem is that we saw the vast majority of terms have differences. This is probably just a large number effect.
 
@@
 
-tested_phenotype_rank_direct_by_race <- get_tested_phenotype_rank_direct_by_race()
-tested_phenotype_rank_all_by_race <- get_tested_phenotype_rank_all_by_race()
+tested_phenotype_rank_direct_by_race <- get_tested_phenotype_rank_direct_by_race(force_rerun = FORCE_RERUN)
+tested_phenotype_rank_all_by_race <- get_tested_phenotype_rank_all_by_race(force_rerun = FORCE_RERUN)
 ```
 
-For each directly measured phenotype, we can determine whether race and the tests are correlated by some form of Chi-squared teste.
+For each directly measured phenotype, we can determine whether race and the tests are correlated by some form of Chi-squared test.
+The current problem is that the most phenotype terms are statistically significant.
+@TODO we should look at phenotypes at each year. 
 ```{r}
-tested_phenotype_rank_direct_by_race %>% filter(race == 'African American') %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
-tested_phenotype_rank_direct_by_race %>% filter(race == 'Asian') %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
-tested_phenotype_rank_direct_by_race %>% filter(race == 'White') %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
+countByRace <- patient_race_sex_distribution %>% 
+  group_by(race) %>% 
+  summarise(dbtotal = sum(n))
+
+phenotype_race_chisq_test <- function(df) {
+  broom::tidy(chisq.test(df$observed, p = df$dbtotal/sum(df$dbtotal)))
+}
+
+phenotype_race_prop_test <- function(df){
+  broom::tidy(prop.test(df$observed, p = df$observed/sum(df$dbtotal)))
+}
+
+tested_phenotype_rank_direct_by_race %>% 
+  select(termid, label, race, n_patient) %>% 
+  rename(observed = n_patient) %>% 
+  group_by(termid, label) %>% 
+  nest() %>% 
+  mutate(data = map(data, function(d) d %>% 
+                      right_join(countByRace, by = 'race') %>%
+                      replace_na(list(termid = first(.$termid), label = first(.$label), observed = 0)))) %>% 
+  mutate(chisq = map(data, phenotype_race_chisq_test)) %>% 
+  select(-data) %>% 
+  unnest(chisq) %>% 
+  arrange(-p.value)
 ```
@@
 ```{r}
-tested_phenotype_rank_all_by_race %>% filter(race == 'African American' & distancetoroot == 1) %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
-tested_phenotype_rank_all_by_race %>% filter(race == 'Asian' & distancetoroot == 1) %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
-tested_phenotype_rank_all_by_race %>% filter(race == 'White' & distancetoroot == 1) %>% arrange(-n_record) %>% head(10) %>% formattable::formattable()
+tested_phenotype_rank_all_by_race %>% filter(distancetoroot == 1) %>% 
+  select(termid, label, race, n_patient) %>% 
+  rename(observed = n_patient) %>% 
+  group_by(termid, label) %>% 
+  nest() %>% 
+  mutate(data = map(data, function(d) d %>% 
+                      right_join(countByRace, by = 'race') %>%
+                      replace_na(list(termid = first(.$termid), label = first(.$label), observed = 0)))) %>% 
+  mutate(chisq = map(data, phenotype_race_chisq_test)) %>% 
+  select(-data) %>% 
+  unnest(chisq) %>% 
+  arrange(-p.value)
 ```
 
-Plot the annual change of tested phenotypes.
+#### racial differences of observed abnormal phenotypes by races
+An improved version of racial analysis: 
+Question: for each race, is it more likely to have a certain abnormal phenotype than the rest of the population?
 
+To do this, we will run Chisq test but on annual basis. 
 ```{r}
-get_tested_phenotype_rank_all_by_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
-  df_path = glue('{cache_dir}/tested_phenotype_rank_all_by_year.csv')
+tested_patient_count_by_year_race <- tested_patient_count_by_year_race_sex %>% 
+  group_by(lab_year, race) %>% 
+  summarise(dbtotal = sum(n_patient), .groups = 'drop')
 
-  if (force_rerun | !file.exists(df_path)){
-    df <-  dbGetQuery(dsca, "
-              with testable_phenotypes_direct as (
-              	select lab.medical_record_number , person.race, lab.context_name , lab.lab_date , lab.lab_time , date_part(year, lab_date) as lab_year, lab.loinc , testable.termid , testable.is_tested_directly
-              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
-              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
-              	join hai_az_prod.person_msdw_2020july person using (medical_record_number)
-              	--where testable.is_tested_directly 
-              ),
-              testable_phenotype_direct_freq as (
-              	select termid, lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
-              	from testable_phenotypes_direct
-              	group by termid, lab_year
-              )
-              select testable.*, hpo.label, hpo.distancetoroot 
-              from testable_phenotype_direct_freq testable
-              join hai_az_prod.hpo hpo on testable.termid = hpo.termid
-              order by termid, lab_year;
-              ")
-
-    write.csv(df, file = df_path, row.names=FALSE)
-  }
-
-  df <- read.csv(df_path)
-
-  return (df)
-}
-
-tested_phenotype_rank_all_by_year <- get_tested_phenotype_rank_all_by_year()
+tested_patient_count_by_year_race %>% filter(lab_year == 2019)
 ```
@@
 ```{r}
-hpo_level_1_terms <- tested_phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
-plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year, normalize = FALSE)
-```
-```{r}
-plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
-```
+RACE_OF_INTEREST <- 'Black or African American'
+df <- phenotype_rank_all_by_year_race_sex %>% 
+  group_by(termid, label, lab_year, race) %>% 
+  summarise(observed = sum(n_patient), .groups = 'drop') %>% 
+  right_join(tested_patient_count_by_year_race, by = c('lab_year', 'race')) %>% 
+  filter(lab_year == 2019) %>% 
+  ##filter(termid == 'HP:0003351')
+  filter(termid == 'HP:0000079')
 
-Some trends are pretty interesting, such as the spike of Abnormal cellular phenotype. The term has three children, but only one is tested: HP:0011017 Abnormal cellular physiology
+cm <- df %>%
+  mutate(roi = if_else(race == RACE_OF_INTEREST, 'isROI', 'notROI')) %>% 
+  group_by(roi) %>% 
+  summarise(observed = sum(observed), dbtotal = sum(dbtotal)) %>% 
+  ungroup() %>%
+  mutate(not_observed = dbtotal - observed) %>% 
+  select(-dbtotal) %>%
+  column_to_rownames('roi') %>% 
+  t() %>% 
+  as.matrix()
 
-```{r}
-children_abnormal_cellular_physiology <- c('HP:0003220', 'HP:0011018', 'HP:0025463', 'HP:0011019', 'HP:0003254', 'HP:4000056', 'HP:0011133', 'HP:0004371', 'HP:0031377', 'HP:0031409', 'HP:0012103', 'HP:0032576', 'HP:0003204', 'HP:0004356', 'HP:0032571', 'HP:0002916', 'HP:0003224', 'HP:0003575', 'HP:0003358')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_abnormal_cellular_physiology, df=tested_phenotype_rank_all_by_year)
-```
+broom::tidy(fisher.test(cm))
 
-Again, the above plot shows that it is driven by one HP:0031409 Abnormal lymphocyte physiology. 
-```{r}
-children_of_interest <- c('HP:0011840', 'HP:0030886', 'HP:0012177', 'HP:0005372', 'HP:0031378')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
-```
-```{r}
-children_of_interest <- c('HP:0002847', 'HP:0002959', 'HP:0005384', 'HP:0041076', 'HP:0010701')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
-```
-
-```{r}
-children_of_interest <- c('HP:0010702', 'HP:0032336', 'HP:0410241', 'HP:0410240', 'HP:0410243', 'HP:0410242', 'HP:0410244', 'HP:0004313')
-plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
-```
-
-Find what local lab codes are mapped to HP:0010702 Increased circulating antibody level and find which are tested more often after 2017.
-```{r}
-get_tested_phenotype_spike_by_loinc_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
-  df_path = glue('{cache_dir}/tested_phenotype_spike_by_loinc_year.csv')
-
-  if (force_rerun | !file.exists(df_path)){
-    df <-  dbGetQuery(dsca, "
-              with lab_spike as (
-            	select context_name , lab.loinc , ltp.termid , date_part('Year', lab_date) as lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
-            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
-            	join hai_az_prod.loinc_testable_phenotypes ltp on lab.loinc = ltp.loincid 
-            	where ltp.termid = 'HP:0010702'
-            	group by context_name , lab.loinc , ltp.termid, date_part('Year', lab_date)
-            )
-            select l.*
-            from lab_spike l
-            order by context_name , loinc , lab_year;
-              ")
-
-    write.csv(df, file = df_path, row.names=FALSE)
+## write a function to test the statistical difference for each race
+racial_abnormality_statistical_test <- function(df){
+  if (nrow(df) <= 1) {
+    return (NULL)
+  }
+  fisher_tests_races <- list()
+  for (RACE_OF_INTEREST in df$race){
+    cm <- df %>%
+        mutate(roi = if_else(race == RACE_OF_INTEREST, 'isROI', 'notROI')) %>% 
+        group_by(roi) %>% 
+        summarise(observed = sum(observed), dbtotal = sum(dbtotal)) %>% 
+        ungroup() %>%
+        mutate(not_observed = dbtotal - observed) %>% 
+        select(-dbtotal) %>%
+        column_to_rownames('roi') %>% 
+        t() %>% 
+        as.matrix()
+    if (length(cm) == 4){
+       fisher_test_coeff <- broom::tidy(fisher.test(cm))
+    } else {
+       fisher_test_coeff <- data.frame(estimate = numeric(), p.value = numeric(), conf.low = numeric(), conf.high = numeric(), method = '', alternative = '')
+    }
+    fisher_test_coeff <- fisher_test_coeff %>% mutate(race = RACE_OF_INTEREST) %>% 
+      select(race, everything())
+    fisher_tests_races[[length(fisher_tests_races) + 1]] <- fisher_test_coeff
+  }
+  fisher_test_coeffs <- bind_rows(fisher_tests_races)
+  ##return (df %>% left_join(fisher_test_coeffs, by = 'race'))
+  return (fisher_test_coeffs)
 }
 
-  df <- read.csv(df_path)
+racial_abnormality_statistical_test(df)
 
-  return (df)
+library(furrr)
+
+plan(multisession, workers = 8)
+
+racial_abnormality_fisher <- phenotype_rank_all_by_year_race_sex %>% 
+  group_by(termid, label, lab_year, race) %>% 
+  summarise(observed = sum(n_patient), .groups = 'drop') %>% 
+  right_join(tested_patient_count_by_year_race, by = c('lab_year', 'race')) %>% 
+  #filter(lab_year == 2019) %>% 
+  #filter(termid %in% c('HP:0003351', 'HP:0000079')) %>% 
+  group_by(lab_year, termid, label) %>% 
+  nest() %>% 
+  ungroup() %>% 
+  mutate(fisher_coeff = future_map(data, racial_abnormality_statistical_test)) %>% 
+  select(-data) %>% 
+  unnest(fisher_coeff)
+
+racial_abnormality_fisher %>% nrow()
+
+write.csv(racial_abnormality_fisher, file = glue('{CACHE_DIR}/racial_abnormality_fisher.csv'), row.names = FALSE)
+
+racial_abnormality_fisher %>% filter(between(lab_year, 2000, 2019)) %>% 
+  filter(race == 'Black or African American') %>% 
+  filter(termid %in% hpo_level_1_terms$termid) %>% 
+  {
+    ggplot(., aes(x = lab_year, y = estimate, group = termid, color = termid)) + 
+    geom_point() +
+    geom_line() +
+    scale_color_discrete(breaks = .$termid, labels = str_c(.$termid, .$label, sep = " ")) +
+    xlab('Year') +
+    ylab('odds ratio') +
+    theme_bw() +
+    theme(panel.grid = element_blank(),
+        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
+        legend.text = element_text(size = 8), legend.title = element_text(size = 10))
+
   }
 
-get_tested_phenotype_spike_by_localcode_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
-  df_path = glue('{cache_dir}/tested_phenotype_spike_by_localcode_year.csv')
-
-  if (force_rerun | !file.exists(df_path)){
-    df <-  dbGetQuery(dsca, "
-              with lab_spike as (
-            	select context_name , local_test_code , ltp.termid, date_part('Year', lab_date) as lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
-            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
-            	join hai_az_prod.loinc_testable_phenotypes ltp on lab.loinc = ltp.loincid 
-            	where ltp.termid = 'HP:0010702'
-            	group by context_name , local_test_code ,ltp.termid,  date_part('Year', lab_date)
-            ),
-            local_lab_names as (
-            	select distinct source, code, test_name
-            	from hai_az_prod.loinc_mapping lm 
-            )
-            select l.*, r.test_name
-            from lab_spike l
-            join local_lab_names r on l.local_test_code = r.code
-            order by context_name , local_test_code , lab_year;
-              ")
-
-    write.csv(df, file = df_path, row.names=FALSE)
-  }
-
-  df <- read.csv(df_path)
-
-  return (df)
-}
-
-tested_phenotype_spike_by_loinc_year <- get_tested_phenotype_spike_by_loinc_year()
-tested_phenotype_spike_by_localcode_year <- get_tested_phenotype_spike_by_localcode_year()
 ```
-
-We can see the spike is caused by LOINC:10834-0 Globulin [Mass/volume] in Serum by calculation.
-
-The SCC spike is for LOINC:2465-3 IgG [Mass/volume] in Serum or Plasma. The increase is significant, but the raw count is still negligible. 
-```{r}
-tested_phenotype_spike_by_loinc_year %>% 
-  left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year') %>%
-    filter(lab_year <= 2019 & lab_year > 2000) %>%
-  #filter(loinc == '10834-0') %>%
-    ggplot(aes(x = lab_year, y = n_patient/n_tested, group = loinc, color = loinc)) + 
-    geom_point() +
-    geom_line() +
-    #scale_color_discrete(breaks = loinc, labels = str_c('LOINC', loinc, sep = ":")) +
-    ylab('patient fraction among tested') +
-  facet_wrap(~context_name, nrow = 2, scales = 'free_y')
-```
-
-Similarly, if we use local codes for the plot, we can also see that EPIC lab test 3277 caused the spike, which has a name of "GLOBULIN". 
-```{r}
-tested_phenotype_spike_by_localcode_year %>% 
-  left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year') %>%
-    filter(lab_year <= 2019 & lab_year > 2000) %>%
-  group_by(context_name, local_test_code) %>%
-  mutate(min_year = min(lab_year), max_year = max(lab_year), max_min_year = max_year - min_year) %>%
-  filter(max_min_year > 2) %>%
-    ggplot(aes(x = lab_year, y = n_patient/n_tested, group = local_test_code, color = local_test_code)) + 
-    geom_point() +
-    geom_line() +
-    #scale_color_discrete(breaks = loinc, labels = str_c('LOINC', loinc, sep = ":")) +
-    ylab('patient fraction among tested') +
-  theme(legend.position = 'bottom') +
-  facet_wrap(~context_name, nrow = 2, scales = 'free_y')
-```

### 2022-03-18 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
update analysis

@@
   group_by(birth_year_bucket, gender) %>% 
-  summarise(n = sum(n)) %>% 
+  summarise(n = sum(n), .groups = 'drop') %>% 
   arrange(birth_year_bucket)
@@
 
-loinc2hpo_success_metric <- get_loinc2hpo_success_metric(TRUE)
+loinc2hpo_success_metric <- get_loinc2hpo_success_metric(force_rerun = FORCE_RERUN)
 
@@
 
-hpo_counts_per_patient <- get_hpo_counts_per_patient(TRUE)
+hpo_counts_per_patient <- get_hpo_counts_per_patient(force_rerun = FORCE_RERUN)
 hpo_counts_per_patient <- hpo_counts_per_patient %>% 
@@
 
+get_tested_phenotype_rank_all_by_race_year_sex <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
+  df_path = glue('{cache_dir}/tested_phenotype_rank_all_by_race_year_sex.csv')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <-  dbGetQuery(dsca, "
+              with testable_phenotypes_direct as (
+              	select lab.medical_record_number , person.race, person.gender , lab.context_name , lab.lab_date , lab.lab_time , date_part(year, lab_date) as lab_year, lab.loinc , testable.termid , testable.is_tested_directly
+              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
+              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
+              	join hai_az_prod.person_msdw_2020july person using (medical_record_number)
+              --where testable.is_tested_directly 
+                ),
+                testable_phenotype_direct_freq as (
+                	select termid, lab_year, race, gender, count(*) as n_record, count(distinct medical_record_number) as n_patient
+                	from testable_phenotypes_direct
+                	group by termid, lab_year, race, gender
+                )
+                select testable.*, hpo.label, hpo.distancetoroot
+                from testable_phenotype_direct_freq testable
+                join hai_az_prod.hpo hpo on testable.termid = hpo.termid
+                order by termid, lab_year, race, gender;")
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
 tested_phenotype_rank_direct_by_race <- get_tested_phenotype_rank_direct_by_race(force_rerun = FORCE_RERUN)
 tested_phenotype_rank_all_by_race <- get_tested_phenotype_rank_all_by_race(force_rerun = FORCE_RERUN)
+tested_phenotype_rank_all_by_race_year_sex <- get_tested_phenotype_rank_all_by_race_year_sex(force_rerun = FORCE_RERUN)
 ```
@@
   unnest(chisq) %>% 
-  arrange(-p.value)
+  arrange(-p.value) %>% 
+  head()
 ```
@@
 
-racial_abnormality_fisher %>% filter(between(lab_year, 2000, 2019)) %>% 
-  filter(race == 'Black or African American') %>% 
+p <- racial_abnormality_fisher %>% filter(between(lab_year, 2003, 2019)) %>% 
+  #filter(race == 'Black or African American') %>% 
+  #filter(race == 'White') %>%
+  filter(race %in% c('White', 'Black or African American', 'Asian')) %>% 
   filter(termid %in% hpo_level_1_terms$termid) %>% 
@@
     scale_color_discrete(breaks = .$termid, labels = str_c(.$termid, .$label, sep = " ")) +
+    #geom_hline(yintercept = c(0.8, 1.2), color = 'red', linetype = 'dotted') +
+    annotate(geom='rect', ymin = 0.8, ymax = 1.2, xmin = -Inf, xmax = Inf, fill = 'magenta', alpha = 0.2) +
+    geom_hline(yintercept = 1, color = 'black') +
     xlab('Year') +
@@
     theme_bw() +
-    theme(panel.grid = element_blank(),
+    theme(panel.grid = element_blank(), legend.position = 'right',
         axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
-        legend.text = element_text(size = 8), legend.title = element_text(size = 10))
-
-  }
-
-```
+        legend.text = element_text(size = 8), legend.title = element_text(size = 10)) +
+    facet_wrap(~race, ncol = 1)
+  }
+
+ggsave(filename = glue('{CACHE_DIR}/plot_by_phenotype_racial_difference_over_years.pdf'), plot = p, width = 8, height = 8)
+
+p
+```
+
+```{r}
+RACE_OF_INTEREST <- 'Black or African American'
+df <- tested_phenotype_rank_all_by_race_year_sex %>% 
+  group_by(termid, label, lab_year, race) %>% 
+  summarise(observed = sum(n_patient), .groups = 'drop') %>% 
+  right_join(tested_patient_count_by_year_race, by = c('lab_year', 'race')) %>% 
+  filter(lab_year == 2019) %>% 
+  ##filter(termid == 'HP:0003351')
+  filter(termid == 'HP:0000079')
+
+cm <- df %>%
+  mutate(roi = if_else(race == RACE_OF_INTEREST, 'isROI', 'notROI')) %>% 
+  group_by(roi) %>% 
+  summarise(observed = sum(observed), dbtotal = sum(dbtotal)) %>% 
+  ungroup() %>%
+  mutate(not_observed = dbtotal - observed) %>% 
+  select(-dbtotal) %>%
+  column_to_rownames('roi') %>% 
+  t() %>% 
+  as.matrix()
+
+broom::tidy(fisher.test(cm))
+
+racial_abnormality_statistical_test(df)
+
+library(furrr)
+
+plan(multisession, workers = 8)
+
+racial_tested_abnormality_fisher <- tested_phenotype_rank_all_by_race_year_sex %>%
+  group_by(termid, label, lab_year, race) %>%
+  summarise(observed = sum(n_patient), .groups = 'drop') %>%
+  right_join(tested_patient_count_by_year_race, by = c('lab_year', 'race')) %>%
+  #filter(lab_year == 2019) %>%
+  #filter(termid %in% c('HP:0003351', 'HP:0000079')) %>%
+  group_by(lab_year, termid, label) %>%
+  nest() %>%
+  ungroup() %>%
+  mutate(fisher_coeff = future_map(data, racial_abnormality_statistical_test)) %>%
+  select(-data) %>%
+  unnest(fisher_coeff)
+
+racial_tested_abnormality_fisher %>% nrow()
+
+write.csv(racial_tested_abnormality_fisher, file = glue('{CACHE_DIR}/racial_tested_abnormality_fisher.csv'), row.names = FALSE)
+
+p <- racial_tested_abnormality_fisher %>% filter(between(lab_year, 2003, 2019)) %>%
+  #filter(race == 'Black or African American') %>%
+  #filter(race == 'White') %>%
+  filter(race %in% c('White', 'Black or African American', 'Asian')) %>%
+  filter(termid %in% hpo_level_1_terms$termid) %>%
+  {
+    ggplot(., aes(x = lab_year, y = estimate, group = termid, color = termid)) +
+    geom_point() +
+    geom_line() +
+    scale_color_discrete(breaks = .$termid, labels = str_c(.$termid, .$label, sep = " ")) +
+    #geom_hline(yintercept = c(0.8, 1.2), color = 'red', linetype = 'dotted') +
+    annotate(geom='rect', ymin = 0.8, ymax = 1.2, xmin = -Inf, xmax = Inf, fill = 'magenta', alpha = 0.2) +
+    geom_hline(yintercept = 1, color = 'black') +
+    xlab('Year') +
+    ylab('odds ratio') +
+    theme_bw() +
+    theme(panel.grid = element_blank(), legend.position = 'right',
+        axis.title = element_text(size = 8), axis.text = element_text(size = 8),
+        legend.text = element_text(size = 8), legend.title = element_text(size = 10)) +
+    facet_wrap(~race, ncol = 1)
+  }
+
+ggsave(filename = glue('{CACHE_DIR}/plot_by_tested_phenotype_racial_difference_over_years.pdf'), plot = p, width = 8, height = 8)
+
+p
+```
+
+For each race, identify the most over-tested phenotypes between 2003 and 2019. 
+```{r}
+overly_tested_abnormalities_by_each_race_top10 <- racial_tested_abnormality_fisher %>% 
+  ## bring in how many patients were tested for each abnormality
+  inner_join(tested_phenotype_rank_all_by_race_year_sex %>% 
+              group_by(termid, lab_year, race) %>% 
+              summarise(n_record = sum(n_record), n_patient = sum(n_patient), .groups = 'drop'), 
+             by = c('termid', 'lab_year', 'race')) %>% 
+  filter(race %in% c('White', 'Black or African American', 'Asian')) %>%
+  ## focus on data between 2003 and 2019
+  filter(between(lab_year, 2003, 2019)) %>% 
+  group_by(termid, label, race) %>% 
+  summarise(median_estimate = median(estimate), 
+            n_record = sum(n_record),
+            n_patient = sum(n_patient),
+            .groups = 'drop') %>% 
+  arrange(desc(median_estimate)) %>%
+  ## do not focus on rarely tested abnormalities
+  ## here we require at least 17 years X 100 patient/year have been tested for the abnormality
+  filter(n_patient > 1700) %>% 
+  ## find top 10 within each race
+  group_by(race) %>% 
+  slice_head(n=10) %>% 
+  ## give some nicer names
+  rename(median_OR = median_estimate,
+         lab_record = n_record,
+         patient_count = n_patient)
+
+write.csv(overly_tested_abnormalities_by_each_race_top10, file=glue('{CACHE_DIR}/overly_tested_abnormalities_by_each_race_top10.csv'), row.names = FALSE)
+
+overly_tested_abnormalities_by_each_race_top10 %>% 
+  ungroup() %>% 
+  #filter(race == 'Asian') %>% 
+  rename(OR = median_OR) %>% 
+  rename(`Lab Record` = lab_record) %>% 
+  rename(`Patient Count` = patient_count) %>% 
+  mutate(OR = round(OR, digits = 3)) %>%
+  group_by(race) %>% 
+  nest() %>% {
+    walk2(.$race, .$data, 
+          function(race_name, df) sjPlot::tab_df(df, 
+                                                 alternate.rows = TRUE, 
+                                                 title = glue("Top 10 Overly Tested Lab Phenotype for {race_name}"), 
+                                                 use.viewer = FALSE, 
+                                                 file = glue('{CACHE_DIR}/table_overly_tested_abnormalities_for_{race_name}_top10.html')))
+  }
+
+race_name = 'Asian'
+table_overly_tested_by_race <- function(race_name){
+  df <- overly_tested_abnormalities_by_each_race_top10 %>% 
+    ungroup() %>% 
+    filter(race == race_name) %>% 
+    rename(OR = median_OR) %>% 
+    rename(`Lab Record` = lab_record) %>% 
+    rename(`Patient Count` = patient_count) %>% 
+    mutate(OR = round(OR, digits = 3)) %>%
+    select(-race)
+
+  sjPlot::tab_df(df, alternate.rows = TRUE, 
+                     title = glue("Top 10 Overly Tested Lab Phenotype for {race_name}: 2003 - 2019"), 
+                 footnote = 'OR: odds of having received lab tests for the phenotype in the race vs other races; <br>
+                phenotypes with less than 1700 tested patients were excluded',
+                 show.footnote = TRUE,
+                     use.viewer = FALSE, 
+                     file = glue('{CACHE_DIR}/table_overly_tested_abnormalities_for_{race_name}_top10.html'))
+}
+table_overly_tested_by_race(race_name = 'Asian')
+table_overly_tested_by_race(race_name = 'Black or African American')
+table_overly_tested_by_race(race_name = 'White')
+```

### 2022-03-18 src/main/r/MSDW_HPO_Phenotype_Analysis_by_SCC_EPIC_archive.Rmd
update analysis

---
title: "LOINC2HPO_EPIC AND SCC"
output:
  html_document:
    df_print: paged
---

## GOAL
This is to LOINC2HPO analysis of both EPIC and SCC labs. 
```{r message=FALSE}
library(tidyverse)
library(glue)
library(readxl)
require(DBI)
require(RPostgres)
require(egg)
require(gridExtra)
source('.db_credential.R')
dsca <- DBI::dbConnect(RPostgres::Postgres(), host='localhost', user=user, dbname = dbname, port = 2345, password=password)
```

## ANALYSIS

### Prepare the lab tables

Run the following script to create the denormalized table for lab_scc and lab_epic. After this step, we should have three tables for SCC lab, EPIC lab, and combined. And each table should have LOINC and mapped HPO. 
```{r}
# sql_folder_path <- '/Users/xingminzhang/git/s4-pd-dmsdw/pd-dmsdw-loinc2hpo-app/src/main/resources/sql'
# sql_denormalized_scc_lab_2020july <- read_file(glue('{sql_folder_path}/loinc2hpo_SCC_MSDW.sql'))
# sql_denormalized_epic_lab_2020july <- read_file(glue('{sql_folder_path}/loinc2hpo_Epic_MSDW.sql'))
# sql_combine_Epic_SCC <- read_file(glue('{sql_folder_path}/loinc2hpo_combine_Epic_SCC.sql')) 
# 
# # make sure the lab table after LOINC2HPO transformation exists
# # if not, we can create it here; but the intended behavior is that we use the Java commandline app for this

```

```{sql connection=dsca, echo=FALSE}
drop table if exists denormalized_scc_lab_2020july;
```
```{sql connection=dsca}
CREATE TEMP TABLE denormalized_scc_lab_2020july as
WITH 
	lab_scc_2020q2 as (
		SELECT l.medical_record_number , l.encounter_key , l.calendar_date , l.time_of_day_key , l.clinical_result_numeric ,  trim(from l.unit_of_measure_numeric) as unit_of_measure_numeric , l.reference_range , l.procedure_key,
		case when l.abnormal_flag is null then 'N' else l.abnormal_flag end as abnormal_flag 
		FROM hai_az_prod.lab_scc_2020july l
		--where medical_record_number in (2966097, 3164683, 3869795)
	), 
	lab_scc_abnormal_flag_mapping AS (
		SELECT case when abf.abnormal_flag is NULL then 'N' else abf.abnormal_flag end as abnormal_flag, abf.mapto 
		FROM hai_az_prod.lab_scc_abnormal_flag_mapping abf
	)
SELECT lab.medical_record_number, lab.encounter_key , lab.calendar_date  as lab_date, dtod.clock_time_24_hour || ':00' as lab_time, lab.clinical_result_numeric , fdp.context_name, fdp.context_procedure_code AS local_test_code , loinc.loinc , lab.clinical_result_numeric AS lab_result_numeric_value, lab.unit_of_measure_numeric As unit_of_measure ,lab.reference_range AS normal_range , lab.abnormal_flag , abf.mapto AS interpretation, l2h.hpotermid , l2h.isnegated 
FROM lab_scc_2020q2 lab 
left JOIN hai_az_prod.fd_procedure fdp using (procedure_key)
left JOIN hai_az_prod.loinc_mapping loinc on fdp.context_procedure_code = loinc.code and (lab.unit_of_measure_numeric = loinc.unit or (lab.unit_of_measure_numeric is NULL and loinc.unit is NULL))
left JOIN lab_scc_abnormal_flag_mapping abf using (abnormal_flag)
left JOIN hai_az_prod.loinc2hpo l2h on loinc.loinc = l2h.loincid and abf.mapto = l2h.code
join prod_msdw.d_time_of_day dtod using (time_of_day_key);
```

```{sql connection=dsca}
DROP TABLE IF EXISTS denormalized_epic_lab_2020july;
```

```{sql connection=dsca}
create temp table denormalized_epic_lab_2020july as
with interpreted as (
	SELECT lab.mrn, lab.order_date , lab.lab_time , lab.test_code , lab.test_result_value , lab.unit_of_measurement , lab.reference_range ,
	case 
			when test_result_value_final :: REAL < range_low_final ::REAL then 'L' 
			when test_result_value_final :: REAL > range_hi_final :: REAL then 'H'
			else 'N' 
	end as interpretation
	FROM hai_az_test.lab_epic_2020july lab
	--where mrn in (3710567, 8014250, 3144637)
	)
SELECT lab.mrn AS medical_record_number, NULL AS encounter_key, lab.order_date as lab_date, lab.lab_time, 'EPIC' AS context_name, lab.test_code AS local_test_code , local2loinc.loinc , lab.test_result_value AS lab_result_numeric_value, lab.unit_of_measurement AS unit_of_measure , lab.reference_range AS normal_range , NULL AS abnormal_flag,  interpretation,
loinc2hpo.hpotermid , loinc2hpo.isnegated 
FROM interpreted lab
LEFT JOIN hai_az_prod.loinc_mapping local2loinc on lab.test_code :: VARCHAR = local2loinc.code and (lab.unit_of_measurement = local2loinc.unit OR (lab.unit_of_measurement is NULL and local2loinc.unit is NULL))
LEFT JOIN hai_az_prod.loinc2hpo loinc2hpo on local2loinc.loinc = loinc2hpo.loincid and interpretation = loinc2hpo.code 
where local2loinc."source" = 'EPIC'; 
```

Join the above tables together
```{sql connection=dsca}
CREATE VIEW denormalized_scc_and_epic_lab_2020july AS
WITH epic_denormalized AS (
  SELECT medical_record_number, lab_date, lab_time, context_name, local_test_code :: VARCHAR, loinc,  lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated
  FROM denormalized_epic_lab_2020july
), 
scc_denormalized AS (
  SELECT medical_record_number, lab_date, lab_time, context_name, local_test_code, loinc, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated
  FROM denormalized_scc_lab_2020july
)
SELECT * FROM scc_denormalized 
UNION all 
SELECT * FROM epic_denormalized; 
```

Right now, we have three sets of denormalized labs, SCC, EPIC and their union.
```{sql connection=dsca}
SELECT * FROM denormalized_scc_lab_2020july limit 4;
```

```{sql connection=dsca}
SELECT * FROM denormalized_epic_lab_2020july limit 4;
```

```{r}
lab_tables <- c('denormalized_scc_lab_2020july', 'denormalized_epic_lab_2020july', 'denormalized_scc_and_epic_lab_2020july')
```

### Summary statistics before running any HPO inference

```{r}
abnormal_phenotypes_before_inference <- function(lab_table_name){
  results <- list()
  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_before_infer_per_patient;')
  sql <- glue::glue('create temp table n_abnormal_before_infer_per_patient AS
        SELECT medical_record_number , count(*) as n_abnormal_before_infer, count(distinct hpotermid ) as n_uniq_abnormal_before_infer
        FROM {lab_table_name}
        where hpotermid is not null and not isnegated
        group by medical_record_number ;')
  dbExecute(dsca, sql)    
  # median of HPO-coded phenotypes per patient from abnormal lab findings
  median_abnormal_hpo <- dbGetQuery(dsca, 'SELECT median(n_abnormal_before_infer) FROM n_abnormal_before_infer_per_patient;')[1,1]
  results$n_abnormal_before_infer <- median_abnormal_hpo

  # median of unique HPO-coded phenotypes per patient from abnormal lab findings
  median_uniq_abnormal_before_infer <- dbGetQuery(dsca, 'SELECT median(n_uniq_abnormal_before_infer) FROM n_abnormal_before_infer_per_patient;')[1,1]
  results$n_uniq_abnormal_before_infer <- median_uniq_abnormal_before_infer

  n_abnormal_before_infer_per_patient <- dbGetQuery(dsca, 
        'SELECT n_abnormal_before_infer, count(*) as n_patient 
      FROM n_abnormal_before_infer_per_patient
      GROUP BY n_abnormal_before_infer
      order by n_abnormal_before_infer;')
  results$n_abnormal_before_infer_per_patient <- n_abnormal_before_infer_per_patient

  n_uniq_abnormal_before_infer_per_patient <- dbGetQuery(dsca, 
          'SELECT n_uniq_abnormal_before_infer, count(*) as n_patient 
            FROM n_abnormal_before_infer_per_patient
            GROUP BY n_uniq_abnormal_before_infer
            order by n_uniq_abnormal_before_infer;')
  results$n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_before_infer_per_patient

  return (results)
}

# the input is the third item in the above function
plot_patient_distribution_by_n_abnormal_phenotypes_before_infer <- function(df){
  n_abnormal_before_infer_per_patient <- df
  n_abnormal_before_infer_per_patient <- n_abnormal_before_infer_per_patient %>% mutate(n_abnormal_before_infer = as.integer(n_abnormal_before_infer), n_patient = as.integer(n_patient))

## roll up the ones > 100
  df_temp <- n_abnormal_before_infer_per_patient %>% mutate(n_abnormal_before_infer = ifelse(n_abnormal_before_infer > 100, 100, n_abnormal_before_infer)) %>% group_by(n_abnormal_before_infer) %>% summarise(n_patient = sum(n_patient))

  p <- ggplot(df_temp) + 
    geom_bar(aes(x=n_abnormal_before_infer, y = n_patient), stat='identity') +
    scale_x_continuous(limits = c(0, 105)) + 
    xlab("# HPO-coded phenotypes") + ylab("patient count") + 
    ggtitle("before HPO inference") +
    theme(axis.text = element_text(size = 8), title = element_text(size = 10))

  rm(df_temp)
  return (p)
}

plot_patient_distribution_by_n_uniq_abnormal_phenotypes_before_infer <- function(df){
  n_uniq_abnormal_before_infer_per_patient <- df
  n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_before_infer_per_patient %>% 
  mutate(n_uniq_abnormal_before_infer = as.integer(n_uniq_abnormal_before_infer), n_patient = as.integer(n_patient))

  p <- ggplot(n_uniq_abnormal_before_infer_per_patient) + 
    geom_bar(aes(x=n_uniq_abnormal_before_infer, y = n_patient), stat='identity') +
    scale_x_continuous(limits = c(0, 105)) +
    xlab("# HPO-coded phenotypes (unique)") + ylab("patient count") + 
    ggtitle("before HPO inference") +
    theme(axis.text = element_text(size = 8), title = element_text(size = 10))

  return (p)
}
```

```{r}
results_before_infer <- lapply(lab_tables, function(x) abnormal_phenotypes_before_inference(x))
```

Median of # of abnormal phenotypes before running HPO inference
```{r}
as.data.frame(list(lab_source = lab_tables, median_abnormal = sapply(results_before_infer, function(x) x$n_abnormal_before_infer), median_uniq_abnormal = sapply(results_before_infer, function(x) x$n_uniq_abnormal_before_infer)) )
```

What are most prevalent abnormal phenotypes? "Prevalent" is defined as unique number of patients having an abnormal phenotype. 
```{sql connection=dsca, max.output=-1}
SELECT l.hpotermid, hpo.label, l.n :: INTEGER
FROM 
      (SELECT hpotermid , count(distinct medical_record_number) as n
      FROM denormalized_scc_and_epic_lab_2020july
      WHERE hpotermid is not null and not isnegated
      group by hpotermid 
      order by n desc
      LIMIT 50) as l
LEFT JOIN hai_az_prod.hpo hpo on l.hpotermid = hpo.termid ;
```

### Summary statistics after running HPO inference

Count the number of HPO phenotypes per patient after inference. Note inference was only done for abnormal labs. It is tricky to infer based on normal findings.

```{r}
abnormal_phenotypes_after_inference <- function(lab_table_name){
  results <- list()
  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_with_inferred_per_patient;')
  sql <- glue::glue('create temp table n_abnormal_with_inferred_per_patient AS
with abnormal_lab as 
	(SELECT * FROM {lab_table_name} where hpotermid is not null and not isnegated),
abnormal_lab_after_infer as 
	(SELECT l.medical_record_number, l.lab_date, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance 
	FROM abnormal_lab l 
	left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair."current")
SELECT medical_record_number, count(*) as n_abnormal_with_inferred, count(DISTINCT ancestor) as n_uniq_abnormal_with_inferred
FROM abnormal_lab_after_infer 
group by medical_record_number;')
  dbExecute(dsca, sql)    
  # median of HPO-coded phenotypes per patient from abnormal lab findings
  median_abnormal_hpo <- dbGetQuery(dsca, 'SELECT median(n_abnormal_with_inferred) FROM n_abnormal_with_inferred_per_patient;')[1,1]
  results$n_abnormal_after_infer <- median_abnormal_hpo

  # median of unique HPO-coded phenotypes per patient from abnormal lab findings
  median_uniq_abnormal_after_infer <- dbGetQuery(dsca, 'SELECT median(n_uniq_abnormal_with_inferred) FROM n_abnormal_with_inferred_per_patient;')[1,1]
  results$n_uniq_abnormal_after_infer <- median_uniq_abnormal_after_infer

  n_abnormal_after_infer_per_patient <- dbGetQuery(dsca, 
        'SELECT n_abnormal_with_inferred, count(*) as n_patient 
        FROM n_abnormal_with_inferred_per_patient 
        group by n_abnormal_with_inferred 
        order by n_abnormal_with_inferred;')
  results$n_abnormal_before_infer_per_patient <- n_abnormal_after_infer_per_patient

  n_uniq_abnormal_after_infer_per_patient <- dbGetQuery(dsca, 
          'SELECT n_uniq_abnormal_with_inferred, count(*) as n_patient 
          FROM n_abnormal_with_inferred_per_patient 
          group by n_uniq_abnormal_with_inferred 
          order by n_uniq_abnormal_with_inferred;')
  results$n_uniq_abnormal_before_infer_per_patient <- n_uniq_abnormal_after_infer_per_patient

  dbExecute(dsca, 'DROP TABLE IF EXISTS n_abnormal_with_inferred_per_patient;')

  return (results)
}

# the input is the third item in the above function
plot_patient_distribution_by_n_abnormal_phenotypes_after_infer <- function(df){
  n_abnormal_with_inferred_per_patient <- df
  n_abnormal_with_inferred_per_patient <- n_abnormal_with_inferred_per_patient %>% mutate(n_abnormal_with_inferred = as.integer(n_abnormal_with_inferred), n_patient = as.integer(n_patient))
n_abnormal_with_inferred_per_patient %>% head()

  MAX_N_ABNORMAL_WITH_INFERRED <- max(n_abnormal_with_inferred_per_patient$n_abnormal_with_inferred)
  df_temp <- n_abnormal_with_inferred_per_patient %>% mutate(strata = cut(n_abnormal_with_inferred_per_patient$n_abnormal_with_inferred, breaks = c(seq(0,100, 10), seq(101, 1001, 100), seq(1002, MAX_N_ABNORMAL_WITH_INFERRED + 30000, 30000)))) %>% group_by(strata) %>% summarise(n_patient = sum(n_patient))

  p <- ggplot(df_temp) +
    geom_bar(aes(x = strata, y = n_patient), stat = 'identity') + theme(axis.text.x = element_text(angle = 30)) +
    xlab("# HPO-coded phenotypes") + ylab("patient count") + 
    ggtitle("after HPO inference") +
    theme(axis.text = element_text(size = 4), title = element_text(size = 10))

  rm(df_temp)
  return (p)
}

plot_patient_distribution_by_n_uniq_abnormal_phenotypes_after_infer <- function(df){
  n_uniq_abnormal_with_inferred_per_patient <- df
  n_uniq_abnormal_with_inferred_per_patient <- n_uniq_abnormal_with_inferred_per_patient %>% mutate(n_uniq_abnormal_with_inferred = as.integer(n_uniq_abnormal_with_inferred), n_patient = as.integer(n_patient))

  MAX_N_ABNORMAL_WITH_INFERRED <- max(n_uniq_abnormal_with_inferred_per_patient$n_uniq_abnormal_with_inferred)
  df_temp <- n_uniq_abnormal_with_inferred_per_patient %>% mutate(strata = cut(n_uniq_abnormal_with_inferred, breaks = c(seq(0,100, 10), seq(101, 1001, 100), seq(1002, MAX_N_ABNORMAL_WITH_INFERRED + 30000, 30000)))) %>% group_by(strata) %>% summarise(n_patient = sum(n_patient))

  p <- ggplot(df_temp) +
    geom_bar(aes(x = strata, y = n_patient), stat = 'identity') + theme(axis.text.x = element_text(angle = 45)) +
    xlab("# HPO-coded phenotypes (unique)") + ylab("patient count") + 
    ggtitle("after HPO inference") +
    theme(axis.text = element_text(size = 4), title = element_text(size = 10))
  rm(df_temp)

  return (p)
}
```

```{r}
results_after_infer <- lapply(lab_tables, abnormal_phenotypes_after_inference)
```

```{r}
as.data.frame(list(lab_source = lab_tables, median_abnormal = sapply(results_after_infer, function(x) x$n_abnormal_after_infer), median_uniq_abnormal = sapply(results_after_infer, function(x) x$n_uniq_abnormal_after_infer)) )
```

### Show patient distribution plots before and after inference on one plot

```{r}
# input is the list item in results_before_infer, and results_after_infer
plot_grid <- function(result_before_infer, result_after_infer){
  # patient distribution ~ abnormal phenotypes before inference
  df_abnormal_before <- result_before_infer[[3]]
  p1 <- plot_patient_distribution_by_n_abnormal_phenotypes_before_infer(df_abnormal_before)
  # patient distribution ~ uniq abnormal phenotypes before inference
  df_uniq_abnormal_before <- result_before_infer[[4]]
  p2 <- plot_patient_distribution_by_n_uniq_abnormal_phenotypes_before_infer(df_uniq_abnormal_before)
  # patient distribution ~ abnormal phenotypes after inference
  df_abnormal_after <- result_after_infer[[3]]
  p3 <- plot_patient_distribution_by_n_abnormal_phenotypes_after_infer(df_abnormal_after)
  # patient distribution ~ uniq abnormal phenotypes after inference
  df_uniq_abnormal_after <- result_after_infer[[4]]
  p4 <- plot_patient_distribution_by_n_uniq_abnormal_phenotypes_after_infer(df_uniq_abnormal_after)

  width = unit(8, "cm")
  height = unit(3, "cm")
  p <- gridExtra::grid.arrange(
    set_panel_size(p1, width = width, height = height),
    set_panel_size(p2, width = width, height = height),
    set_panel_size(p3, width = width, height = height),
    set_panel_size(p4, width = width, height = height),
    ncol=2)

  return (p)
}
```

Patient distribution by looking at SCC labs alone
```{r}
# Lab_SCC
plot_grid(results_before_infer[[1]], results_after_infer[[1]])
```

Patient distribution by looking at EPIC labs alone
```{r}
# Lab_EPIC
plot_grid(results_before_infer[[2]], results_after_infer[[2]])
```

Patient distribution by looking at all labs combined
```{r}
# Lab_SCC_and_EPIC
plot_grid(results_before_infer[[3]], results_after_infer[[3]])
```

We can also look at the most prevalent phenotypes after inference. It is not fair to compare all labs because parent terms are always going to be more prevalent. So we just look at level 1 terms (terms right below "Phenotypic Abnormality").
Look at patients at level = 1 (right below phenotypic abnormality)
```{sql connection=dsca, max.output=-1}
with abnormal_lab as 
	(SELECT * FROM denormalized_scc_and_epic_lab_2020july where hpotermid is not null and not isnegated),
abnormal_lab_after_infer as 
	(SELECT l.medical_record_number, l.lab_date, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance , hpo.distancetoroot, hpo.label
	FROM abnormal_lab l 
	left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair."current"
	left join hai_az_prod.hpo hpo on pair.ancestor = hpo.termid)
SELECT ancestor, label, count(distinct medical_record_number) :: INTEGER as n_patient FROM abnormal_lab_after_infer where distancetoroot = 1 group by ancestor, label
ORDER BY n_patient DESC;
```

```{sql connection=dsca}
DROP VIEW IF EXISTS denormalized_scc_and_epic_lab_2020july;
```

### 2022-03-22 src/main/r/biomarker.Rmd
update analysis

@@
 ## explore other statistical models
-model2 <- glmer(formula = 'case_or_control ~ gender + race + age + is_observed + (1 + is_observed | is_tested)', data = df_model, family = binomial(link = 'logit') )
-summary(model2)
+# model2 <- glmer(formula = 'case_or_control ~ gender + race + age + is_observed + (1 + is_observed | is_tested)', data = df_model, family = binomial(link = 'logit') )
+# summary(model2)
 ```
@@
 
-plan(multisession, workers = 6)
+plan(multisession, workers = 8)
 
-screen_results2 <- phenotype_features %>% 
+screen_results <- phenotype_features %>% 
   #filter(termid == 'HP:0001880') %>%

### 2022-03-22 src/main/r/numeric_vs_hpo_combined.Rmd
update analysis

---
title: "numeric vs HPO"
output: html_notebook
---

In this page, we combine the analysis of numeric value vs HPO terms for prediction tasks.

```{r}
library(tidyverse)
library(glue)
```

Load previous data analysis results
```{r}
crc_analysis <- read.csv('cache/crc_case_matched_controls_analysis.csv')
acute_kidney_failure_analysis <- read.csv('cache/acute_kidney_failure_analysis.csv')
anemia_analysis <- read.csv('cache/anemia_analysis.csv')
```

```{r}
numeric_vs_hpo <- bind_rows(crc_analysis %>% mutate(disease = 'colon cancer'), 
                            acute_kidney_failure_analysis %>% mutate(disease = 'acute kidney failure'), 
                            anemia_analysis %>% mutate(disease = 'aplastic anemia'))

numeric_vs_hpo %>% sample_n(4)

numeric_vs_hpo_final <- numeric_vs_hpo %>% 
  select(disease, loinc, estimate1, estimate2, p.value, interpretation, hpotermid) %>% 
  rename(auc_continuous = estimate1, auc_hpo = estimate2) %>% 
  group_by(disease, loinc) %>% 
  arrange(desc(auc_hpo)) %>% 
  mutate(seq = row_number()) %>% 
  ungroup() %>% 
  arrange(disease, loinc, seq) %>% 
  filter(seq == 1) %>% mutate(frac = auc_hpo/auc_continuous) %>% 
  mutate(frac_rm_rand = (auc_hpo - 0.5) / (auc_continuous - 0.5))

numeric_vs_hpo_final
```
```{r}
numeric_vs_hpo_final  %>% 
  #filter(estimate1 > 0.6) %>%
  summarise(frac_median = median(frac), frac_rm_rand_median = median(frac_rm_rand))
```
```{r}
df <- numeric_vs_hpo_final %>% 
  select(disease, loinc, auc_continuous, auc_hpo) %>% 
  pivot_longer(cols = starts_with('auc'), names_to = 'method', names_prefix = 'auc_', values_to = 'auc')

plot_AUC_differences_numeric_vs_hpo <- ggplot(df, aes(x = method, y = auc, group = interaction(disease, loinc))) + 
  geom_point(size = 0.5) +
  geom_line(aes(color = disease)) +
  scale_x_discrete(breaks = c('continuous', 'hpo'), labels = c('original numeric value', 'transformed to HPO')) +
  xlab("Use Lab Test Results As") +
  ylab("AUC") +
  theme_bw() + 
    theme(panel.grid = element_blank(),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 8), 
          strip.text.x = element_text(margin = margin(0.3, 0, 0.3, 0), size = 5),
          plot.title = element_text(size = 8))

ggsave(filename = 'cache/plot_AUC_differences_numeric_vs_hpo.pdf', width = 5, height = 3)

plot_AUC_differences_numeric_vs_hpo
```

### 2022-03-22 src/main/r/numeric_vs_hpo_for_CRC.Rmd
update analysis

@@
 }
+FORCE_RERUN <- TRUE
 ```
@@
 sql <- "SELECT * FROM cohort"
-df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'colorectal_cancer_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'colorectal_cancer_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
 
@@
 sql <- read_file('../sql/get_labs_for_colorectal_cancer.sql')
-df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_colorectal_cancer.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
+df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_colorectal_cancer.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FORCE_RERUN)
 df_labs %>% nrow()
@@
 
-p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '6690-2', ylim = c(0, 50)) + ggtitle('Leukocytes')
+p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '26464-8', ylim = c(0, 50)) + ggtitle('Leukocytes')
 
@@
 require(furrr)
-plan(multisession, workers = 4)
+plan(multisession, workers = 5)
 
@@
   nest() %>%
+  ungroup() %>%
   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
@@
 sql <- read_file('../sql/colorectal_cancer_cohort_demographics.sql')
-df_case_demographics <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'demographics_colorectal_cancer_cases.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_case_demographics <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'demographics_colorectal_cancer_cases.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
 
@@
 sql <- 'SELECT * FROM control_cohort'
-df_control_demographics <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'demographics_colorectal_cancer_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_control_demographics <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'demographics_colorectal_cancer_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
 
 sql <- read_file('../sql/get_labs_for_colorectal_cancer_control.sql')
-df_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_sampled_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_sampled_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
 ```
@@
   as_tibble() %>%
+  mutate(medical_record_number = as.character(medical_record_number)) %>% 
+  mutate(local_test_code = as.character(local_test_code)) %>% 
   #filter(medical_record_number == '7007952')  %>%
@@
 
+crc_case_matched_controls_analysis %>% write.csv(file = 'cache/crc_case_matched_controls_analysis.csv', row.names = FALSE)
+
 crc_case_matched_controls_analysis
+
 ```
@@
   nest() %>%
+  ungroup() %>% 
   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, include_demographics=TRUE, .options = furrr_options(seed = T))) %>%

### 2022-04-12 src/main/r/biomarker.Rmd
update analysis

@@
   model_coeff <- data.frame(term = character(), estimate = double(), std.error = double(), statistic = double(), p.value = double())
-  tryCatch({model_coeff = glm_Xy(df_model)}, error = function(e) print('cannot fit model'))
+  tryCatch({model_coeff = glm_Xy(df_model)}, error = function(e) warning('cannot fit model'))
   #error = function(e) print('model did not run'))
@@
 ## at each time window
+N_BIOMARKERS_SCREENED <- screen_results %>% select(lab_date_rel_bracket, termid) %>% distinct() %>% nrow()
 screen_results %>% group_by(lab_date_rel_bracket) %>% summarise(n = n_distinct(termid))
@@
   left_join(hpo_term_list, by = 'termid') %>%
-  select(lab_date_rel_bracket, termid, label, term, estimate, std.error, statistic, p.value)
+  select(lab_date_rel_bracket, termid, label, distance_to_root, term, distance_to_root, estimate, std.error, statistic, p.value)
 
@@
 ```{r}
-# rank results by weight of is_tested
-screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
-  arrange(term) %>%
-  mutate(term_rank = row_number()) %>%
-  ungroup() %>% 
-  arrange(lab_date_rel_bracket, termid) %>% 
-  group_by(lab_date_rel_bracket, termid, label) %>%
-  nest() %>% 
-  mutate(ranking_metric_value = map_dbl(data, function(df) df$estimate[df$term == 'is_testedTRUE'])) %>% 
-  arrange(-ranking_metric_value) %>% 
-  mutate(overall_rank = row_number()) %>%
-  unnest(cols = c(data)) %>%
-  write.csv(file = '../cache/screen_results_annotated_rank_by_weight_of_is_tested.csv', row.names = FALSE)
+screen_results_annotated_formatted <- screen_results_annotated %>% 
+  mutate(OR = exp(estimate)) %>% 
+  mutate(OR = round(OR, 2)) %>% 
+  #mutate(p.value = scales::scientific(p.value, digits = 3)) %>% 
+  select(-estimate, -std.error, -statistic) %>% 
+  mutate(term = case_when(term == 'is_testedTRUE' ~ 'tested',
+                          term == 'is_testedTRUE:is_observedTRUE' ~ 'observed',
+                          TRUE ~ 'NEVER GONNA HAPPEN')) %>% 
+  pivot_wider(id_cols = c(lab_date_rel_bracket, termid, label, distance_to_root), names_from = term, values_from = c(OR, p.value)) %>% 
+  relocate(p.value_tested, .after = 'OR_tested') %>% 
+  rename(`Odds Ratio (Been Tested)` = OR_tested) %>% 
+  rename(`p value (Been Tested)` = p.value_tested) %>% 
+  rename(`Odds Ratio (Been Observed)` = 'OR_observed') %>% 
+  rename(`p value (Been Observed)` = 'p.value_observed')
 
-# rank results by weight of is_testedTRUE:is_observedTRUE
-screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
-  arrange(term) %>%
-  mutate(term_rank = row_number()) %>%
-  ungroup() %>% 
-  arrange(lab_date_rel_bracket, termid) %>% 
-  group_by(lab_date_rel_bracket, termid, label) %>%
-  nest() %>% 
-  mutate(ranking_metric_value = map_dbl(data, function(df) df$estimate[df$term == 'is_testedTRUE:is_observedTRUE'])) %>% 
-  arrange(-ranking_metric_value) %>% 
-  mutate(overall_rank = row_number()) %>%
-  unnest(cols = c(data)) %>%
-  write.csv(file = '../cache/screen_results_annotated_rank_by_weight_of_is_tested_is_observed.csv', row.names = FALSE)
+write.csv(screen_results_annotated_formatted, file = '../cache/screen_results_annotated.csv', row.names = FALSE)
 
-# rank results by p value of is_tested
-screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
-  arrange(term) %>%
-  mutate(term_rank = row_number()) %>%
-  ungroup() %>% 
-  arrange(lab_date_rel_bracket, termid) %>% 
-  group_by(lab_date_rel_bracket, termid, label) %>%
-  nest() %>% 
-  mutate(ranking_metric_value = map_dbl(data, function(df) df$p.value[df$term == 'is_testedTRUE'])) %>% 
-  arrange(ranking_metric_value) %>% 
-  mutate(overall_rank = row_number()) %>%
-  unnest(cols = c(data)) %>%
-  write.csv(file = '../cache/screen_results_annotated_rank_by_p_of_is_tested.csv', row.names = FALSE)
+screen_results_annotated_formatted %>% head()
+```
 
-# rank results by p value of is_testedTRUE:is_observedTRUE
-screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
-  arrange(term) %>%
-  mutate(term_rank = row_number()) %>%
-  ungroup() %>% 
-  arrange(lab_date_rel_bracket, termid) %>% 
-  group_by(lab_date_rel_bracket, termid, label) %>%
-  nest() %>% 
-  mutate(ranking_metric_value = map_dbl(data, function(df) df$p.value[df$term == 'is_testedTRUE:is_observedTRUE'])) %>% 
-  arrange(ranking_metric_value) %>% 
-  mutate(overall_rank = row_number()) %>%
-  unnest(cols = c(data)) %>%
-  write.csv(file = '../cache/screen_results_annotated_rank_by_p_of_is_tested_is_observed.csv', row.names = FALSE)
+Highlight some biomarkers:
+
+Eosinophilia
+```{r}
+eosinophilia <- screen_results_annotated_formatted %>% 
+  filter(label == 'Eosinophilia') %>% 
+  select(-distance_to_root) %>% 
+  mutate(across(starts_with('p value'), scales::scientific))
+
+sjPlot::tab_df(eosinophilia %>% rename(`Lab Year Bracket` = lab_date_rel_bracket), 
+                title = 'Table Association of Eosinophilia with Severe Asthma',
+               file = '../cache/table_eosinophilia_with_severe_asthma.html',
+               alternate.rows = TRUE)
+
+eosinophilia
+```
+
+Rank by Odds Ratio (Been Observed)
+```{r}
+significant_by_tested_and_observed <- screen_results_annotated_formatted %>% 
+  arrange(desc(`Odds Ratio (Been Observed)`)) %>% 
+  filter(`p value (Been Tested)` < 0.05/N_BIOMARKERS_SCREENED) %>%
+  filter(`p value (Been Observed)` < 0.05/N_BIOMARKERS_SCREENED) %>%
+  filter(`Odds Ratio (Been Tested)` > 1) %>% 
+  filter(`Odds Ratio (Been Observed)` > 1) %>% 
+  filter(lab_date_rel_bracket == '(-1,0]') %>%
+  mutate(across(starts_with('p value'), scales::scientific))
+
+significant_by_tested_and_observed_deduped <- significant_by_tested_and_observed %>% 
+  rows_delete(tibble(label = c('Abnormal eosinophil morphology', 
+                               'Abnormal leukocyte morphology',
+                               'Abnormal immune system morphology',
+                               'Abnormal cellular immune system morphology',
+                               'Abnormal leukocyte count',
+                               'Abnormal granulocyte morphology',
+                               'Abnormality of the respiratory system',
+                               'Functional respiratory abnormality')), by = 'label')
+
+significant_by_tested_and_observed_deduped
+```
+```{r}
+significant_by_tested_and_observed_deduped_three_years <- screen_results_annotated_formatted %>% 
+  filter(termid %in% significant_by_tested_and_observed_deduped$termid) %>% 
+  mutate(termid = factor(termid, levels = significant_by_tested_and_observed_deduped$termid)) %>% 
+  arrange(termid, lab_date_rel_bracket) %>%
+  mutate(across(starts_with('p value'), scales::scientific))
+
+significant_by_tested_and_observed_deduped_three_years
+```
+
+```{r}
+sjPlot::tab_df(significant_by_tested_and_observed_deduped %>% select(-distance_to_root, -lab_date_rel_bracket), 
+                title = 'Table Selected Phenotypes (within 1 year prior) Associated with Severe Asthma',
+               file = '../cache/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html', 
+               alternate.rows = TRUE)
+sjPlot::tab_df(significant_by_tested_and_observed_deduped_three_years %>% 
+                 select(-distance_to_root) %>% rename(`Lab Year Bracket` = lab_date_rel_bracket), 
+                title = 'Table Selected Phenotypes (within past 3 years) Associated with Severe Asthma',
+               file = '../cache/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html',
+               alternate.rows = TRUE)
 ```

### 2022-04-12 src/main/r/numeric_vs_hpo_for_liver_function.Rmd
update analysis

@@
 }
+FORCE_RERUN <- FALSE
 ```
@@
 sql <- "SELECT * FROM cohort"
-df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'liver_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'liver_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
 
@@
 sql <- read_file('../sql/get_labs_for_liver_function.sql')
-df_liver_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_liver_function.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
+df_liver_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_liver_function.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FORCE_RERUN)
 df_liver_labs %>% nrow()
@@
 require(furrr)
-plan(multisession, workers = 4)
+plan(multisession, workers = 6)
 
@@
   nest() %>%
+  ungroup() %>% 
   mutate(result = future_map(data, pipeline_compare_for_liver_function)) %>%
@@
   unnest(result)
+
+write.csv(liver_analysis, file = 'cache/liver_analysis.csv', row.names = FALSE)
 
@@
 roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
-roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'H'))
-roc_comparisons$H <- roc_comparison
 

### 2022-04-21 code/r/MSDW_HPO_Phenotype_Analysis.Rmd
reorg project structure

--- 2022-03-18 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
@@
 plot_by_distinct_abnormal_hpo_infer_per_patient
+```
+
+### Visualize patient phenotypic journey
+```{r}
+get_patient_lab_phenotype_journeys <- function(medical_record_numbers, force_rerun=FALSE, cache_dir=CACHE_DIR) {
+  df_path = glue('{cache_dir}/patient_lab_phenotype_journey.csv')
+
+  mrns_str <- str_c("'", medical_record_numbers, "'", collapse = ',')
+
+  if (force_rerun | !file.exists(df_path)){
+    df <- dbGetQuery(dsca, glue("
+            with patient_lab_results as (
+            	select *, date_part(year, lab_date) as lab_year
+            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july saelalhj 
+            	where medical_record_number in ({mrns_str}) and hpotermid is not null 
+            ),
+            annual_phenotypes as (
+            	select medical_record_number , lab_year, isnegated , hpotermid , count(*) as n 
+            	from patient_lab_results 
+            	group by medical_record_number , lab_year, isnegated , hpotermid
+            ),
+            annual_phenotypes_annotated as (
+            	select *
+            	from annual_phenotypes p
+            	left join hai_az_prod.hpo h on p.hpotermid  = h.termid  
+            	order by medical_record_number , lab_year, isnegated , hpotermid )
+            select medical_record_number , lab_year, isnegated , hpotermid , n, label
+            from annual_phenotypes_annotated
+            order by medical_record_number , lab_year, isnegated , n desc, hpotermid
+            ;
+            "))
+
+    write.csv(df, file = df_path, row.names=FALSE)
+  }
+
+  df <- read.csv(df_path)
+
+  return (df)
+}
+
+patient_lab_phenotype_journeys <- get_patient_lab_phenotype_journeys(medical_record_numbers = c('7611233'), force_rerun = FORCE_RERUN)
+```
+
+```{r}
+patient_lab_phenotype_journeys_formated <- patient_lab_phenotype_journeys %>% 
+  filter(medical_record_number == '7611233') %>% 
+  select(-medical_record_number) %>% 
+  group_by(lab_year, isnegated) %>% 
+  slice_head(n = 5) %>% 
+  mutate(display = glue('{hpotermid} {label} ({n} times)')) %>% 
+  summarise(display = str_c(display, collapse = '\n'), .groups = 'drop')
+write.csv(patient_lab_phenotype_journeys_formated, file = glue('{CACHE_DIR}/patient_lab_phenotype_journey_formatted.csv'))
 ```

### 2022-04-21 code/r/classifier_severe_asthma.Rmd
reorg project structure

---
title: "classifier_severe_asthma"
author: "Aaron Zhang"
date: "3/30/2022"
output: html_document
---

The goal of this document is to build ML classifiers to predict progression into severe asthma using biomarkers idenfied from lab phenotype screening.

```{r message=FALSE}
library(tidyverse)
library(glue)
```

## Load data
```{r}
CACHE_DIR <- '../cache'

confounders_outcome <- read_csv(glue("{CACHE_DIR}/confounders_outcome.csv"))
phenotype_features <- read_csv(glue("{CACHE_DIR}/phenotype_features.csv"))
```

## Prepare feature matrix

We will use the biomarker in the past year, and the count in the past three
```{r}
biomarkers_identified <- c("HP:0001880", "HP:0001899", "HP:0012312", 
                           "HP:0032309", "HP:0011897", "HP:0012416")

# biomarkers from Year (-1, 0] only
biomarker_feature_matrix <- phenotype_features %>% filter(termid %in% biomarkers_identified) %>% 
  mutate(termid = make.names(termid)) %>% 
  rename(tested = tested_count) %>% 
  rename(observed = observed_count) %>% 
  select(medical_record_number, lab_date_rel_bracket, termid, tested, observed) %>% 
  #filter(lab_date_rel_bracket == '(-1,0]') %>% 
  pivot_wider(id_cols = medical_record_number, names_from = c(lab_date_rel_bracket, termid), values_from = c(tested, observed))

Xy <- confounders_outcome %>% 
  left_join(biomarker_feature_matrix, by = 'medical_record_number') %>% 
  mutate(across(starts_with('tested'), ~ replace_na(.x, 0))) %>% 
  mutate(across(starts_with('observed'), ~ replace_na(.x, 0))) %>% 
  mutate(gender = as_factor(gender)) %>%
  mutate(gender = fct_relevel(gender, 'Male')) %>% 
  mutate(race = as_factor(race)) %>% 
  mutate(race = fct_relevel(race, 'White')) %>% 
  mutate(age = round(age / 10, 0) * 10) %>% 
  relocate(case_or_control, .after = last_col())
```

```{r message=FALSE}
library(caret)
require(doSNOW)
require(pROC)
```

```{r}
set.seed(7531)

# how to balance classes
class_balance_method <- c('downsample' = TRUE, 'upsample' = FALSE, 'asis' = FALSE)
stopifnot(sum(class_balance_method) == 1)
if (class_balance_method['downsample']){
  Xy <- downSample(Xy %>% select(-case_or_control), factor(Xy$case_or_control), yname='case_or_control')
} else if (class_balance_method['upsample']){
  Xy <- upSample(Xy %>% select(-case_or_control), factor(Xy$case_or_control), yname='case_or_control')
} else {
  # do nothing
}
```

```{r}
trainIndex <- createDataPartition(Xy$case_or_control, p = 0.80, list = FALSE)

df_ml_train <- Xy[trainIndex, ]
df_ml_test <- Xy[-trainIndex, ]
```

```{r}
# train a random forest, start with default
control <- trainControl(method='repeatedcv', 
                        number=5, 
                        repeats=3,
                        classProbs = TRUE)
mtry <- sqrt(ncol(df_ml_train))
tunegrid <- expand.grid(.mtry=mtry)

c1 <- makeCluster(6, type = "SOCK")
registerDoSNOW(c1)
rf_default <- train(case_or_control ~ ., 
                    data = df_ml_train %>% select(-medical_record_number) %>% mutate(case_or_control = factor(case_or_control)),
                    method = 'rf', 
                    metric = 'Kappa', 
                    truneGrid = tunegrid,
                    trControl = control)
stopCluster(c1)

importance <- varImp(rf_default)

feature_rank_rf <- as.data.frame(importance$importance) %>% rownames_to_column(var = 'feature') %>% arrange(-Overall)

prediction <- predict(rf_default, df_ml_test %>% select(-medical_record_number, -case_or_control))

confusionMatrix(prediction, factor(df_ml_test$case_or_control), positive = 'case')
```

```{r}
# use a boosting algorithm
c1 <- makeCluster(6, type = "SOCK")
registerDoSNOW(c1)
xgbGrid <- expand.grid(nrounds = c(30, 60, 90),
                       max_depth = c(10, 15, 20, 25),
                       colsample_bytree = seq(0.5, 0.9, length.out = 5),
                       eta = c(0.1, 0.2, 0.4),
                       gamma=0,
                       min_child_weight = 1,
                       subsample = c(0.5, 0.75)
                      )
xgboost_model <- train(case_or_control ~ ., 
                    data = df_ml_train %>% select(-medical_record_number) %>% mutate(case_or_control = factor(case_or_control)),
                    method = 'xgbTree', 
                    tuneGrid = xgbGrid,
                    trControl = control)
stopCluster(c1)
prediction <- predict(xgboost_model, df_ml_test %>% select(-medical_record_number, -case_or_control))

confusionMatrix(prediction, factor(df_ml_test$case_or_control), positive = 'case')
```

### 2022-04-22 code/r/MSDW_HPO_Phenotype_Analysis.Rmd
refactor

@@
 require(gridExtra)
-source('.db_credential.R')
-dsca <- DBI::dbConnect(RPostgres::Postgres(), host='localhost', user=user, dbname = dbname, port = 2345, password=password)
+dsca <- DBI::dbConnect(RPostgres::Postgres(), 
+                       host=Sys.getenv('DSCA_PHI_HOST'), 
+                       dbname = Sys.getenv('DSCA_PHI_DBNAME'), 
+                       port = Sys.getenv('DSCA_PHI_PORT'), 
+                       user=Sys.getenv('DSCA_PHI_USERNAME'),
+                       password= Sys.getenv('DSCA_PHI_PASSWORD'))
 ```
@@
 ```{r}
-CACHE_DIR = 'cache'
+CACHE_DIR = '../../cache'
 if (!dir.exists(CACHE_DIR)){
@@
 
+# if true, ignore cached data and reconnect to database to get latest data;
+# otherwise, use cached data first before trying to connect to database
 FORCE_RERUN <- FALSE
@@
 
-patient_race_sex_distribution <- get_patient_race_sex_distribution(force_rerun = FALSE)
+patient_race_sex_distribution <- get_patient_race_sex_distribution(force_rerun = FORCE_RERUN)
 
@@
 ```{r}
-tested_phenotype_rank_direct %>% head(10) %>% formattable::formattable()
+tested_phenotype_rank_direct %>% head(10)
 ```

### 2022-04-22 code/r/02_B_numeric_vs_hpo_for_acute_kidney_failure.Rmd
rename files

--- 2021-12-01 src/main/r/numeric_vs_hpo_for_acute_kidney_failure.Rmd
@@
 sql <- read_file('../sql/get_labs_for_acute_kidney_failure.sql')
-df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_acute_kidney_failure.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
+df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_acute_kidney_failure.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FORCE_RERUN)
 df_labs %>% nrow()
@@
   nest() %>%
+  ungroup() %>% 
   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
@@
   unnest(result)
+
+write.csv(kidney_analysis, file = 'cache/acute_kidney_failure_analysis.csv', row.names = FALSE)
 

### 2022-04-22 code/r/02_C_numeric_vs_hpo_for_blood_disorder.Rmd
rename files

--- 2021-12-10 src/main/r/numeric_vs_hpo_for_blood_disorder.Rmd
@@
 }
+
+FORCE_RERUN <- TRUE
 ```
@@
 sql <- "SELECT * FROM cohort"
-df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'aplastic_anemia_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'aplastic_anemia_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
 
@@
 sql <- read_file('../sql/get_labs_for_aplastic_anemia.sql')
-df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_aplastic_anemia.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
+df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_aplastic_anemia.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FORCE_RERUN)
 df_labs %>% nrow()
@@
   unnest(result)
+
+write.csv(anemia_analysis, file = 'cache/anemia_analysis.csv', row.names = FALSE)
 

### 2022-04-22 code/r/02_E_numeric_vs_hpo_combined.Rmd
rename files

--- 2022-03-22 src/main/r/numeric_vs_hpo_combined.Rmd
@@
 ```{r}
+liver_analysis <- read.csv('cache/liver_analysis.csv')
 crc_analysis <- read.csv('cache/crc_case_matched_controls_analysis.csv')
@@
 ```{r}
-numeric_vs_hpo <- bind_rows(crc_analysis %>% mutate(disease = 'colon cancer'), 
+numeric_vs_hpo <- bind_rows(liver_analysis %>% mutate(disease = 'abnormal liver function'),
+                            crc_analysis %>% mutate(disease = 'colon cancer'), 
                             acute_kidney_failure_analysis %>% mutate(disease = 'acute kidney failure'), 
@@
 plot_AUC_differences_numeric_vs_hpo <- ggplot(df, aes(x = method, y = auc, group = interaction(disease, loinc))) + 
-  geom_point(size = 0.5) +
+  geom_point(size = 1) +
   geom_line(aes(color = disease)) +
@@
   theme_bw() + 
-    theme(panel.grid = element_blank(),
+    theme(panel.grid = element_blank(), legend.position = 'none',
           axis.text = element_text(size = 8),
@@
           strip.text.x = element_text(margin = margin(0.3, 0, 0.3, 0), size = 5),
-          plot.title = element_text(size = 8))
+          plot.title = element_text(size = 8)) +
+  facet_wrap(~ disease)
 
-ggsave(filename = 'cache/plot_AUC_differences_numeric_vs_hpo.pdf', width = 5, height = 3)
+ggsave(filename = 'cache/plot_AUC_differences_numeric_vs_hpo.pdf', width = 6, height = 6)
 

### 2022-04-22 code/r/02_C_numeric_vs_hpo_for_blood_disorder.Rmd
update

@@
 require(gridExtra)
-source('.db_credential.R')
 source('functions.R')
-dsca <- DBI::dbConnect(RPostgres::Postgres(), host='localhost', user=user, dbname = dbname, port = 2345, password=password)
+dsca <- DBI::dbConnect(RPostgres::Postgres(), 
+                       host=Sys.getenv('DSCA_PHI_HOST'), 
+                       dbname = Sys.getenv('DSCA_PHI_DBNAME'), 
+                       port = Sys.getenv('DSCA_PHI_PORT'), 
+                       user=Sys.getenv('DSCA_PHI_USERNAME'),
+                       password= Sys.getenv('DSCA_PHI_PASSWORD'))
 
 # cache directory to read from or write to
-CACHE_DIR <- './cache'
+CACHE_DIR = '../../cache'
+
 if (!dir.exists(CACHE_DIR)){
@@
 
-FORCE_RERUN <- TRUE
+FORCE_RERUN <- FALSE
 ```

### 2022-04-22 code/r/02_E_numeric_vs_hpo_combined.Rmd
update

@@
 library(glue)
+
+# cache directory to read from or write to
+CACHE_DIR = '../../cache'
 ```
@@
 ```{r}
-liver_analysis <- read.csv('cache/liver_analysis.csv')
-crc_analysis <- read.csv('cache/crc_case_matched_controls_analysis.csv')
-acute_kidney_failure_analysis <- read.csv('cache/acute_kidney_failure_analysis.csv')
-anemia_analysis <- read.csv('cache/anemia_analysis.csv')
+liver_analysis <- read.csv(glue('{CACHE_DIR}/liver_analysis.csv'))
+crc_analysis <- read.csv(glue('{CACHE_DIR}/crc_case_matched_controls_analysis.csv'))
+acute_kidney_failure_analysis <- read.csv(glue('{CACHE_DIR}/acute_kidney_failure_analysis.csv'))
+anemia_analysis <- read.csv(glue('{CACHE_DIR}/anemia_analysis.csv'))
 ```
@@
 
-ggsave(filename = 'cache/plot_AUC_differences_numeric_vs_hpo.pdf', width = 6, height = 6)
+ggsave(filename = glue('{CACHE_DIR}/plot_AUC_differences_numeric_vs_hpo.pdf'), width = 6, height = 6)
 

### 2022-04-22 code/r/03_biomarker.Rmd
update

--- 2022-04-12 src/main/r/biomarker.Rmd
@@
 require(furrr)
-source('.db_credential.R')
-dsca <- DBI::dbConnect(RPostgres::Postgres(), host='localhost', user=user, dbname = dbname, port = 2345, password=password)
+require(dbplyr)
+
+dsca <- DBI::dbConnect(RPostgres::Postgres(), 
+                       host=Sys.getenv('DSCA_PHI_HOST'), 
+                       dbname = Sys.getenv('DSCA_PHI_DBNAME'), 
+                       port = Sys.getenv('DSCA_PHI_PORT'), 
+                       user=Sys.getenv('DSCA_PHI_USERNAME'),
+                       password= Sys.getenv('DSCA_PHI_PASSWORD'))
+
+# cache directory to read from or write to
+CACHE_DIR = '../../cache'
+
+if (!dir.exists(CACHE_DIR)){
+  dir.create(CACHE_DIR)
+}
+FORCE_RERUN <- TRUE
 ```
@@
 
-```{sql connection=dsca, output.var=cohort}
-select * from cohort;
-```
-
-```{sql connection=dsca, output.var=tested_phenotypes}
+Get data from the database. The following method are lazily executed.
+```{r}
+cohort <- tbl(dsca, 'cohort')
+tested_phenotypes <- tbl(dsca, sql("
     with testable_phenotypes_all as (
@@
      select medical_record_number, lab_date, termid
- from testable_phenotypes_all;
-```
-
-```{sql connection=dsca, output.var=observed_phenotypes}
+     from testable_phenotypes_all"))
+observed_phenotypes <- tbl(dsca, sql("
     with abnormal_lab as (
@@
     select medical_record_number , lab_date, ancestor as termid
-from abnormal_lab_after_infer;
-```
-
-```{sql connection=dsca, output.var='cohort_demographics'}
+    from abnormal_lab_after_infer"))
+cohort_demographics <- tbl(dsca, sql("
     select person.*
     from hai_az_prod.person_msdw_2020july person
-join cohort using (medical_record_number);
+    join cohort using (medical_record_number)"))
+
 ```
@@
 ```{r}
-hpo_term_list <- read.csv('/Users/xingminzhang/git/s4-pd-dmsdw/pd-dmsdw-loinc2hpo-app/src/main/resources/hp_term_list.csv', header = FALSE, col.names = c('termid', 'distance_to_root', 'label'), stringsAsFactors = FALSE)
+hpo_term_list <- read.csv('../../in/hp_term_list.csv', header = FALSE, col.names = c('termid', 'distance_to_root', 'label'), stringsAsFactors = FALSE)
 
@@
 
-write.csv(screen_results_annotated, file = '../cache/screen_results_annotated.csv', row.names = FALSE)
+write.csv(screen_results_annotated, file = glue('{CACHE_DIR}/screen_results_annotated.csv'), row.names = FALSE)
 
@@
 
-write.csv(screen_results_annotated_formatted, file = '../cache/screen_results_annotated.csv', row.names = FALSE)
+write.csv(screen_results_annotated_formatted, file = glue('{CACHE_DIR}/screen_results_annotated.csv', row.names = FALSE)
 
@@
                 title = 'Table Association of Eosinophilia with Severe Asthma',
-               file = '../cache/table_eosinophilia_with_severe_asthma.html',
+               file = glue('{CACHE_DIR}/table_eosinophilia_with_severe_asthma.html',
                alternate.rows = TRUE)
@@
                 title = 'Table Selected Phenotypes (within 1 year prior) Associated with Severe Asthma',
-               file = '../cache/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html', 
+               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html', 
                alternate.rows = TRUE)
@@
                 title = 'Table Selected Phenotypes (within past 3 years) Associated with Severe Asthma',
-               file = '../cache/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html',
+               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html',
                alternate.rows = TRUE)
@@
     filename <- names(cache_list)[[i]]
-    write.table(cache_list[[i]], file = glue('../cache/{filename}.csv'), sep = ',', row.names = FALSE, col.names = TRUE)
+    write.table(cache_list[[i]], file = glue('{CACHE_DIR}/{filename}.csv'), sep = ',', row.names = FALSE, col.names = TRUE)
   }

### 2022-04-25 code/r/02_A_numeric_vs_hpo_for_liver_function.Rmd

--- 2022-04-12 src/main/r/numeric_vs_hpo_for_liver_function.Rmd
@@
 require(gridExtra)
-source('.db_credential.R')
 source('functions.R')
-dsca <- DBI::dbConnect(RPostgres::Postgres(), host='localhost', user=user, dbname = dbname, port = 2345, password=password)
+dsca <- DBI::dbConnect(RPostgres::Postgres(), 
+                       host=Sys.getenv('DSCA_PHI_HOST'), 
+                       dbname = Sys.getenv('DSCA_PHI_DBNAME'), 
+                       port = Sys.getenv('DSCA_PHI_PORT'), 
+                       user=Sys.getenv('DSCA_PHI_USERNAME'),
+                       password= Sys.getenv('DSCA_PHI_PASSWORD'))
 
 # cache directory to read from or write to
-CACHE_DIR <- './cache'
+CACHE_DIR = '../../cache'
 if (!dir.exists(CACHE_DIR)){
@@
 
-ggsave(filename = 'cache/liver_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
+ggsave(filename = glue('{CACHE_DIR}/liver_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6))
 ```
 
-![liver function labs](./cache/liver_labs_raw_data_distribution_by_rel_date.png)
+![liver function labs](./../../cache/liver_labs_raw_data_distribution_by_rel_date.png)
 
@@
 
-write.csv(liver_analysis, file = 'cache/liver_analysis.csv', row.names = FALSE)
+write.csv(liver_analysis, file = glue('{CACHE_DIR}/liver_analysis.csv'), row.names = FALSE)
 

### 2022-04-25 code/r/02_B_numeric_vs_hpo_for_acute_kidney_failure.Rmd

@@
 }
-FORCE_RERUN <- TRUE
+FORCE_RERUN <- FALSE
 ```
@@
 sql <- "SELECT * FROM cohort"
-df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'acute_kidney_failure_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
+df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'acute_kidney_failure_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
 

### 2022-04-25 code/r/02_C_numeric_vs_hpo_for_blood_disorder.Rmd
update

@@
 require(gridExtra)
+require(caret)
+require(broom)
+require(pROC)
 source('functions.R')
@@
 ```{r}
-require(caret)
-require(broom)
-require(pROC)
-
 # preprocess: convert the HPO column (and the isnegated column) into three separate features:

### 2022-04-26 code/r/03_A_biomarker.Rmd
separate biomarker screen into two sections

--- 2022-04-22 code/r/03_biomarker.Rmd
@@
 title: "Biomarker screening"
-output: html_notebook
+output:
+  html_document:
+    df_print: paged
 ---
@@
 
+This is Section A, data extraction from database.
+
+## Database connection
 ```{r message=FALSE}
@@
 
+Create a table for all those with J45 and R06 asthma and wheezing. 
 ```{sql connection=dsca}
@@
 
+Create a table for all those with severe asthma.
 ```{sql connection=dsca}
@@
 
-Combine cases and controls in one table. For now, we will use the 
+Combine cases and controls in one table. For now, we will use 2x case count for controls. 
 ```{sql connection=dsca}
@@
   left_join(cohort, by = 'medical_record_number') %>%
-  mutate(lab_date_rel = interval(t0, lab_date) / ddays()) %>% 
+  # replace below with sql
+  #mutate(lab_date_rel = interval(t0, lab_date) / ddays()) %>% 
+  mutate(lab_date_rel = sql("lab_date - t0")) %>% 
   filter(lab_date_rel < 0) %>%
   filter(lab_date_rel > -365 * 3) %>%
+  collect() 
+cohort_tested_phenotypes_yearly_count <- cohort_tested_phenotypes_yearly_count %>%
   mutate(lab_date_rel_bracket = cut(lab_date_rel / 365, breaks=seq(-3,0))) %>%
@@
   left_join(cohort, by = 'medical_record_number') %>%
-  mutate(lab_date_rel = interval(t0, lab_date) / ddays()) %>%
+  # replace below with sql
+  #mutate(lab_date_rel = interval(t0, lab_date) / ddays()) %>% 
+  mutate(lab_date_rel = sql("lab_date - t0")) %>% 
   filter(lab_date_rel < 0) %>%
   filter(lab_date_rel > -365 * 3) %>%
+  collect()
+cohort_observed_phenotypes_yearly_count <- cohort_observed_phenotypes_yearly_count %>% 
   mutate(lab_date_rel_bracket = cut(lab_date_rel / 365, breaks=seq(-3,0))) %>%
@@
 
+Combine phenotypes in one dataframe: one patient * phenotype per row.
 ```{r}
@@
   mutate(observed_count = ifelse(is.na(observed_count), 0, observed_count)) %>% 
-  mutate(is_tested = tested_count > 0) %>%
-  mutate(is_observed = observed_count > 0)
+  mutate(is_tested = tested_count > 0) %>% # a binary feature
+  mutate(is_observed = observed_count > 0) # a binary feature
+```
+
+The following is the dataframe for confounders (gender, race and age) + outcome (case or control)
+```{r}
+confounders_outcome <- cohort %>% 
+  inner_join(cohort_demographics, by = 'medical_record_number') %>% 
+  collect()
+confounders_outcome <- confounders_outcome %>%
+  mutate(age = interval(date_of_birth, t0) / dyears()) %>%
+  select(medical_record_number, gender, race, age, case_or_control) %>% 
+  filter(!is.na(age))
 ```
@@
 
-## Statistical analysis
-For the initial analysis, the model is as below:
+## Cache data for future use
 
-outcome ~ is_tested + is_tested:is_observed
-
-```{r}
-confounders_outcome <- cohort %>% 
-  inner_join(cohort_demographics, by = 'medical_record_number') %>%
-  mutate(age = interval(date_of_birth, t0) / dyears()) %>%
-  select(medical_record_number, gender, race, age, case_or_control) %>% 
-  filter(!is.na(age))
-
-d <- phenotype_features %>% 
-  filter(termid == 'HP:0001880') %>%
-  filter(lab_date_rel_bracket == '(-1,0]')
-
-d <- phenotype_features %>% 
-  filter(termid == 'HP:0001873') %>%
-  filter(lab_date_rel_bracket == '(-1,0]')
-
-prepare_Xy <- function(df_phenotype_feature, df_confounders_outcome){
-  df_model <- df_confounders_outcome %>% 
-    left_join(df_phenotype_feature, by = 'medical_record_number') %>%
-    select(medical_record_number, gender, race, age, tested_count, observed_count, is_tested, is_observed, case_or_control) %>%
-    ## fill in missing values
-    replace_na(list(tested_count = 0, observed_count = 0, is_tested = FALSE, is_observed = FALSE)) %>%
-    ## factorize categorical variables
-    mutate(gender = factor(gender)) %>%
-    mutate(race = factor(race)) %>%
-    mutate(is_tested = factor(is_tested)) %>%
-    mutate(is_observed = factor(is_observed)) %>%
-    mutate(case_or_control = factor(case_or_control)) %>%
-    ## set the reference levels for categorical variables
-    mutate(gender = relevel(gender, ref='Male')) %>%
-    mutate(race = relevel(race, ref='White')) %>%
-    mutate(is_tested = relevel(is_tested, ref = 'FALSE')) %>%
-    mutate(is_observed = relevel(is_observed, ref = 'FALSE')) %>%
-    mutate(case_or_control = relevel(case_or_control, ref = 'control'))
-  return (df_model)
-}
-
-glm_Xy <- function(Xy){
-  model <- glm(formula = 'case_or_control ~ gender + race + age + is_tested / is_observed', data = Xy, family = binomial(link = 'logit'))
-  model_coeff <- broom::tidy(model)
-  return (model_coeff)
-}
-
-df_model <- prepare_Xy(df_phenotype_feature = d %>% select(-lab_date_rel_bracket, -termid), 
-                       df_confounders_outcome = confounders_outcome)
-
-# df_model %>% sample_n(10)
-
-glm_Xy(df_model) %>% filter(str_starts(term, 'is_testedTRUE'))
-
-fit_model <- function(df_phenotype_feature, df_confounders_outcome) {
-  df_model <- prepare_Xy(df_phenotype_feature, df_confounders_outcome)
-  model_coeff <- data.frame(term = character(), estimate = double(), std.error = double(), statistic = double(), p.value = double())
-  tryCatch({model_coeff = glm_Xy(df_model)}, error = function(e) warning('cannot fit model'))
-  #error = function(e) print('model did not run'))
-  return (model_coeff)
-}
-
-fit_model(df_phenotype_feature = d %>% select(-lab_date_rel_bracket, -termid), 
-                       df_confounders_outcome = confounders_outcome)
-
-## explore other statistical models
-# model2 <- glmer(formula = 'case_or_control ~ gender + race + age + is_observed + (1 + is_observed | is_tested)', data = df_model, family = binomial(link = 'logit') )
-# summary(model2)
-```
-
-Screen all the phenotypes
-
-```{r}
-tic <- Sys.time()
-
-plan(multisession, workers = 8)
-
-screen_results <- phenotype_features %>% 
-  #filter(termid == 'HP:0001880') %>%
-  #filter(lab_date_rel_bracket == '(-1,0]') %>%
-  group_by(lab_date_rel_bracket, termid) %>%
-  nest() %>%
-  #calling ungroup() is critical for furrr to work in parallel
-  ungroup() %>%
-  mutate(coefficient = future_map(data, fit_model, df_confounders_outcome = confounders_outcome)) %>%
-  select(-data) %>%
-  unnest(coefficient)
-
-toc <- Sys.time()
-
-toc - tic
-```
-```{r}
-## how many biomarkers we have screened
-## at each time window
-N_BIOMARKERS_SCREENED <- screen_results %>% select(lab_date_rel_bracket, termid) %>% distinct() %>% nrow()
-screen_results %>% group_by(lab_date_rel_bracket) %>% summarise(n = n_distinct(termid))
-```
-
-Annotate the screen results by adding the HPO term label, and also create a few different views that rank the results differently. 
-```{r}
-hpo_term_list <- read.csv('../../in/hp_term_list.csv', header = FALSE, col.names = c('termid', 'distance_to_root', 'label'), stringsAsFactors = FALSE)
-
-screen_results_annotated <- screen_results %>%  filter(str_starts(term, 'is_testedTRUE')) %>% arrange(lab_date_rel_bracket) %>%
-  left_join(hpo_term_list, by = 'termid') %>%
-  select(lab_date_rel_bracket, termid, label, distance_to_root, term, distance_to_root, estimate, std.error, statistic, p.value)
-
-write.csv(screen_results_annotated, file = glue('{CACHE_DIR}/screen_results_annotated.csv'), row.names = FALSE)
-
-screen_results_annotated %>% head()
-```
-
-```{r}
-screen_results_annotated_formatted <- screen_results_annotated %>% 
-  mutate(OR = exp(estimate)) %>% 
-  mutate(OR = round(OR, 2)) %>% 
-  #mutate(p.value = scales::scientific(p.value, digits = 3)) %>% 
-  select(-estimate, -std.error, -statistic) %>% 
-  mutate(term = case_when(term == 'is_testedTRUE' ~ 'tested',
-                          term == 'is_testedTRUE:is_observedTRUE' ~ 'observed',
-                          TRUE ~ 'NEVER GONNA HAPPEN')) %>% 
-  pivot_wider(id_cols = c(lab_date_rel_bracket, termid, label, distance_to_root), names_from = term, values_from = c(OR, p.value)) %>% 
-  relocate(p.value_tested, .after = 'OR_tested') %>% 
-  rename(`Odds Ratio (Been Tested)` = OR_tested) %>% 
-  rename(`p value (Been Tested)` = p.value_tested) %>% 
-  rename(`Odds Ratio (Been Observed)` = 'OR_observed') %>% 
-  rename(`p value (Been Observed)` = 'p.value_observed')
-
-write.csv(screen_results_annotated_formatted, file = glue('{CACHE_DIR}/screen_results_annotated.csv', row.names = FALSE)
-
-screen_results_annotated_formatted %>% head()
-```
-
-Highlight some biomarkers:
-
-Eosinophilia
-```{r}
-eosinophilia <- screen_results_annotated_formatted %>% 
-  filter(label == 'Eosinophilia') %>% 
-  select(-distance_to_root) %>% 
-  mutate(across(starts_with('p value'), scales::scientific))
-
-sjPlot::tab_df(eosinophilia %>% rename(`Lab Year Bracket` = lab_date_rel_bracket), 
-                title = 'Table Association of Eosinophilia with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_eosinophilia_with_severe_asthma.html',
-               alternate.rows = TRUE)
-
-eosinophilia
-```
-
-Rank by Odds Ratio (Been Observed)
-```{r}
-significant_by_tested_and_observed <- screen_results_annotated_formatted %>% 
-  arrange(desc(`Odds Ratio (Been Observed)`)) %>% 
-  filter(`p value (Been Tested)` < 0.05/N_BIOMARKERS_SCREENED) %>%
-  filter(`p value (Been Observed)` < 0.05/N_BIOMARKERS_SCREENED) %>%
-  filter(`Odds Ratio (Been Tested)` > 1) %>% 
-  filter(`Odds Ratio (Been Observed)` > 1) %>% 
-  filter(lab_date_rel_bracket == '(-1,0]') %>%
-  mutate(across(starts_with('p value'), scales::scientific))
-
-significant_by_tested_and_observed_deduped <- significant_by_tested_and_observed %>% 
-  rows_delete(tibble(label = c('Abnormal eosinophil morphology', 
-                               'Abnormal leukocyte morphology',
-                               'Abnormal immune system morphology',
-                               'Abnormal cellular immune system morphology',
-                               'Abnormal leukocyte count',
-                               'Abnormal granulocyte morphology',
-                               'Abnormality of the respiratory system',
-                               'Functional respiratory abnormality')), by = 'label')
-
-significant_by_tested_and_observed_deduped
-```
-```{r}
-significant_by_tested_and_observed_deduped_three_years <- screen_results_annotated_formatted %>% 
-  filter(termid %in% significant_by_tested_and_observed_deduped$termid) %>% 
-  mutate(termid = factor(termid, levels = significant_by_tested_and_observed_deduped$termid)) %>% 
-  arrange(termid, lab_date_rel_bracket) %>%
-  mutate(across(starts_with('p value'), scales::scientific))
-
-significant_by_tested_and_observed_deduped_three_years
-```
-
-```{r}
-sjPlot::tab_df(significant_by_tested_and_observed_deduped %>% select(-distance_to_root, -lab_date_rel_bracket), 
-                title = 'Table Selected Phenotypes (within 1 year prior) Associated with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html', 
-               alternate.rows = TRUE)
-sjPlot::tab_df(significant_by_tested_and_observed_deduped_three_years %>% 
-                 select(-distance_to_root) %>% rename(`Lab Year Bracket` = lab_date_rel_bracket), 
-                title = 'Table Selected Phenotypes (within past 3 years) Associated with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html',
-               alternate.rows = TRUE)
-```
-
-## Cache data for future use
+Cache two dataframes for downstream statistical analysis: confounders_outcome and phenotype_features.
 ```{r}

### 2022-04-26 code/r/03_B_biomarker.Rmd
separate biomarker screen into two sections

--- 2022-04-22 code/r/03_biomarker.Rmd
@@
 title: "Biomarker screening"
-output: html_notebook
+output:
+  html_document:
+    df_print: paged
 ---
 
-In this section, we try to screen the phenotypes for biomarkers associated with pregression to severe asthma.
+This is Section B, statistical analysis.
 
@@
 library(readxl)
-require(DBI)
-require(RPostgres)
 require(egg)
@@
 require(furrr)
-require(dbplyr)
-
-dsca <- DBI::dbConnect(RPostgres::Postgres(), 
-                       host=Sys.getenv('DSCA_PHI_HOST'), 
-                       dbname = Sys.getenv('DSCA_PHI_DBNAME'), 
-                       port = Sys.getenv('DSCA_PHI_PORT'), 
-                       user=Sys.getenv('DSCA_PHI_USERNAME'),
-                       password= Sys.getenv('DSCA_PHI_PASSWORD'))
 
@@
 }
-FORCE_RERUN <- TRUE
+
+confounders_outcome <- read_csv(glue('{CACHE_DIR}/confounders_outcome.csv'))
+phenotype_features <- read_csv(glue('{CACHE_DIR}/phenotype_features.csv'))
 ```
-
-## Cohort 
-The starting cohort are patients with any asthma (J45) and asthma-related diagnosis.
-
-Within this starting cohort, we find one subset with severe asthma diagnosis.
-We further find a subset that has non-severe asthma diagnosis before they progressed to the severe form.
-These patients are the cases.
-
-Another subcohort are the ones that have never been associated with severe asthma. These patients are the controls.
-
-Final table has three columns:
-
-- medical_record_number
-- cohort: case or control
-- t0: earliest severe asthma dx date for cases and last asthma diagnosis for controls
-
-```{sql connection=dsca}
--- dx of asthma and asthma-related symptoms
-create temp table asthma_related_dx as 
-select medical_record_number , calendar_date , context_name , context_diagnosis_code 
-from hai_az_prod.diagnosis_2020july dj 
-where context_diagnosis_code ~* '^J45' or context_diagnosis_code ~* '^R06';
-```
-
-```{sql connection=dsca}
--- severe asthma
-create temp table severe_asthma_patients as
-select medical_record_number , min(calendar_date) as min_calendar_date
-from hai_az_prod.diagnosis_2020july dj 
-where context_diagnosis_code ~* '^J45\\.5'
-group by medical_record_number;  
-```
-
-A subset of severe asthma patients who were progressed from non-severe to severe. These are cases. 
-```{sql connection=dsca}
--- severe asthma progressed from non-severe: 2593
-create temp table progressed_to_severe_asthma_patients as
-select distinct medical_record_number 
-from hai_az_prod.diagnosis_2020july dj 
-join severe_asthma_patients using (medical_record_number)
-where (context_diagnosis_code ~* '^J45(\\.)?[2349]' or context_diagnosis_code ~* '^R06') and calendar_date < min_calendar_date;
-```
-
-Patients who had asthma but not severe. These are controls. 
-```{sql connection=dsca}
--- control : 323420
-create temp table never_severe_asthma_patients as
-select medical_record_number, max(calendar_date) as max_calendar_date, random() as random 
-from asthma_related_dx
-where medical_record_number not in (select medical_record_number from severe_asthma_patients)
-group by medical_record_number ;
-```
-
-Combine cases and controls in one table. For now, we will use the 
-```{sql connection=dsca}
--- combined cohort 
-create temp table cohort as 
-select medical_record_number , min_calendar_date as t0, 'case' as case_or_control
-from severe_asthma_patients 
-join progressed_to_severe_asthma_patients using (medical_record_number)
-union all
-select medical_record_number , max_calendar_date as t0, 'control' as case_or_control
-from never_severe_asthma_patients
-where random < 2593.0 * 2 / 323420;
-```
-
-## Extraction of data
-
-We need to extract the following information:
-
-- the cohort table: medical_record_number, cohort, t0
-- tested phenotypess: medical_record_number, termid, testdate
-- observed phenotypes: medical_record_number, termid, testdate
-- demographics: medical_record_number, sex, race, birthdate
-
-Get data from the database. The following method are lazily executed.
-```{r}
-cohort <- tbl(dsca, 'cohort')
-tested_phenotypes <- tbl(dsca, sql("
-    with testable_phenotypes_all as (
-    	select lab.medical_record_number , lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly
-    	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
-    	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid
-    	join cohort using (medical_record_number)
-      )
-     select medical_record_number, lab_date, termid
-     from testable_phenotypes_all"))
-observed_phenotypes <- tbl(dsca, sql("
-    with abnormal_lab as (
-      select lab.*
-      from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
-      join cohort using (medical_record_number)
-      where hpotermid is not null and not isnegated),
-    abnormal_lab_after_infer as (
-      SELECT l.medical_record_number, l.lab_date, date_part(year, lab_date) as lab_year, l.local_test_code, l.loinc, l.hpotermid, l.isnegated, pair.ancestor , pair.distance
-      FROM abnormal_lab l
-      left join hai_az_prod.hpo_is_a_pairs pair on l.hpotermid = pair.current)
-    select medical_record_number , lab_date, ancestor as termid
-    from abnormal_lab_after_infer"))
-cohort_demographics <- tbl(dsca, sql("
-    select person.*
-    from hai_az_prod.person_msdw_2020july person
-    join cohort using (medical_record_number)"))
-
-```
-
-## Data transformation
-
-- tested phenotypes: calculate testdate relative to t0; cut the relative date into yearly brackets and aggregated on yearly basis
-- observed phenotypes: similar to the above
-- demographics: calculate age at the time of t0
-
-```{r}
-cohort_tested_phenotypes_yearly_count <- tested_phenotypes %>%
-  left_join(cohort, by = 'medical_record_number') %>%
-  mutate(lab_date_rel = interval(t0, lab_date) / ddays()) %>% 
-  filter(lab_date_rel < 0) %>%
-  filter(lab_date_rel > -365 * 3) %>%
-  mutate(lab_date_rel_bracket = cut(lab_date_rel / 365, breaks=seq(-3,0))) %>%
-  group_by(medical_record_number, lab_date_rel_bracket, termid) %>% 
-  summarise(termid_count = n(), .groups = 'drop')
-```
-```{r}
-cohort_observed_phenotypes_yearly_count <- observed_phenotypes %>% 
-  left_join(cohort, by = 'medical_record_number') %>%
-  mutate(lab_date_rel = interval(t0, lab_date) / ddays()) %>%
-  filter(lab_date_rel < 0) %>%
-  filter(lab_date_rel > -365 * 3) %>%
-  mutate(lab_date_rel_bracket = cut(lab_date_rel / 365, breaks=seq(-3,0))) %>%
-  group_by(medical_record_number, lab_date_rel_bracket, termid) %>% 
-  summarise(termid_count = n(), .groups = 'drop')
-```
-
-```{r}
-## collect all the phenotypes in one dataframe
-phenotype_features <- cohort_tested_phenotypes_yearly_count %>% 
-  rename(tested_count = termid_count) %>%
-  left_join(cohort_observed_phenotypes_yearly_count %>% rename(observed_count = termid_count), 
-            by = c('medical_record_number', 'lab_date_rel_bracket', 'termid')) %>%
-  mutate(observed_count = ifelse(is.na(observed_count), 0, observed_count)) %>% 
-  mutate(is_tested = tested_count > 0) %>%
-  mutate(is_observed = observed_count > 0)
-```
-
-The final data frame should have the following columns:
-medical_record_number: patient id 
-termid: phenotype
-is_tested: TRUE or FALSE
-is_observed: TRUE or FALSE
-time_window: [-3, -2), [-2, -1), [-1, 0)
-sex
-race
-age: age at first severe asthma diagnosis (cases) or last asthma diagnosis (control)
-outcome: progressed to severe asthma or not
 
@@
 ```{r}
-confounders_outcome <- cohort %>% 
-  inner_join(cohort_demographics, by = 'medical_record_number') %>%
-  mutate(age = interval(date_of_birth, t0) / dyears()) %>%
-  select(medical_record_number, gender, race, age, case_or_control) %>% 
-  filter(!is.na(age))
-
+# test statistical modeling with one phenotype
 d <- phenotype_features %>% 
@@
 
-Screen all the phenotypes
+Screen all the phenotypes. Many phenotypes will fail to pass statistical modeling because small patient size, which is fine.
 
 ```{r}
+set.seed(45657)
 tic <- Sys.time()
@@
 
-write.csv(screen_results_annotated_formatted, file = glue('{CACHE_DIR}/screen_results_annotated.csv', row.names = FALSE)
+write.csv(screen_results_annotated_formatted, file = glue('{CACHE_DIR}/screen_results_annotated.csv'), row.names = FALSE)
 
@@
                 title = 'Table Association of Eosinophilia with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_eosinophilia_with_severe_asthma.html',
+               file = glue('{CACHE_DIR}/table_eosinophilia_with_severe_asthma.html'),
                alternate.rows = TRUE)
@@
                 title = 'Table Selected Phenotypes (within 1 year prior) Associated with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html', 
+               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html'), 
                alternate.rows = TRUE)
@@
                 title = 'Table Selected Phenotypes (within past 3 years) Associated with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html',
+               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html'),
                alternate.rows = TRUE)
 ```
-
-## Cache data for future use
-```{r}
-cache_data <- TRUE
-
-if (cache_data){
-  cache_list <- list(confounders_outcome = confounders_outcome,
-                     phenotype_features = phenotype_features)
-
-  for (i in 1:length(cache_list)){
-    filename <- names(cache_list)[[i]]
-    write.table(cache_list[[i]], file = glue('{CACHE_DIR}/{filename}.csv'), sep = ',', row.names = FALSE, col.names = TRUE)
-  }
-}
-```

### 2022-04-26 code/r/02_D_numeric_vs_hpo_for_CRC.Rmd
continue refactoring

@@
 
-ggsave(filename = 'cache/CRC_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 8, height = 6)
+ggsave(filename = glue('{CACHE_DIR}/CRC_labs_raw_data_distribution_by_rel_date.png'), plot = p, width = 8, height = 6)
 ```
 
-![liver function labs](./cache/CRC_labs_raw_data_distribution_by_rel_date.png)
+![CRC labs](../../cache/CRC_labs_raw_data_distribution_by_rel_date.png)
 
-Use -180 as the time cutoff and separate the patient journey into two sections: 0 - 180 days before the first diagnosis of abnormal liver function, and 180 days before the first diagnosis of liver abnormal functions.
+Use -180 as the time cutoff and separate the patient journey into two sections: 0 - 180 days before the first diagnosis of CRC, and >180 days before the first diagnosis of CRC.
 Within each section, we will pick the latest lab measurement for each LOINC lab test (features) and assign either + (within 180 days) or - outcome (outside of 180 days).
@@
 # briefly, the following snippet uses the following algorithm:
-# calculate the relative date to the first abnormal liver function diagnosis: if before diagnosis, - days otherwise + days
-# then cut the relative days into a few windows (-Inf, -365), (-365, -30), (-30, -7), (-7, 0) and (0, Inf)
+# calculate the relative date to the first diagnosis: if before diagnosis, - days otherwise + days
+# then cut the relative days into a few windows (-Inf, -180), (-180, 0) and (0, Inf)
 # for each window, we only use the closest one to diagnosis date as a representative, although we can use other methods such as the mean
-# at last, only the (-7, 0) window has + outcome, meaning having abnormal liver function; all other windows have normal liver function
 
@@
 
-```{r max.output=-1, message=FALSE}
-require(furrr)
-plan(multisession, workers = 5)
+Skip the following sections.
+<!-- ```{r max.output=-1, message=FALSE} -->
+<!-- require(furrr) -->
+<!-- plan(multisession, workers = 5) -->
 
-crc_analysis <- df_labs_and_outcomes %>%
-  group_by(loinc) %>%
-  nest() %>%
-  ungroup() %>%
-  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
-  select(-data) %>%
-  unnest(result)
+<!-- crc_analysis <- df_labs_and_outcomes %>% -->
+<!--   group_by(loinc) %>% -->
+<!--   nest() %>% -->
+<!--   ungroup() %>% -->
+<!--   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>% -->
+<!--   select(-data) %>% -->
+<!--   unnest(result) -->
 
-crc_analysis
-```
+<!-- crc_analysis -->
+<!-- ``` -->
 
-```{r message=FALSE}
-df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '4544-3')
+<!-- ```{r message=FALSE} -->
+<!-- df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '4544-3') -->
 
-df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc)
+<!-- df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc) -->
 
-df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE)
+<!-- df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE) -->
 
-df_train <- df_train_test$df_train
-df_test <- df_train_test$df_test
+<!-- df_train <- df_train_test$df_train -->
+<!-- df_test <- df_train_test$df_test -->
 
-lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
-roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
-roc_comparison <- bind_cols(roc_comparison)
+<!-- lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test) -->
+<!-- roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]]) -->
+<!-- roc_comparison <- bind_cols(roc_comparison) -->
 
-plot(lr_L[[5]])
-plot(lr_L[[6]])
+<!-- plot(lr_L[[5]]) -->
+<!-- plot(lr_L[[6]]) -->
 
-roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap")
-```
+<!-- roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap") -->
+<!-- ``` -->
 
@@
 
-```{r max.output=-1, message=FALSE}
-require(furrr)
-plan(multisession, workers=8)
-start_time <- Sys.time()
+Skip the following code sections.
 
-crc_separate_controls_analysis <- df_labs_and_outcomes_case_control %>%
-  group_by(loinc) %>%
-  nest() %>%
-  ungroup() %>%
-  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, .options = furrr_options(seed = T))) %>%
-  select(-data) %>%
-  unnest(result)
+<!-- ```{r max.output=-1, message=FALSE} -->
+<!-- require(furrr) -->
+<!-- plan(multisession, workers=8) -->
+<!-- start_time <- Sys.time() -->
 
-end_time <- Sys.time()
+<!-- crc_separate_controls_analysis <- df_labs_and_outcomes_case_control %>% -->
+<!--   group_by(loinc) %>% -->
+<!--   nest() %>% -->
+<!--   ungroup() %>% -->
+<!--   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, .options = furrr_options(seed = T))) %>% -->
+<!--   select(-data) %>% -->
+<!--   unnest(result) -->
 
-print(end_time - start_time)
+<!-- end_time <- Sys.time() -->
 
-crc_separate_controls_analysis
-```
+<!-- print(end_time - start_time) -->
 
-```{r message=FALSE}
-df_one_loinc <- df_labs_and_outcomes_case_control %>% filter(loinc == '4544-3')
+<!-- crc_separate_controls_analysis -->
+<!-- ``` -->
 
-df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc)
+<!-- ```{r message=FALSE} -->
+<!-- df_one_loinc <- df_labs_and_outcomes_case_control %>% filter(loinc == '4544-3') -->
 
-df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE)
+<!-- df_hpo_features <- create_hpo_features(df_one_loinc = df_one_loinc) -->
 
-df_train <- df_train_test$df_train
-df_test <- df_train_test$df_test
+<!-- df_train_test <- train_test_split(df_hpo_features, p = 0.7, list=FALSE) -->
 
-lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test)
-roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]])
-roc_comparison <- bind_cols(roc_comparison)
+<!-- df_train <- df_train_test$df_train -->
+<!-- df_test <- df_train_test$df_test -->
 
-plot(lr_L[[5]])
-plot(lr_L[[6]])
+<!-- lr_L <- lr_for_hpo_L(df_train = df_train, df_test = df_test) -->
+<!-- roc_comparison <- compare_rocs(lr_L[[5]], lr_L[[6]]) -->
+<!-- roc_comparison <- bind_cols(roc_comparison) -->
 
-roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap")
-```
+<!-- plot(lr_L[[5]]) -->
+<!-- plot(lr_L[[6]]) -->
+
+<!-- roc.test(lr_L[[5]], lr_L[[6]], method = "bootstrap") -->
+<!-- ``` -->
 
@@
 
-ggplot(df_demographics_with_outcomes, aes(x = outcome, y = propensity_score)) +
-  geom_violin()
+p <- ggplot(df_demographics_with_outcomes, aes(x = outcome, y = propensity_score)) +
+  geom_violin() +
+  theme_az()
+
+ggsave(glue("{CACHE_DIR}/plot_CRC_before_matching.png"), width = 2, height = 1.5)
+p
 ```
@@
 # still slow, but works
-df_demographics_with_outcomes %>% colnames()
-m.out <- matchit(outcome ~ medical_record_number, data = df_demographics_with_outcomes, method = "nearest", distance = df_demographics_with_outcomes$propensity_score)
-save(m.out, file='m.out.obj')
+# df_demographics_with_outcomes %>% colnames()
+
+run_matchit <- function(filepath, force_rerun) {
+  if (force_rerun | !file.exists(filepath)){
+    # place m.out in the global environment with "<<-" (instead of regular "<-")
+    m.out <<- matchit(outcome ~ medical_record_number, data = df_demographics_with_outcomes, method = "nearest", distance = df_demographics_with_outcomes$propensity_score)
+    # save the matchit object in cache
+    save(m.out, file=filepath)
+  }
+  load(file=filepath)
+}
+
+run_matchit(filepath = glue('{CACHE_DIR}/m.out.obj'), force_rerun = FALSE)
+
 matched_unique_ids_etc <- match.data(m.out, data = df_demographics_with_outcomes)
 
-matched_unique_ids_etc %>% group_by(outcome) %>% summarise(n =n())
-ggplot(matched_unique_ids_etc, aes(x = outcome, y = propensity_score)) +
-  geom_violin()
+#matched_unique_ids_etc %>% group_by(outcome) %>% summarise(n =n())
+p <- ggplot(matched_unique_ids_etc, aes(x = outcome, y = propensity_score)) +
+  geom_violin() + 
+  theme_az()
+
+ggsave(glue("{CACHE_DIR}/plot_CRC_after_matching.png"), plot = p, width = 2, height = 1.5)
 
@@
   geom_violin(draw_quantiles = c(0.5)) +
-  theme(legend.position = 'none')
+  theme(legend.position = 'none') + 
+  theme_az()
 
 plot_match_gender <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = gender)) +
-  geom_bar(stat = 'count')
+  geom_bar(stat = 'count') + 
+  theme_az()
 
 plot_match_race <- ggplot(matched_unique_ids_etc, aes(x = outcome, fill = race)) +
-  geom_bar(stat = 'count')
+  geom_bar(stat = 'count')+ 
+  theme_az()
 
 plot_match <- gridExtra::arrangeGrob(plot_match_age, plot_match_gender, plot_match_race, layout_matrix = rbind(c(1,2), c(3,3)), ncol = 1)
+
+ggsave(filename = glue('{CACHE_DIR}/plot_CRC_age_gender_race_after_matching.png'), plot = plot_match, width = 6, height = 5)
 plot(plot_match)
@@
 ```{r}
-dbExecute(dsca, "create table if not exists matched_control_cohort (medical_record_number VARCHAR)")
-dbExecute(dsca, "truncate table matched_control_cohort")
+dbExecute(dsca, "create temp table if not exists matched_control_cohort (medical_record_number VARCHAR)")
 
@@
 
+# insert mrns one by one
 matched_unique_ids_etc %>% filter(outcome == 'normal') %>% 
@@
 sql <- read_file('../sql/get_labs_for_colorectal_cancer_matched_control.sql')
-df_matched_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_matched_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = TRUE)
-
-# drop the matched control cohort
-# persist it for a while for now
-# n <- dbExecute(dsca, "DROP TABLE IF EXISTS matched_control_cohort")
-# assertthat::are_equal(n, 0)
+df_matched_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_matched_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)
 ```
@@
 
-```{r max.output=-1, message=FALSE}
+```{r message=FALSE, max.output=-1}
 require(furrr)
-plan(multisession, workers=4)
+plan(multisession, workers=8)
 start_time <- Sys.time()
@@
 
-crc_case_matched_controls_analysis %>% write.csv(file = 'cache/crc_case_matched_controls_analysis.csv', row.names = FALSE)
+crc_case_matched_controls_analysis %>% write.csv(file = glue('{CACHE_DIR}/crc_case_matched_controls_analysis.csv'), row.names = FALSE)
 
 crc_case_matched_controls_analysis
-
 ```
 
-Combine demographics to see whether the AUC can be improved. 
-```{r}
-df_demographics_case_matched_control <- bind_rows(df_case_demographics %>% select(-min_diag_date), 
-                                                  df_control_demographics %>% filter(medical_record_number %in% 
-                                                                                       unique(df_matched_control_labs$medical_record_number)) %>%
-                                                    filter(!(medical_record_number %in% df_case_demographics$medical_record_number)))
+<!-- Combine demographics to see whether the AUC can be improved. It is not used in the manuscript.  -->
+<!-- ```{r} -->
+<!-- df_demographics_case_matched_control <- bind_rows(df_case_demographics %>% select(-min_diag_date),  -->
+<!--                                                   df_control_demographics %>% filter(medical_record_number %in%  -->
+<!--                                                                                        unique(df_matched_control_labs$medical_record_number)) %>% -->
+<!--                                                     filter(!(medical_record_number %in% df_case_demographics$medical_record_number))) -->
 
-plan(multisession, workers=4)
-start_time <- Sys.time()
+<!-- plan(multisession, workers=4) -->
+<!-- start_time <- Sys.time() -->
 
-crc_case_matched_controls_analysis_with_demographics <- df_labs_and_outcomes_case_matched_control %>%
-  left_join(df_demographics_case_matched_control, by = 'medical_record_number') %>%
-  mutate(age = as.duration(interval(date_of_birth, lab_date)) / dyears()) %>%
-  filter(!(race %in% c('African American', 'American Indian or Alaska Native', 'Native Hawaiian or Other Pacific Islander'))) %>%
-  # mutate(gender = as.factor(gender),
-  #        race = as.factor(race)) %>% 
-  # mutate(gender = relevel(gender, 'Male'),
-  #        race = relevel(race, 'White')) %>%
-  group_by(loinc) %>%
-  nest() %>%
-  ungroup() %>% 
-  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, include_demographics=TRUE, .options = furrr_options(seed = T))) %>%
-  select(-data) %>%
-  unnest(result)
+<!-- crc_case_matched_controls_analysis_with_demographics <- df_labs_and_outcomes_case_matched_control %>% -->
+<!--   left_join(df_demographics_case_matched_control, by = 'medical_record_number') %>% -->
+<!--   mutate(age = as.duration(interval(date_of_birth, lab_date)) / dyears()) %>% -->
+<!--   filter(!(race %in% c('African American', 'American Indian or Alaska Native', 'Native Hawaiian or Other Pacific Islander'))) %>% -->
+<!--   # mutate(gender = as.factor(gender), -->
+<!--   #        race = as.factor(race)) %>%  -->
+<!--   # mutate(gender = relevel(gender, 'Male'), -->
+<!--   #        race = relevel(race, 'White')) %>% -->
+<!--   group_by(loinc) %>% -->
+<!--   nest() %>% -->
+<!--   ungroup() %>%  -->
+<!--   mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, include_demographics=TRUE, .options = furrr_options(seed = T))) %>% -->
+<!--   select(-data) %>% -->
+<!--   unnest(result) -->
 
-end_time <- Sys.time()
+<!-- end_time <- Sys.time() -->
 
-print(end_time - start_time)
+<!-- print(end_time - start_time) -->
 
-crc_case_matched_controls_analysis_with_demographics
-```
+<!-- crc_case_matched_controls_analysis_with_demographics -->
+<!-- ``` -->

### 2022-04-26 code/r/02_E_numeric_vs_hpo_combined.Rmd
update
continue refactoring

@@
 title: "numeric vs HPO"
-output: html_notebook
+output:
+  html_document:
+    df_print: paged
 ---
@@
           axis.title = element_text(size = 8), 
-          strip.text.x = element_text(margin = margin(0.3, 0, 0.3, 0), size = 5),
+          strip.text.x = element_text(margin = margin(0.3, 0, 0.3, 0), size = 8),
           plot.title = element_text(size = 8)) +
@@
 
-ggsave(filename = glue('{CACHE_DIR}/plot_AUC_differences_numeric_vs_hpo.pdf'), width = 6, height = 6)
+ggsave(filename = glue('{CACHE_DIR}/plot_AUC_differences_numeric_vs_hpo.pdf'), width = 6, height = 5)
 

### 2022-04-26 code/r/functions.R

@@ -22,3 +22,13 @@ get_df_from_db_or_cache <- function(cache_dir, file_name,
   # data.table::fread is typically preferred
   df <- data.table::fread(df_path, ...)
 }
+
+
+#' my default theme
+theme_az <- function(){
+  theme_bw() + 
+    theme(panel.grid = element_blank(), legend.position = "right", 
+          axis.title = element_text(size = 8), 
+          axis.text = element_text(size = 8), 
+          legend.text = element_text(size = 8), legend.title = element_text(size = 10))
+}

### 2022-04-27 code/r/01_MSDW_HPO_Phenotype_Analysis.Rmd
update

--- 2022-04-22 code/r/MSDW_HPO_Phenotype_Analysis.Rmd
@@
 require(gridExtra)
+source('functions.R')
 dsca <- DBI::dbConnect(RPostgres::Postgres(), 
@@
   mutate(gender = recode(gender, .missing = 'Unknown'))
-
-# the following cleaning are no longer needed after updating the race mapping file
-# recode "African American" as "Black or African American"
-# patient_race_sex_distribution <- patient_race_sex_distribution %>%
-#   mutate(race = recode(race, 'African American' = 'Black or African American'))
 ```
@@
   group_by(gender) %>% 
-  summarise(n = sum(n))
+  summarise(n = sum(n)) %>% 
+  mutate(fraction = n / sum(n))
 
@@
   coord_polar(theta = "y") + 
-  theme_bw() + 
-  theme(panel.grid = element_blank(), legend.position = c(0.25, 0.7), 
+  theme_az() + 
+  theme(legend.position = c(0.25, 0.7), 
         axis.text = element_blank(), axis.ticks = element_blank(), 
-        axis.title = element_text(size = 8), legend.text = element_text(size = 8), 
-        legend.title = element_text(size = 10), 
         legend.background = element_rect(fill=alpha('white', 0.1)))
@@
 ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.png'), plot = plot_by_sex, width = 2.5, height = 2.5)
+ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.pdf'), plot = plot_by_sex, width = 2.5, height = 2.5)
 plot_by_sex
@@
   coord_flip() + 
-  theme_bw() + 
-  theme(panel.grid = element_blank(), legend.position = "right", 
-        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
-        legend.text = element_text(size = 8), legend.title = element_text(size = 10), 
-        legend.background = element_rect(fill=alpha('white', 0.1)))
+  theme_az() + 
+  theme(legend.background = element_rect(fill=alpha('white', 0.1)))
 
 ggsave(filename = glue("{CACHE_DIR}/plotByRace.png"), plot = plot_by_race, width = 5, height = 2)
+ggsave(filename = glue("{CACHE_DIR}/plotByRace.pdf"), plot = plot_by_race, width = 5, height = 2)
 plot_by_race
+```
+Fraction of each race:
+```{r}
+countByRace_Gender %>% 
+  group_by(race) %>% 
+  summarise(n = sum(n)) %>%
+  mutate(fraction = n / sum(n)) %>% 
+  arrange(desc(fraction))
 ```
@@
 
-patient_birth_year_sex_distribution <- get_patient_birth_year_distribution(FALSE)
+patient_birth_year_sex_distribution <- get_patient_birth_year_distribution(force_rerun = FORCE_RERUN)
 
@@
   coord_flip() + 
-      theme_bw() + 
-  theme(panel.grid = element_blank(), legend.position = "right", 
-        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
-        legend.text = element_text(size = 8), legend.title = element_text(size = 10), 
-        legend.background = element_rect(fill=alpha('white', 0.1)))
+  theme_az() + 
+  theme(legend.background = element_rect(fill=alpha('white', 0.1)))
   }
@@
 ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.png"), plot = plot_by_birth_year_sex, width = 4, height = 3)
+ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.pdf"), plot = plot_by_birth_year_sex, width = 4, height = 3)
 plot_by_birth_year_sex
-
 ```

### 2022-04-27 code/r/01_MSDW_HPO_Phenotype_Analysis.Rmd
update analysis

@@
 
+OUT_DIR = '../../out'
+if (!dir.exists(OUT_DIR)){
+  dir.create(OUT_DIR)
+}
+
 get_db_data <- function(sql, df_path, force_rerun=FALSE, cache_dir=CACHE_DIR){
@@
 
-ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.png'), plot = plot_by_sex, width = 2.5, height = 2.5)
-ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.pdf'), plot = plot_by_sex, width = 2.5, height = 2.5)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_by_sex.png'), glue('{OUT_DIR}/plot_by_sex.pdf')), plot = plot_by_sex, width = 2.5, height = 2.5)
 plot_by_sex
@@
 
-ggsave(filename = glue("{CACHE_DIR}/plotByRace.png"), plot = plot_by_race, width = 5, height = 2)
-ggsave(filename = glue("{CACHE_DIR}/plotByRace.pdf"), plot = plot_by_race, width = 5, height = 2)
+ggsave2(filenames = c(glue("{OUT_DIR}/plotByRace.png"), glue("{OUT_DIR}/plotByRace.pdf")), plot = plot_by_race, width = 5, height = 2)
 plot_by_race
@@
 
-ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.png"), plot = plot_by_birth_year_sex, width = 4, height = 3)
-ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.pdf"), plot = plot_by_birth_year_sex, width = 4, height = 3)
+ggsave2(filenames = c(glue("{OUT_DIR}/plotByBirthyearSex.png"), glue("{OUT_DIR}/plotByBirthyearSex.pdf")), plot = plot_by_birth_year_sex, width = 4, height = 3)
 plot_by_birth_year_sex
@@
         legend.text = element_text(size = 8), legend.title = element_text(size = 10))
-ggsave(filename = glue("{CACHE_DIR}/plotByPhenotypesPerPatient.png"), plot = plot_by_phenotypes_per_patient, width = 3, height = 2)
+ggsave2(filenames = c(glue("{OUT_DIR}/plotByPhenotypesPerPatient.png"), glue("{OUT_DIR}/plotByPhenotypesPerPatient.pdf")), plot = plot_by_phenotypes_per_patient, width = 3, height = 2)
 plot_by_phenotypes_per_patient
@@
         legend.text = element_text(size = 8), legend.title = element_text(size = 10))
-ggsave(filename = glue("{CACHE_DIR}/plotByDistinctAbnormalPhenotypeInferPerPatient.png"), plot = plot_by_distinct_abnormal_hpo_infer_per_patient, width = 3, height = 2)
+ggsave2(filenames = c(glue("{OUT_DIR}/plotByDistinctAbnormalPhenotypeInferPerPatient.png"), glue("{OUT_DIR}/plotByDistinctAbnormalPhenotypeInferPerPatient.pdf")), plot = plot_by_distinct_abnormal_hpo_infer_per_patient, width = 3, height = 2)
 plot_by_distinct_abnormal_hpo_infer_per_patient
@@
   summarise(display = str_c(display, collapse = '\n'), .groups = 'drop')
-write.csv(patient_lab_phenotype_journeys_formated, file = glue('{CACHE_DIR}/patient_lab_phenotype_journey_formatted.csv'))
+write.csv(patient_lab_phenotype_journeys_formated, file = glue('{OUT_DIR}/patient_lab_phenotype_journey_formatted.csv'))
 ```
@@
 p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=phenotype_rank_all_by_year, normalize = FALSE)
-ggsave(filename = glue('{CACHE_DIR}/plot_abnormal_phenotypes_by_year_level_1.pdf'), plot = p, width = 8, height = 4)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_abnormal_phenotypes_by_year_level_1.png'), glue('{OUT_DIR}/plot_abnormal_phenotypes_by_year_level_1.pdf')), plot = p, width = 8, height = 4)
 p
@@
                                            left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
-ggsave(filename = glue('{CACHE_DIR}/plot_abnormal_phenotypes_by_year_level_1_normalized.pdf'), plot = p, width = 8, height = 4)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_abnormal_phenotypes_by_year_level_1_normalized.png'), glue('{OUT_DIR}/plot_abnormal_phenotypes_by_year_level_1_normalized.pdf')), plot = p, width = 8, height = 4)
 p
@@
 
-write.csv(trendline_lm_test_formatted, file = glue('{CACHE_DIR}/table_abnormal_phenotypes_by_year_level_1_normalized_trendline_lm_model.csv'), row.names=FALSE)
+write.csv(trendline_lm_test_formatted, file = glue('{OUT_DIR}/table_abnormal_phenotypes_by_year_level_1_normalized_trendline_lm_model.csv'), row.names=FALSE)
 
@@
 p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year, normalize = FALSE)
-ggsave(filename = glue('{CACHE_DIR}/plot_tested_phenotypes_by_year_level_1.pdf'), plot = p, width = 8, height = 4)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_tested_phenotypes_by_year_level_1.png'), glue('{OUT_DIR}/plot_tested_phenotypes_by_year_level_1.pdf')), plot = p, width = 8, height = 4)
 p
@@
 p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
-ggsave(filename = glue('{CACHE_DIR}/plot_tested_phenotypes_by_year_level_1_normalized.pdf'), plot = p, width = 8, height = 4)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_tested_phenotypes_by_year_level_1_normalized.png'), glue('{OUT_DIR}/plot_tested_phenotypes_by_year_level_1_normalized.pdf')), plot = p, width = 8, height = 4)
 p
@@
 
-write.csv(trendline_lm_test_formatted, file = glue('{CACHE_DIR}/table_tested_phenotypes_by_year_level_1_normalized_trendline_lm_model.csv'), row.names=FALSE)
+write.csv(trendline_lm_test_formatted, file = glue('{OUT_DIR}/table_tested_phenotypes_by_year_level_1_normalized_trendline_lm_model.csv'), row.names=FALSE)
 
@@
 
-write.csv(racial_abnormality_fisher, file = glue('{CACHE_DIR}/racial_abnormality_fisher.csv'), row.names = FALSE)
+write.csv(racial_abnormality_fisher, file = glue('{OUT_DIR}/racial_abnormality_fisher.csv'), row.names = FALSE)
 
@@
 
-ggsave(filename = glue('{CACHE_DIR}/plot_by_phenotype_racial_difference_over_years.pdf'), plot = p, width = 8, height = 8)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_by_phenotype_racial_difference_over_years.png'), glue('{OUT_DIR}/plot_by_phenotype_racial_difference_over_years.pdf')), plot = p, width = 8, height = 8)
 
@@
 
-write.csv(racial_tested_abnormality_fisher, file = glue('{CACHE_DIR}/racial_tested_abnormality_fisher.csv'), row.names = FALSE)
+write.csv(racial_tested_abnormality_fisher, file = glue('{OUT_DIR}/racial_tested_abnormality_fisher.csv'), row.names = FALSE)
 
@@
 
-ggsave(filename = glue('{CACHE_DIR}/plot_by_tested_phenotype_racial_difference_over_years.pdf'), plot = p, width = 8, height = 8)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_by_tested_phenotype_racial_difference_over_years.png'), glue('{OUT_DIR}/plot_by_tested_phenotype_racial_difference_over_years.pdf')), plot = p, width = 8, height = 8)
 
@@
 
-write.csv(overly_tested_abnormalities_by_each_race_top10, file=glue('{CACHE_DIR}/overly_tested_abnormalities_by_each_race_top10.csv'), row.names = FALSE)
+write.csv(overly_tested_abnormalities_by_each_race_top10, file=glue('{OUT_DIR}/overly_tested_abnormalities_by_each_race_top10.csv'), row.names = FALSE)
 
@@
                                                  use.viewer = FALSE, 
-                                                 file = glue('{CACHE_DIR}/table_overly_tested_abnormalities_for_{race_name}_top10.html')))
+                                                 file = glue('{OUT_DIR}/table_overly_tested_abnormalities_for_{race_name}_top10.html')))
   }
@@
                      use.viewer = FALSE, 
-                     file = glue('{CACHE_DIR}/table_overly_tested_abnormalities_for_{race_name}_top10.html'))
+                     file = glue('{OUT_DIR}/table_overly_tested_abnormalities_for_{race_name}_top10.html'))
 }

### 2022-04-27 code/r/02_A_numeric_vs_hpo_for_liver_function.Rmd
update analysis

@@
 }
+
+# output directory to write analysis results
+OUT_DIR = '../../out'
+if (!dir.exists(OUT_DIR)){
+  dir.create(OUT_DIR)
+}
+
+# if true, ignore cached data and reconnect to database to get latest data;
+# otherwise, use cached data first before trying to connect to database
 FORCE_RERUN <- FALSE
@@
 
-ggsave(filename = glue('{CACHE_DIR}/liver_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6))
+ggsave2(filenames = c(glue('{OUT_DIR}/liver_labs_raw_data_distribution_by_rel_date.png'), glue('{OUT_DIR}/liver_labs_raw_data_distribution_by_rel_date.pdf')), plot = p, width = 6, height = 6)
 ```
@@
 
-write.csv(liver_analysis, file = glue('{CACHE_DIR}/liver_analysis.csv'), row.names = FALSE)
+write.csv(liver_analysis, file = glue('{OUT_DIR}/liver_analysis.csv'), row.names = FALSE)
 

### 2022-04-27 code/r/02_B_numeric_vs_hpo_for_acute_kidney_failure.Rmd
update analysis

@@
 
-ggsave(filename = 'cache/kidney_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
+ggsave2(filenames = c(glue('{OUT_DIR}/kidney_labs_raw_data_distribution_by_rel_date.png'), glue('{OUT_DIR}/kidney_labs_raw_data_distribution_by_rel_date.pdf')), plot = p, width = 6, height = 6)
 ```
 
-![liver function labs](./cache/kidney_labs_raw_data_distribution_by_rel_date.png)
+![liver function labs](./../../out/kidney_labs_raw_data_distribution_by_rel_date.png)
 
@@
 
-write.csv(kidney_analysis, file = 'cache/acute_kidney_failure_analysis.csv', row.names = FALSE)
+write.csv(kidney_analysis, file = glue('{OUT_DIR}/acute_kidney_failure_analysis.csv'), row.names = FALSE)
 

### 2022-04-27 code/r/02_C_numeric_vs_hpo_for_blood_disorder.Rmd
update analysis

@@
 ## Goal
-Compare the performance of raw lab values vs HPO interpretations for seme blood disorders
+Compare the performance of raw lab values vs HPO interpretations for some blood disorders
 
@@
 
+# output directory to write analysis results
+OUT_DIR = '../../out'
+if (!dir.exists(OUT_DIR)){
+  dir.create(OUT_DIR)
+}
+
+# if true, ignore cached data and reconnect to database to get latest data;
+# otherwise, use cached data first before trying to connect to database
 FORCE_RERUN <- FALSE
@@
 
-ggsave(filename = 'cache/anemia_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 5, height = 5)
+ggsave2(filenames = c(glue('{OUT_DIR}/anemia_labs_raw_data_distribution_by_rel_date.png'), glue('{OUT_DIR}/anemia_labs_raw_data_distribution_by_rel_date.pdf')), plot = p, width = 5, height = 5)
 ```
 
-![liver function labs](./cache/anemia_labs_raw_data_distribution_by_rel_date.png)
+![liver function labs](./../../out/anemia_labs_raw_data_distribution_by_rel_date.png)
 
@@
 
-write.csv(anemia_analysis, file = 'cache/anemia_analysis.csv', row.names = FALSE)
+write.csv(anemia_analysis, file = glue('{OUT_DIR}/anemia_analysis.csv'), row.names = FALSE)
 

### 2022-04-27 code/r/02_E_numeric_vs_hpo_combined.Rmd
update analysis

@@
 
+source('functions.R')
+
 # cache directory to read from or write to
 CACHE_DIR = '../../cache'
+
+# output directory to write analysis results
+OUT_DIR = '../../out'
+if (!dir.exists(OUT_DIR)){
+  dir.create(OUT_DIR)
+}
+
 ```
@@
 ```{r}
-liver_analysis <- read.csv(glue('{CACHE_DIR}/liver_analysis.csv'))
-crc_analysis <- read.csv(glue('{CACHE_DIR}/crc_case_matched_controls_analysis.csv'))
-acute_kidney_failure_analysis <- read.csv(glue('{CACHE_DIR}/acute_kidney_failure_analysis.csv'))
-anemia_analysis <- read.csv(glue('{CACHE_DIR}/anemia_analysis.csv'))
+liver_analysis <- read.csv(glue('{OUT_DIR}/liver_analysis.csv'))
+crc_analysis <- read.csv(glue('{OUT_DIR}/crc_case_matched_controls_analysis.csv'))
+acute_kidney_failure_analysis <- read.csv(glue('{OUT_DIR}/acute_kidney_failure_analysis.csv'))
+anemia_analysis <- read.csv(glue('{OUT_DIR}/anemia_analysis.csv'))
 ```
@@
   geom_line(aes(color = disease)) +
-  scale_x_discrete(breaks = c('continuous', 'hpo'), labels = c('original numeric value', 'transformed to HPO')) +
+  scale_x_discrete(breaks = c('continuous', 'hpo'), labels = c('numeric', 'HPO')) +
   xlab("Use Lab Test Results As") +
@@
 
-ggsave(filename = glue('{CACHE_DIR}/plot_AUC_differences_numeric_vs_hpo.pdf'), width = 6, height = 5)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_AUC_differences_numeric_vs_hpo.png'), glue('{OUT_DIR}/plot_AUC_differences_numeric_vs_hpo.pdf')), plot = plot_AUC_differences_numeric_vs_hpo, width = 3.2, height = 3.2)
 

### 2022-04-27 code/r/03_A_biomarker.Rmd
update analysis

@@
 CACHE_DIR = '../../cache'
-
 if (!dir.exists(CACHE_DIR)){
@@
 }
-FORCE_RERUN <- TRUE
+
+# if true, ignore cached data and reconnect to database to get latest data;
+# otherwise, use cached data first before trying to connect to database
+FORCE_RERUN <- FALSE
 ```

### 2022-04-27 code/r/functions.R

@@ -32,3 +32,11 @@ theme_az <- function(){
           axis.text = element_text(size = 8), 
           legend.text = element_text(size = 8), legend.title = element_text(size = 10))
 }
+
+#' my ggsave
+#' it extends ggplot2::ggsave by saving one plot to multiple files
+ggsave2 <- function(filenames, ...){
+  for (filename in filenames){
+    ggsave(filename, ...)
+  }
+}

### 2022-04-28 code/r/01_A_MSDW_HPO_Phenotype_Analysis.Rmd
observed normalized by tested

--- 2022-04-27 code/r/01_MSDW_HPO_Phenotype_Analysis.Rmd
@@
     ylab(y_label) +
-    theme_bw() +
-    theme(panel.grid = element_blank(),
-        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
-        legend.text = element_text(size = 8), legend.title = element_text(size = 10))
+    theme_az()
 

### 2022-04-28 code/r/01_B.Rmd

---
title: "01_B_observed_normalized_by_tested"
output: html_document
date: '2022-04-27'
author: Aaron Zhang
---

This is a small extension for the main analysis of "01_MSDW_HPO_Phenotype_Analysis.Rmd". 
The goal is to analyze the trendlines of observed phenotypes but now normalize by how many patients were tested for each corresponding phenotype.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(glue)
source('functions.R')
```

## load data
```{r}
# cache directory to read from or write to
CACHE_DIR = '../../cache'
if (!dir.exists(CACHE_DIR)){
  dir.create(CACHE_DIR)
}

# output directory to write analysis results
OUT_DIR = '../../out'
if (!dir.exists(OUT_DIR)){
  dir.create(OUT_DIR)
}

tested_phenotype_rank_all_by_year <- read_csv(glue('{CACHE_DIR}/tested_phenotype_rank_all_by_year.csv'))
observed_phenotype_rank_all_by_year <- read_csv(glue('{CACHE_DIR}/observed_phenotype_rank_all_by_year.csv'))
```

## analysis
Calculate fraction of patients observed for each phenotype by patients tested for that phenotype. 
```{r}
observed_divide_tested_all_by_year <- observed_phenotype_rank_all_by_year %>% 
  rename(n_patient_observed = n_patient) %>% 
  select(-n_record) %>%
  left_join(tested_phenotype_rank_all_by_year %>% 
              rename(n_patient_tested = n_patient) %>% 
              select(termid, lab_year, n_patient_tested),
            by = c('termid', 'lab_year')) %>%
  relocate(n_patient_tested, .after = n_patient_observed) %>%
  mutate(frac_observed_tested = n_patient_observed / n_patient_tested)
```

Below, we plot level HPO terms.
```{r}
plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df){
  hpo_terms_selected <- df %>% filter(termid %in% terms_of_interest) %>% select(termid, label) %>% distinct()

  y_str = 'frac_observed_tested'
  y_label = 'patient fraction (observed / tested)'

  p <- df %>% filter(termid %in% terms_of_interest) %>% 
    filter(lab_year <= 2019 & lab_year > 2000) %>%
    ggplot(aes(x = lab_year, y = eval(parse(text=y_str)), group = termid, color = termid)) + 
    geom_point() +
    geom_line() +
    scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " ")) +
    xlab('Year') +
    ylab(y_label) +
    theme_az()

  return (p)
}
```

```{r}
hpo_level_1_terms <- observed_phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()

p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=observed_divide_tested_all_by_year)

p
```

Run statistical testing of the trendlines. It is probably a better metric to claim whether the cohort is more or less healthier for one phenotype as this is normalized by patients tested for that phenotype.

```{r output.max=-1}
trendline_lm_test <- observed_divide_tested_all_by_year %>% 
  filter(distancetoroot == 1) %>% 
  filter(lab_year > 2000) %>%
  rename(fraction = frac_observed_tested) %>%
  select(termid, lab_year, label, fraction) %>% 
  group_by(termid, label) %>% 
  nest() %>% 
  mutate(lm_model = map(data, function(d) lm(fraction ~ lab_year, data = d))) %>%
  mutate(lm_coeff = map(lm_model, broom::tidy)) %>%
  select(-data, -lm_model) %>% 
  unnest(lm_coeff) %>% 
  filter(term == 'lab_year') %>% 
  mutate(p.value.bonferroni = p.value * 11)

trendline_lm_test_formatted <- trendline_lm_test %>% select(termid, label, estimate, p.value, p.value.bonferroni) %>% 
  arrange(p.value.bonferroni) %>%
  rename(beta = estimate) %>% 
  mutate(significance = symnum(p.value.bonferroni, cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, Inf), symbols = c("***", "**", "*", ".", " ")) ) %>% 
  mutate(across(.cols = function(x) is.double(x), function(x) scales::scientific(x, digits = 3)))

write.csv(trendline_lm_test_formatted, file = glue('{OUT_DIR}/table_observed_normalized_by_tested_phenotypes_by_year_level_1_normalized_trendline_lm_model.csv'), row.names=FALSE)

trendline_lm_test_formatted
```

In the last section, we show two sets of phenotypes. Set 1, those phenotype trend lower when looking at observed phenotypes only; Set 2, those phenotype trend higher when looking at observed phenotypes only.

The terms have significantly decreased trendlines when looking at patient counts of observed phenotypes. 
```{r}
hp_terms_of_interest <- c('HP:0001871','HP:0002715','HP:0025031','HP:0001626')
p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hp_terms_of_interest, df=observed_divide_tested_all_by_year)

ggsave2(filenames = c(glue('{OUT_DIR}/plot_observed_by_tested_set1.png'), glue('{OUT_DIR}/plot_observed_by_tested_set1.pdf')), plot = p, width = 8, height = 4)

p
```

The terms have significantly increased trendlines when looking at patient counts of observed phenotypes. 
```{r}
hp_terms_of_interest <- c('HP:0000119','HP:0001939')
p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hp_terms_of_interest, df=observed_divide_tested_all_by_year)

ggsave2(filenames = c(glue('{OUT_DIR}/plot_observed_by_tested_set2.png'), glue('{OUT_DIR}/plot_observed_by_tested_set2.pdf')), plot = p, width = 8, height = 4)

p
```

### 2022-04-28 code/r/01_C.Rmd

---
title: "01_B_racial_disparity_observed_normalized_by_tested"
output: html_document
date: '2022-04-28'
author: Aaron Zhang
---

This is another small extension for the main analysis of "01_MSDW_HPO_Phenotype_Analysis.Rmd". 
The goal is to analyze the trendlines of observed phenotypes for each race but now normalize by how many patients were tested for each corresponding phenotype.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(glue)
source('functions.R')
```

## load data
```{r}
# cache directory to read from or write to
CACHE_DIR = '../../cache'
if (!dir.exists(CACHE_DIR)){
  dir.create(CACHE_DIR)
}

# output directory to write analysis results
OUT_DIR = '../../out'
if (!dir.exists(OUT_DIR)){
  dir.create(OUT_DIR)
}

IN_DIR = '../../in'

observed_phenotype_rank_all_by_year_race_sex <- read_csv(glue('{CACHE_DIR}/observed_phenotype_rank_all_by_year_race_sex.csv'))
tested_phenotype_rank_all_by_race_year_sex <- read_csv(glue('{CACHE_DIR}/tested_phenotype_rank_all_by_race_year_sex.csv'))
hp_term_list <- read.csv(glue('{IN_DIR}/hp_term_list.csv'), header = FALSE, col.names = c('termid', 'distancetoroot', 'label'))
```

## analysis
Calculate fraction of patients observed for each phenotype by patients tested for that phenotype.

```{r}
observed_phenotype_rank_all_by_year_race <- observed_phenotype_rank_all_by_year_race_sex %>% 
  group_by(termid, lab_year, race) %>%
  summarise(n_patient_observed = sum(n_patient), .groups = 'drop')

tested_phenotype_rank_all_by_race_year <- tested_phenotype_rank_all_by_race_year_sex %>%
  group_by(termid, lab_year, race) %>%
  summarise(n_patient_tested = sum(n_patient), .groups = 'drop')
```

```{r}
observed_normalized_by_tested_all_year_race <- observed_phenotype_rank_all_by_year_race %>% 
  left_join(tested_phenotype_rank_all_by_race_year, by = c('termid', 'lab_year', 'race')) %>%
  mutate(frac_observed_tested = n_patient_observed / n_patient_tested)

observed_normalized_by_tested_all_year_race <- observed_normalized_by_tested_all_year_race %>% 
  left_join(hp_term_list, by = 'termid')
```

Below, we plot level HPO terms.
```{r}
plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df){
  hpo_terms_selected <- df %>% filter(termid %in% terms_of_interest) %>% select(termid, label) %>% distinct()

  y_str = 'frac_observed_tested'
  y_label = 'patient fraction (observed / tested)'

  df <- df %>% mutate(termid_label = str_c(label, glue('[{termid}]'), sep = '\n'))

  p <- df %>% filter(termid %in% terms_of_interest) %>% 
    filter(lab_year <= 2019 & lab_year > 2000) %>%
    #ggplot(aes(x = lab_year, y = eval(parse(text=y_str)), group = termid, color = termid)) + 
    ggplot(aes(x = lab_year, y = eval(parse(text=y_str)), group = race, color = race)) +
    geom_point() +
    geom_line() +
    #scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " ")) +
    xlab('Year') +
    ylab(y_label) +
    theme_az() + 
    facet_wrap(~termid_label, ncol = 2)

  return (p)
}
```

```{r}
hpo_level_1_terms <- observed_normalized_by_tested_all_year_race %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()

p <- plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=observed_normalized_by_tested_all_year_race %>% filter(race %in% c('White', 'Black or African American', 'Asian')))

ggsave2(filenames = c(glue('{OUT_DIR}/plot_racial_difference_observed_normalized_by_tested.png'), glue('{OUT_DIR}/plot_racial_difference_observed_normalized_by_tested.pdf')), width=8, height = 10)
```

The plot shows racial differences of observed phenotypes.

![racial differences](./../../out/plot_racial_difference_observed_normalized_by_tested.png)

### 2022-04-28 code/r/02_D_numeric_vs_hpo_for_CRC.Rmd

@@
 
-ggsave(filename = glue('{CACHE_DIR}/CRC_labs_raw_data_distribution_by_rel_date.png'), plot = p, width = 8, height = 6)
+ggsave2(filenames = c(glue('{OUT_DIR}/CRC_labs_raw_data_distribution_by_rel_date.png'), glue('{OUT_DIR}/CRC_labs_raw_data_distribution_by_rel_date.pdf')), plot = p, width = 8, height = 6)
 ```
@@
 
-ggsave(glue("{CACHE_DIR}/plot_CRC_before_matching.png"), width = 2, height = 1.5)
+ggsave2(filenames = c(glue("{OUT_DIR}/plot_CRC_before_matching.png"), glue("{OUT_DIR}/plot_CRC_before_matching.pdf")), width = 2, height = 1.5)
 p
@@
     m.out <<- matchit(outcome ~ medical_record_number, data = df_demographics_with_outcomes, method = "nearest", distance = df_demographics_with_outcomes$propensity_score)
-    # save the matchit object in cache
+    # save the matchit object in cache to the global environment
     save(m.out, file=filepath)
   }
-  load(file=filepath)
+  load(file=filepath, envir = globalenv())
 }
@@
 
-ggsave(glue("{CACHE_DIR}/plot_CRC_after_matching.png"), plot = p, width = 2, height = 1.5)
+ggsave2(filenames = c(glue("{OUT_DIR}/plot_CRC_after_matching.png"), glue("{OUT_DIR}/plot_CRC_after_matching.pdf")) , plot = p, width = 2, height = 1.5)
 
@@
 
-ggsave(filename = glue('{CACHE_DIR}/plot_CRC_age_gender_race_after_matching.png'), plot = plot_match, width = 6, height = 5)
+ggsave2(filenames = c(glue('{OUT_DIR}/plot_CRC_age_gender_race_after_matching.png'), glue('{OUT_DIR}/plot_CRC_age_gender_race_after_matching.pdf')), plot = plot_match, width = 6, height = 5)
 plot(plot_match)
 ```
+<br>
 We need to put back the identifiers of controls to the database in order to selectively retrieve their labs. 
@@
 
-crc_case_matched_controls_analysis %>% write.csv(file = glue('{CACHE_DIR}/crc_case_matched_controls_analysis.csv'), row.names = FALSE)
+crc_case_matched_controls_analysis %>% write.csv(file = glue('{OUT_DIR}/crc_case_matched_controls_analysis.csv'), row.names = FALSE)
 

### 2022-04-28 code/r/03_B_biomarker.Rmd

@@
 CACHE_DIR = '../../cache'
-
 if (!dir.exists(CACHE_DIR)){
   dir.create(CACHE_DIR)
+}
+
+# output directory to write analysis results
+OUT_DIR = '../../out'
+if (!dir.exists(OUT_DIR)){
+  dir.create(OUT_DIR)
 }
@@
 
-write.csv(screen_results_annotated, file = glue('{CACHE_DIR}/screen_results_annotated.csv'), row.names = FALSE)
+write.csv(screen_results_annotated, file = glue('{OUT_DIR}/screen_results_annotated.csv'), row.names = FALSE)
 
@@
 
-write.csv(screen_results_annotated_formatted, file = glue('{CACHE_DIR}/screen_results_annotated.csv'), row.names = FALSE)
+write.csv(screen_results_annotated_formatted, file = glue('{OUT_DIR}/screen_results_annotated.csv'), row.names = FALSE)
 
@@
                 title = 'Table Association of Eosinophilia with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_eosinophilia_with_severe_asthma.html'),
+               file = glue('{OUT_DIR}/table_eosinophilia_with_severe_asthma.html'),
                alternate.rows = TRUE)
@@
                 title = 'Table Selected Phenotypes (within 1 year prior) Associated with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html'), 
+               file = glue('{OUT_DIR}/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html'), 
                alternate.rows = TRUE)
@@
                 title = 'Table Selected Phenotypes (within past 3 years) Associated with Severe Asthma',
-               file = glue('{CACHE_DIR}/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html'),
+               file = glue('{OUT_DIR}/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html'),
                alternate.rows = TRUE)
